# National CPS Profiles Behind The Panel

This appendix now reads only the derived parquet outputs produced by `build_panels.R`. It focuses on the dominant occupation and class-of-worker profiles most relevant to the UI take-up interpretation.

```{r ui-cps-national-load}
#| cache: true

ui_profile_data <- read_final_parquet("unemployment_dominant_profiles.parquet") |>
  filter(dimension %in% c("Occupation", "Class"))

ui_profile_metadata <- read_final_parquet("unemployment_metadata.parquet")

ui_profile_variants <- c(
  pooled = "1994-present",
  year_2025 = "Calendar year 2025",
  delta_2025 = "2025 minus pooled"
)

ui_profile_colors <- c(
  "Temporary Layoff" = "#D55E00",
  "Permanent Job Losers" = "#0072B2",
  "Completed Temp Job" = "#009E73",
  "Job Leavers" = "#CC79A7",
  "Reentrants" = "#E69F00",
  "New Entrants" = "#7F7F7F"
)
```

## National Work-Background Snapshot

The strongest work-background signal in the parquet-backed summaries is the dominant subgroup on the occupation and class margins for each unemployment cause. That is the part most directly relevant for the UI take-up design: whether the unemployed pool is tilting toward white-collar, blue-collar, private-sector, government, or no-prior-job backgrounds.

```{r ui-work-profile-table}
#| tbl-cap: "Dominant occupation and class profile by unemployment cause."

ui_profile_data |>
  mutate(
    Window = recode(variant, !!!ui_profile_variants),
    Share = if_else(
      variant == "delta_2025",
      paste0(if_else(delta > 0, "+", ""), scales::number(delta * 100, accuracy = 0.1), " pp"),
      scales::percent(share, accuracy = 0.1)
    )
  ) |>
  select(
    Cause = cause,
    Dimension = dimension,
    Window,
    `Dominant Group` = dominant_group,
    Share
  ) |>
  gt(groupname_col = "Window") |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

```{r ui-work-profile-chart}
#| fig-cap: "Dominant occupation/class share by unemployment cause. The 2025-minus-pooled panel reports percentage-point change in the dominant subgroup's share."
#| fig-height: 8

ui_profile_data |>
  mutate(
    Window = factor(
      recode(variant, !!!ui_profile_variants),
      levels = unname(ui_profile_variants)
    ),
    plot_value = if_else(variant == "delta_2025", delta, share),
    label = if_else(
      variant == "delta_2025",
      paste0(dominant_group, "\n", if_else(delta > 0, "+", ""), scales::number(delta * 100, accuracy = 0.1), " pp"),
      paste0(dominant_group, "\n", scales::percent(share, accuracy = 1))
    ),
    label_y = case_when(
      variant == "delta_2025" & plot_value >= 0 ~ plot_value + 0.015,
      variant == "delta_2025" & plot_value < 0 ~ plot_value - 0.015,
      TRUE ~ plot_value + 0.03
    ),
    label_vjust = case_when(
      variant == "delta_2025" & plot_value < 0 ~ 1,
      TRUE ~ 0
    ),
    cause = factor(
      cause,
      levels = c(
        "Temporary Layoff",
        "Permanent Job Losers",
        "Completed Temp Job",
        "Job Leavers",
        "Reentrants",
        "New Entrants"
      )
    )
  ) |>
  ggplot(aes(x = cause, y = plot_value, fill = cause)) +
  geom_col(width = 0.72, alpha = 0.9) +
  geom_hline(yintercept = 0, color = "grey70", linewidth = 0.4) +
  facet_grid(Window ~ dimension, scales = "free_y") +
  geom_text(
    aes(y = label_y, label = label, vjust = label_vjust),
    size = 3,
    lineheight = 0.9
  ) +
  scale_fill_manual(values = ui_profile_colors) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  labs(
    title = "Dominant Work-Background Profiles by Cause",
    subtitle = "Occupation and class margins only; derived from the parquet-backed national CPS summaries",
    x = NULL,
    y = NULL,
    fill = NULL,
    caption = "Source: data/final/unemployment_dominant_profiles.parquet"
  ) +
  theme_report +
  theme(
    axis.text.x = element_text(angle = 25, hjust = 1),
    legend.position = "none"
  )
```

These dominant-profile summaries preserve the main point for the memo: layoffs and permanent job losers retain a distinct work-background signature, while new entrants remain mechanically concentrated in no-prior-job backgrounds. That is enough to keep the UI design discussion connected to the CPS composition evidence without re-reading raw microdata at render time.
