---
title: "Temporary Unemployed to Employed"
subtitle: "Linked CPS Follow-Up for Temporary Layoffs vs Permanent Job Losers"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 5.5
    embed-resources: true
    self-contained-math: true
execute:
  warning: false
  message: false
---

## Overview

This note asks a simple panel-style question using linked monthly CPS records: when someone reports being unemployed because of a **temporary layoff**, how often do they later show up as **employed** in subsequent CPS interviews, and how has that changed over time?

The comparison group is **permanent job losers**. The report tracks three linked outcomes by the respondent's **origin month**:

1. employed in the **next observed interview**
2. employed **within the next 3 observed interviews**
3. employed at the respondent's **first reappearance after the long CPS rotation gap**

All three are shown **over time**, with temporary layoffs and permanent job losers plotted together.

## Rotation Design

The CPS is not a clean 16-month monthly panel. It uses a **4-8-4 rotation**:

- interviewed for 4 consecutive months
- out of sample for 8 months
- interviewed for 4 consecutive months again

So this report does **not** measure "employed within the next 6 calendar months" for everyone. Instead it measures employment at later **observed CPS interviews**. For linkage, the report uses the validated person identifier `CPSIDV`.

```{r setup}
library(tidyverse)
library(here)
library(ipumsr)
library(lubridate)
library(gt)
library(scales)

here::i_am("temporary_to_employed.qmd")

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )

theme_set(theme_report)

analysis_dir <- here("data", "CPS", "temporary_to_employed")
dir.create(analysis_dir, recursive = TRUE, showWarnings = FALSE)

extract_ddi_path <- here("data", "CPS", "temporary_to_employed", "cps_temp_to_employed.xml")
extract_csv_path <- here("data", "CPS", "temporary_to_employed", "cps_temp_to_employed.csv.gz")
monthly_transition_path <- here("data", "CPS", "temporary_to_employed", "monthly_transition_rates.csv")
origin_record_path <- here("data", "CPS", "temporary_to_employed", "origin_records_with_followup.csv.gz")
linkage_summary_path <- here("data", "CPS", "temporary_to_employed", "linkage_summary.csv")
```

## Extract Definition

```{r extract-download}
#| cache: true

extract_description <- paste(
  "Temporary to employed monthly CPS extract:",
  "CPSIDV/CPSIDP, EMPSTAT, WHYUNEMP, WTFINL, AGE, 1994-present"
)

ensure_temp_to_employed_extract <- function(target_ddi, target_csv, extract_desc) {
  month_index <- function(year, month) {
    as.integer(year) * 12L + as.integer(month)
  }

  recent_year_threshold <- max(1994L, year(Sys.Date()) - 1L)

  ddi_months <- function(ddi_path) {
    if (!file.exists(ddi_path)) {
      return(tibble(year = integer(), month = integer(), ym = integer()))
    }

    read_lines(ddi_path) |>
      str_match("timePrd date=\"(\\d{4})-(\\d{2})\"") |>
      as_tibble(.name_repair = "minimal") |>
      transmute(
        year = as.integer(V2),
        month = as.integer(V3),
        ym = month_index(year, month)
      ) |>
      filter(!is.na(year), !is.na(month)) |>
      distinct(ym, .keep_all = TRUE)
  }

  csv_months <- function(csv_path) {
    if (!file.exists(csv_path)) {
      return(tibble(year = integer(), month = integer(), ym = integer()))
    }

    tryCatch(
      {
        read_csv(
          csv_path,
          col_select = c("YEAR", "MONTH"),
          show_col_types = FALSE,
          progress = FALSE
        ) |>
          transmute(
            year = as.integer(YEAR),
            month = as.integer(MONTH),
            ym = month_index(year, month)
          ) |>
          filter(!is.na(year), !is.na(month)) |>
          distinct(ym, .keep_all = TRUE)
      },
      error = function(e) tibble(year = integer(), month = integer(), ym = integer())
    )
  }

  has_expected_coverage <- function(ddi_path, csv_path) {
    declared_months <- ddi_months(ddi_path)
    observed_months <- csv_months(csv_path)

    if (nrow(declared_months) == 0 || nrow(observed_months) == 0) {
      return(FALSE)
    }

    min(declared_months$year, na.rm = TRUE) <= 1994L &&
      max(declared_months$year, na.rm = TRUE) >= recent_year_threshold &&
      all(declared_months$ym %in% observed_months$ym)
  }

  local_files_exist <- file.exists(target_ddi) && file.exists(target_csv)

  if (local_files_exist && has_expected_coverage(target_ddi, target_csv)) {
    return(list(ddi = target_ddi, csv = target_csv, source = "local-validated"))
  }

  if (local_files_exist) {
    file.remove(c(target_ddi, target_csv))
  }

  extract_history <- tryCatch(
    get_extract_history("cps", how_many = 25),
    error = function(e) NULL
  )

  matching_completed <- list()
  if (!is.null(extract_history)) {
    matching_completed <- keep(
      extract_history,
      \(x) identical(x$description, extract_desc) && identical(x$status, "completed")
    )
  }

  copy_download_to_targets <- function(path_to_ddi) {
    downloaded_csv <- sub("\\.xml$", ".csv.gz", path_to_ddi)

    if (file.exists(target_ddi)) file.remove(target_ddi)
    if (file.exists(target_csv)) file.remove(target_csv)

    ok_ddi <- FALSE
    ok_csv <- FALSE

    for (attempt in 1:3) {
      ok_ddi <- file.copy(path_to_ddi, target_ddi, overwrite = TRUE)
      ok_csv <- file.copy(downloaded_csv, target_csv, overwrite = TRUE)

      if (ok_ddi && ok_csv && file.exists(target_ddi) && file.exists(target_csv)) {
        break
      }

      Sys.sleep(attempt)
    }

    if (!ok_ddi || !ok_csv || !file.exists(target_ddi) || !file.exists(target_csv)) {
      stop("Failed to copy downloaded CPS extract into the analysis directory.")
    }

    file.remove(path_to_ddi)
    file.remove(downloaded_csv)
  }

  try_download_extract <- function(extract_obj) {
    partial_xml <- list.files(analysis_dir, pattern = "^cps_\\d+\\.xml$", full.names = TRUE)
    partial_csv <- list.files(analysis_dir, pattern = "^cps_\\d+\\.csv\\.gz$", full.names = TRUE)

    if (length(c(partial_xml, partial_csv)) > 0) {
      file.remove(c(partial_xml, partial_csv))
    }

    path_to_ddi <- NULL
    for (attempt in 1:5) {
      path_to_ddi <- tryCatch(
        download_extract(extract_obj, download_dir = analysis_dir),
        error = function(e) {
          if (attempt < 5) {
            Sys.sleep(attempt * 10)
          }
          NULL
        }
      )

      if (!is.null(path_to_ddi)) {
        downloaded_csv <- sub("\\.xml$", ".csv.gz", path_to_ddi)

        if (has_expected_coverage(path_to_ddi, downloaded_csv)) {
          return(path_to_ddi)
        }

        file.remove(c(path_to_ddi, downloaded_csv))
      }
    }

    NULL
  }

  for (candidate in matching_completed) {
    path_to_ddi <- try_download_extract(candidate)

    if (is.null(path_to_ddi)) {
      next
    }

    if (has_expected_coverage(path_to_ddi, sub("\\.xml$", ".csv.gz", path_to_ddi))) {
      copy_download_to_targets(path_to_ddi)
      return(list(ddi = target_ddi, csv = target_csv, source = "history-validated"))
    }

    downloaded_csv <- sub("\\.xml$", ".csv.gz", path_to_ddi)
    file.remove(c(path_to_ddi, downloaded_csv))
  }

  all_samples <- get_sample_info("cps")

  ipums_samples <- all_samples |>
    filter(
      grepl("^cps\\d{4}_\\d{2}[bs]$", name),
      as.numeric(gsub("cps(\\d{4}).*", "\\1", name)) >= 1994,
      !str_detect(.data$description, "ASEC")
    ) |>
    pull(name)

  extract_def <- define_extract_micro(
    collection = "cps",
    description = extract_desc,
    samples = ipums_samples,
    variables = list(
      "YEAR", "MONTH", "PERNUM", "WTFINL",
      "CPSIDP", "CPSIDV",
      "AGE", "EMPSTAT", "WHYUNEMP"
    ),
    data_format = "csv"
  )

  submitted <- submit_extract(extract_def)
  downloadable <- wait_for_extract(submitted)
  path_to_ddi <- try_download_extract(downloadable)

  if (is.null(path_to_ddi)) {
    stop("Failed to download the temporary-to-employed CPS extract.")
  }

  if (!has_expected_coverage(sub("\\.xml$", ".csv.gz", path_to_ddi), expected_months)) {
    downloaded_csv <- sub("\\.xml$", ".csv.gz", path_to_ddi)
    file.remove(c(path_to_ddi, downloaded_csv))
    stop("Downloaded temporary-to-employed extract is still missing requested monthly coverage.")
  }

  copy_download_to_targets(path_to_ddi)

  list(ddi = target_ddi, csv = target_csv, source = "downloaded-validated")
}

extract_files <- ensure_temp_to_employed_extract(
  extract_ddi_path,
  extract_csv_path,
  extract_description
)

cps_summary <- read_csv(
  extract_files$csv,
  col_select = c("YEAR", "MONTH"),
  show_col_types = FALSE,
  progress = FALSE
)

cps_var_names <- names(
  read_csv(
    extract_files$csv,
    n_max = 0,
    show_col_types = FALSE,
    progress = FALSE
  )
)
```

```{r extract-summary}
#| tbl-cap: "Linked CPS extract used for the temporary-to-employed analysis."

tibble(
  Item = c(
    "Source",
    "Rows",
    "Years",
    "Months",
    "Variables"
  ),
  Value = c(
    extract_files$source,
    comma(nrow(cps_summary)),
    paste0(min(cps_summary$YEAR, na.rm = TRUE), "-", max(cps_summary$YEAR, na.rm = TRUE)),
    n_distinct(paste(cps_summary$YEAR, cps_summary$MONTH)),
    paste(cps_var_names, collapse = ", ")
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Link Construction

```{r build-linked-panel}
#| cache: true

if (
  file.exists(monthly_transition_path) &&
  file.exists(origin_record_path) &&
  file.exists(linkage_summary_path) &&
  file.mtime(monthly_transition_path) >= file.mtime(extract_csv_path) &&
  file.mtime(origin_record_path) >= file.mtime(extract_csv_path) &&
  file.mtime(linkage_summary_path) >= file.mtime(extract_csv_path)
) {
  monthly_transition_rates <- read_csv(
    monthly_transition_path,
    show_col_types = FALSE,
    progress = FALSE
  )
  origin_records <- read_csv(
    origin_record_path,
    show_col_types = FALSE,
    progress = FALSE
  )
  linkage_summary <- read_csv(
    linkage_summary_path,
    show_col_types = FALSE,
    progress = FALSE
  )
} else {
  stop("Prebuilt panel outputs are missing or stale. Run build_temporary_to_employed_panel.R first.")
}
```

## Linkage Coverage

```{r linkage-coverage}
#| tbl-cap: "Origin-cohort linkage coverage for temporary layoffs and permanent job losers."

linkage_summary |>
  gt() |>
  fmt_number(columns = Observations, decimals = 0) |>
  fmt_percent(columns = `Weighted Linkable Share`, decimals = 1) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Employed in Next Observed Interview

```{r next-observed-plot}
monthly_transition_rates |>
  ggplot(aes(x = date, y = next_obs_employed_share, color = unemployed_reason)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "Temporary Layoff" = "#0073C2",
      "Permanent Job Loser" = "#CD534C"
    )
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Employed in the Next Observed CPS Interview",
    subtitle = "Monthly origin cohorts, weighted by origin-month WTFINL",
    x = NULL,
    y = "Share employed",
    color = NULL,
    caption = "Denominator is the origin cohort with a nonmissing next observed interview."
  )
```

## Employed Within Next 3 Observed Interviews

```{r next3-plot}
monthly_transition_rates |>
  ggplot(aes(x = date, y = next3_employed_share, color = unemployed_reason)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "Temporary Layoff" = "#0073C2",
      "Permanent Job Loser" = "#CD534C"
    )
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Employed Within the Next 3 Observed CPS Interviews",
    subtitle = "Monthly origin cohorts, weighted by origin-month WTFINL",
    x = NULL,
    y = "Share employed",
    color = NULL,
    caption = "Denominator is the origin cohort with at least one observed interview in the next 3 waves."
  )
```

## Employed at Reappearance After the Long Gap

```{r gap-plot}
monthly_transition_rates |>
  ggplot(aes(x = date, y = gap_employed_share, color = unemployed_reason)) +
  geom_line(linewidth = 0.7, alpha = 0.9) +
  scale_color_manual(
    values = c(
      "Temporary Layoff" = "#0073C2",
      "Permanent Job Loser" = "#CD534C"
    )
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Employed at First Reappearance After the Long CPS Gap",
    subtitle = "First later interview with a gap of at least 9 months from the origin month",
    x = NULL,
    y = "Share employed",
    color = NULL,
    caption = "Denominator is the origin cohort that reappears after the long rotation gap."
  )
```

## Coverage Over Time

```{r coverage-plot}
coverage_long <- monthly_transition_rates |>
  select(
    date, unemployed_reason,
    next_obs_followup_share, next3_followup_share, gap_followup_share
  ) |>
  pivot_longer(
    cols = -c(date, unemployed_reason),
    names_to = "metric",
    values_to = "value"
  ) |>
  mutate(
    metric = recode(
      metric,
      next_obs_followup_share = "Next observed interview available",
      next3_followup_share = "At least one of next 3 observed interviews available",
      gap_followup_share = "Reappears after long gap"
    )
  )

coverage_long |>
  ggplot(aes(x = date, y = value, color = unemployed_reason)) +
  geom_line(linewidth = 0.6, alpha = 0.9) +
  facet_wrap(~metric, ncol = 1) +
  scale_color_manual(
    values = c(
      "Temporary Layoff" = "#0073C2",
      "Permanent Job Loser" = "#CD534C"
    )
  ) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  labs(
    title = "Follow-Up Coverage by Origin Month",
    subtitle = "Later origin months lose evaluable follow-up near the sample end",
    x = NULL,
    y = "Coverage share",
    color = NULL
  )
```

## Latest Available Comparison

```{r latest-table}
#| tbl-cap: "Latest available monthly origin-cohort transition rates."

latest_complete <- monthly_transition_rates |>
  filter(!is.na(next_obs_employed_share), !is.na(next3_employed_share), !is.na(gap_employed_share)) |>
  group_by(unemployed_reason) |>
  filter(date == max(date, na.rm = TRUE)) |>
  ungroup() |>
  transmute(
    Date = date,
    Group = unemployed_reason,
    `Employed next observed interview` = next_obs_employed_share,
    `Employed within next 3 observed interviews` = next3_employed_share,
    `Employed after long-gap reappearance` = gap_employed_share
  )

latest_complete |>
  gt() |>
  fmt_date(columns = Date, date_style = 10) |>
  fmt_percent(
    columns = c(
      `Employed next observed interview`,
      `Employed within next 3 observed interviews`,
      `Employed after long-gap reappearance`
    ),
    decimals = 1
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Interpretation Notes

- The "next observed interview" outcome is the cleanest short-run linked measure.
- "Within next 3 observed interviews" is broader: it asks whether the respondent gets back to employment over the next observed stretch of CPS interviews, not necessarily in consecutive calendar months.
- The long-gap reappearance outcome is a separate persistence test. It is not purely a short-run recall measure, but it is useful for seeing whether temporary layoffs eventually reconnect to employment more strongly than permanent job losers.
- Near the end of the sample, later origin months have less follow-up available. The coverage plots above make that explicit.
