---
title: "UI Take-Up Study"
subtitle: "State-Level Research Design Memo"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 5.5
    embed-resources: true
    self-contained-math: true
execute:
  warning: false
  message: false
execute-dir: project
---

# Overview

This memo documents a proposed state-level study of why **UI claiming intensity appears low in recent years**. The core hypothesis is that take-up has fallen partly because the **composition of unemployment** has shifted toward groups that are less likely to be eligible for, or less likely to claim, unemployment insurance.

Two broad stories motivate the exercise:

1. **Eligibility/composition story**. A larger share of the unemployed may be made up of workers with weak UI eligibility, especially **new entrants** with no prior work history, and possibly some **migrant** groups.
2. **Claim-propensity story**. Even among plausibly eligible workers, some groups may be less likely to claim UI because the costs of claiming are high relative to the expected benefit. One candidate mechanism is that some displaced white-collar workers receive severance packages or expect rapid re-employment and therefore do not bother filing.

The clean empirical question is therefore not simply why claims are low, but whether **cross-state, over-time variation in the composition of unemployment helps explain variation in UI claiming intensity and insured unemployment**.

# Recommended Framing

The most defensible baseline outcome is a **normalized claims-intensity measure**, not raw claims. Raw UI claims combine three things at once: the scale of covered employment, the composition of unemployment, and claiming behavior. A ratio is better aligned with the underlying hypothesis.

Recommended outcomes:

- `initial claims / covered employment`
- `insured unemployment rate`
- `continued claims / covered employment` only as an equivalent restatement of the insured unemployment rate, not as a distinct outcome

For population-weighted regressions, the memo also carries **annual state resident population**, repeated within year when the panel is monthly or quarterly.

The first-pass regression should use **state fixed effects** and **date fixed effects**:

$$
\text{takeup}_{st} =
\beta_1 \text{NewEntrantShare}_{st} +
\beta_2 \text{NoncitizenShare}_{st} +
\beta_3 \text{BlueCollarShare}_{st} +
\beta_4 \text{WorkingAgeShare}_{st} +
\alpha_s + \gamma_t + \varepsilon_{st}
$$

where `WorkingAgeShare` refers to the unemployed share ages **25-54**, $\alpha_s$ are state fixed effects, and $\gamma_t$ are date fixed effects — one dummy per period in the panel (month-year for monthly panels, quarter-year for quarterly, year for yearly). These are *not* 12 seasonal dummies; they absorb all common national time variation. Identification therefore comes from **within-period, cross-state variation** in composition after absorbing persistent state differences. Baseline inference uses **standard errors clustered by state** and **annual state population weights**.

The key design choices are:

- **New entrants** are a relatively clean mechanism because they are definitionally searching for a first job and are therefore likely to have weak or zero UI eligibility.
- **Noncitizen share** is the sharper baseline migrant proxy at monthly frequency. Foreign-born share remains a possible later robustness check, but it is not part of the current baseline.
- **Blue-collar share** is included to capture shifts in industry and occupation mix.
- **Working-age (25-54) share** is included to capture changes in the age composition of the unemployed pool.

# National UI Take-Up Evidence

Before moving to the state panel, it helps to establish the core national descriptive patterns in UI take-up and unemployment composition.

```{r}

library(tidyverse)
library(arrow)
library(here)
library(gt)
library(lubridate)
library(plotly)
library(scales)
library(DT)

here::i_am("ui-takeup-study.qmd")

read_final_parquet = function(filename) {
  path = here("data", "final", filename)
  if (!file.exists(path)) {
    stop("Missing derived panel: ", path, ". Run 'Rscript build_panels.R'.")
  }
  arrow::read_parquet(path) |>
    as_tibble()
}

state_meta <- tibble(
  state = state.name,
  abb = state.abb,
  fips = c(
    "01", "02", "04", "05", "06", "08", "09", "10", "12", "13",
    "15", "16", "17", "18", "19", "20", "21", "22", "23", "24",
    "25", "26", "27", "28", "29", "30", "31", "32", "33", "34",
    "35", "36", "37", "38", "39", "40", "41", "42", "44", "45",
    "46", "47", "48", "49", "50", "51", "53", "54", "55", "56"
  )
) |>
  arrange(state)

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )

theme_set(theme_report)
```

## National UI Take-Up: Recent Detail vs. Historical Average

The two panels below show national **continued claims / covered employment** (CC/CE) and **initial claims / covered employment** (IC/CE) at monthly frequency, with the Jan 1994 – Dec 2025 historical average overlaid as a reference line. All three input series come from FRED:

- [CCSA](https://fred.stlouisfed.org/series/CCSA) — continued claims, seasonally adjusted (weekly)
- [ICSA](https://fred.stlouisfed.org/series/ICSA) — initial claims, seasonally adjusted (weekly)
- [COVEMP](https://fred.stlouisfed.org/series/COVEMP) — covered employment (weekly)

Weekly claims are summed (initial) or averaged (continued) within each calendar month; covered employment is averaged within each month.

```{r national-takeup-panel}
#| fig-height: 7

national_monthly = read_final_parquet("ui_national_takeup_monthly.parquet")

hist_avg = national_monthly |>
  summarise(cc_ce_avg = mean(cc_ce, na.rm = TRUE),
            ic_ce_avg = mean(ic_ce, na.rm = TRUE))

recent = national_monthly |> filter(month >= as.Date("2023-01-01"))

p_cc = ggplot(recent, aes(month, cc_ce)) +
  geom_hline(yintercept = hist_avg$cc_ce_avg, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  annotate("text", x = min(recent$month), y = hist_avg$cc_ce_avg,
           label = paste0("1994-2025 avg: ", scales::percent(hist_avg$cc_ce_avg, accuracy = 0.01)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey40") +
  geom_line(color = "#0073C2", linewidth = 0.8) +
  geom_point(color = "#0073C2", size = 1.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(title = "Continued Claims / Covered Employment",
       subtitle = "National monthly, 2023-present vs. historical average",
       x = NULL, y = "CC / CE")

p_ic = ggplot(recent, aes(month, ic_ce)) +
  geom_hline(yintercept = hist_avg$ic_ce_avg, linetype = "dashed", color = "grey50", linewidth = 0.5) +
  annotate("text", x = min(recent$month), y = hist_avg$ic_ce_avg,
           label = paste0("1994-2025 avg: ", scales::percent(hist_avg$ic_ce_avg, accuracy = 0.01)),
           hjust = 0, vjust = 1.5, size = 3, color = "grey40") +
  geom_line(color = "#EFC000", linewidth = 0.8) +
  geom_point(color = "#EFC000", size = 1.2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(title = "Initial Claims / Covered Employment",
       subtitle = "National monthly, 2023-present vs. historical average",
       x = NULL, y = "IC / CE")

library(patchwork)
p_cc / p_ic +
  plot_annotation(
    caption = "Source: FRED (CCSA, ICSA, COVEMP). Weekly claims aggregated to monthly."
  )
```

## UI Claims Intensity vs. Unemployment Rate

How closely do UI claiming measures track the headline unemployment rate? The scatterplots below compare national **continued claims / covered employment** ([CCSA](https://fred.stlouisfed.org/series/CCSA) / [COVEMP](https://fred.stlouisfed.org/series/COVEMP)) and **initial claims / covered employment** ([ICSA](https://fred.stlouisfed.org/series/ICSA) / COVEMP) against the unemployment rate, at monthly frequency from 1994 onward.

```{r fig-claims-vs-urate}
#| fig-cap: "National CC/CE and IC/CE vs. unemployment rate (monthly, 1994-present)."
#| fig-height: 5

scatter_df = read_final_parquet("ui_national_takeup_cause.parquet") |>
  select(date, urate, cc_ce, ic_ce)

cor_cc = cor(scatter_df$urate, scatter_df$cc_ce, use = "complete.obs")
cor_ic = cor(scatter_df$urate, scatter_df$ic_ce, use = "complete.obs")

p_cc_scatter = ggplot(scatter_df, aes(urate, cc_ce)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#0073C2") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", sprintf("%.3f", cor_cc)),
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "bold") +
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(title = "CC / CE vs. Unemployment Rate",
       x = "Unemployment Rate (%)", y = "CC / CE")

p_ic_scatter = ggplot(scatter_df, aes(urate, ic_ce)) +
  geom_point(alpha = 0.4, size = 1.5, color = "#EFC000") +
  geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.6) +
  annotate("text", x = Inf, y = Inf,
           label = paste0("r = ", sprintf("%.3f", cor_ic)),
           hjust = 1.1, vjust = 1.5, size = 4, fontface = "bold") +
  scale_x_continuous(labels = \(x) paste0(x, "%")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.01)) +
  labs(title = "IC / CE vs. Unemployment Rate",
       x = "Unemployment Rate (%)", y = "IC / CE")

p_cc_scatter + p_ic_scatter +
  plot_annotation(caption = "Source: BLS via FRED (CCSA, ICSA, COVEMP, UNEMPLOY, CLF16OV). Monthly, 1994-present.")
```

## UI Take-Up vs. By-Cause Shares

This is a useful descriptive extension. At the national level, the clean question is not causal identification but whether months with different **reason-for-unemployment composition** also have systematically different **UI take-up intensity**, after absorbing simple seasonality. The regressions below therefore treat **permanent job losers** as the omitted reference category and ask how `CC/CE` and `IC/CE` move when unemployment shifts toward temporary layoffs, completed temporary jobs, job leavers, reentrants, or new entrants instead.

Month-of-year fixed effects absorb regular seasonal patterns in claiming, so the fitted values should be read as the take-up level predicted by the **mix of unemployment causes plus seasonality**. The residual is then the gap between observed take-up and what that composition model would have expected.

Formally, for outcome $y_t \in \{CC/CE_t,\ IC/CE_t\}$, the national monthly regression is:

$$
y_t = \alpha
+ \beta_1 \,\text{TempLayoffShare}_t
+ \beta_2 \,\text{CompletedTempJobShare}_t
+ \beta_3 \,\text{JobLeaverShare}_t
+ \beta_4 \,\text{ReentrantShare}_t
+ \beta_5 \,\text{NewEntrantShare}_t
+ \sum_{m=2}^{12} \gamma_m \mathbf{1}\{\text{month}(t)=m\}
+ \varepsilon_t,
$$

where **Permanent Job Losers** is the omitted reference category, so each $\beta_j$ is interpreted relative to a higher permanent-job-loser share holding the other included shares and calendar month fixed.

```{r national-takeup-by-cause-vars}
#| tbl-cap: "Regression specification for the national monthly take-up models."

tibble(
  Term = c(
    "$y_t$",
    "$CC/CE_t$",
    "$IC/CE_t$",
    "$\\text{TempLayoffShare}_t$",
    "$\\text{CompletedTempJobShare}_t$",
    "$\\text{JobLeaverShare}_t$",
    "$\\text{ReentrantShare}_t$",
    "$\\text{NewEntrantShare}_t$",
    "Omitted category",
    "$\\mathbf{1}\\{\\text{month}(t)=m\\}$",
    "$\\varepsilon_t$"
  ),
  Definition = c(
    "Outcome in month $t$; estimated separately for $CC/CE$ and $IC/CE$.",
    "Continued claims divided by covered employment, aggregated to monthly frequency.",
    "Initial claims divided by covered employment, aggregated to monthly frequency.",
    "Share of total unemployment accounted for by temporary layoffs in month $t$.",
    "Share of total unemployment accounted for by completed temporary jobs in month $t$.",
    "Share of total unemployment accounted for by job leavers in month $t$.",
    "Share of total unemployment accounted for by reentrants in month $t$.",
    "Share of total unemployment accounted for by new entrants in month $t$.",
    "Permanent job losers. Because the six shares sum to one, this category is excluded and serves as the reference group.",
    "Month-of-year fixed effects for February through December; January is the omitted seasonal reference month.",
    "Residual take-up variation not explained by the by-cause mix and month-of-year seasonality."
  )
) |>
  gt() |>
  fmt_markdown(columns = everything()) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Term ~ px(180),
    Definition ~ px(520)
  )
```

```{r national-takeup-by-cause-setup}

takeup_cause_df = read_final_parquet("ui_national_takeup_cause.parquet")

fit_national_takeup_composition = function(outcome) {
  lm(
    as.formula(
      paste0(
        outcome,
        " ~ temp_layoff_share + completed_temp_job_share + job_leaver_share + ",
        "reentrant_share + new_entrant_share + month_of_year"
      )
    ),
    data = takeup_cause_df
  )
}

cc_cause_model = fit_national_takeup_composition("cc_ce")
ic_cause_model = fit_national_takeup_composition("ic_ce")

share_term_labels = c(
  temp_layoff_share = "Temporary layoff share",
  completed_temp_job_share = "Completed temp-job share",
  job_leaver_share = "Job leaver share",
  reentrant_share = "Reentrant share",
  new_entrant_share = "New entrant share"
)

extract_share_terms = function(model, outcome_label) {
  coef_mat = summary(model)$coefficients |>
    as.data.frame() |>
    rownames_to_column("term") |>
    as_tibble() |>
    filter(term %in% names(share_term_labels)) |>
    mutate(
      Outcome = outcome_label,
      Term = recode(term, !!!share_term_labels),
      `Coef (10 pp shift)` = Estimate * 0.10,
      `Std. Error (10 pp shift)` = `Std. Error` * 0.10,
      `p-value` = `Pr(>|t|)`
    ) |>
    select(Outcome, Term, `Coef (10 pp shift)`, `Std. Error (10 pp shift)`, `p-value`)
}

national_takeup_coef_table = bind_rows(
  extract_share_terms(cc_cause_model, "CC / CE"),
  extract_share_terms(ic_cause_model, "IC / CE")
)

national_takeup_fit_table = tibble(
  Outcome = c("CC / CE", "IC / CE"),
  `R-squared` = c(summary(cc_cause_model)$r.squared, summary(ic_cause_model)$r.squared),
  RMSE = c(
    sqrt(mean(residuals(cc_cause_model)^2, na.rm = TRUE)),
    sqrt(mean(residuals(ic_cause_model)^2, na.rm = TRUE))
  ),
  `Sample Start` = min(takeup_cause_df$date),
  `Sample End` = max(takeup_cause_df$date)
)

takeup_cause_fitted = takeup_cause_df |>
  mutate(
    cc_ce_hat = predict(cc_cause_model, newdata = takeup_cause_df),
    ic_ce_hat = predict(ic_cause_model, newdata = takeup_cause_df),
    cc_gap = cc_ce - cc_ce_hat,
    ic_gap = ic_ce - ic_ce_hat
  )

recent_takeup_gap_summary = tibble(
  Outcome = c("CC / CE", "IC / CE"),
  `Average Gap Since 2023` = c(
    mean(takeup_cause_fitted$cc_gap[takeup_cause_fitted$date >= as.Date("2023-01-01")], na.rm = TRUE),
    mean(takeup_cause_fitted$ic_gap[takeup_cause_fitted$date >= as.Date("2023-01-01")], na.rm = TRUE)
  ),
  `Latest Gap` = c(
    takeup_cause_fitted$cc_gap[which.max(takeup_cause_fitted$date)],
    takeup_cause_fitted$ic_gap[which.max(takeup_cause_fitted$date)]
  ),
  `Latest Month` = max(takeup_cause_fitted$date)
)
```

```{r national-takeup-by-cause-coefs}
#| tbl-cap: "National monthly take-up regressions on by-cause unemployment shares. Permanent job losers are the omitted reference category; coefficients are reported for a 10 percentage-point higher share."

national_takeup_coef_table |>
  mutate(
    `Coef (10 pp shift)` = scales::number(`Coef (10 pp shift)`, accuracy = 0.0001),
    `Std. Error (10 pp shift)` = scales::number(`Std. Error (10 pp shift)`, accuracy = 0.0001),
    `p-value` = scales::number(`p-value`, accuracy = 0.001)
  ) |>
  gt(groupname_col = "Outcome") |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

```{r national-takeup-by-cause-fit}
#| tbl-cap: "Model fit for the national take-up composition regressions."

national_takeup_fit_table |>
  mutate(
    `R-squared` = scales::number(`R-squared`, accuracy = 0.001),
    RMSE = scales::number(RMSE, accuracy = 0.0001),
    `Sample Start` = format(`Sample Start`, "%Y-%m"),
    `Sample End` = format(`Sample End`, "%Y-%m")
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

```{r fig-national-takeup-by-cause-fitted}
#| fig-cap: "Observed versus composition-predicted national take-up, using monthly by-cause shares and month-of-year fixed effects. Displayed window starts in 2019; models are fit on the full monthly sample from 1994 onward."
#| fig-height: 6.5

takeup_cause_recent <- takeup_cause_fitted |>
  filter(date >= as.Date("2019-01-01"))

p_cc_fit <- ggplot(takeup_cause_recent, aes(date)) +
  geom_line(aes(y = cc_ce, color = "Observed"), linewidth = 0.9) +
  geom_line(aes(y = cc_ce_hat, color = "Expected"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Observed" = "#0073C2", "Expected" = "black")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "CC / CE",
    subtitle = "Observed vs. expected from cause shares",
    x = NULL, y = NULL
  )

p_ic_fit <- ggplot(takeup_cause_recent, aes(date)) +
  geom_line(aes(y = ic_ce, color = "Observed"), linewidth = 0.9) +
  geom_line(aes(y = ic_ce_hat, color = "Expected"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Observed" = "#EFC000", "Expected" = "black")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "IC / CE",
    subtitle = "Observed vs. expected from cause shares",
    x = NULL, y = NULL
  )

p_cc_fit + p_ic_fit +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

```{r fig-national-takeup-by-cause-gap}
#| fig-cap: "Gap between observed and composition-predicted national take-up. Positive values mean take-up was above the level implied by the unemployment mix; negative values mean it was below."
#| fig-height: 6.5

p_cc_gap <- ggplot(takeup_cause_recent, aes(date, cc_gap)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_col(fill = "#0073C2", alpha = 0.85, width = 25) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "CC / CE Gap",
    subtitle = "Observed minus expected",
    x = NULL, y = NULL
  )

p_ic_gap <- ggplot(takeup_cause_recent, aes(date, ic_gap)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_col(fill = "#EFC000", alpha = 0.9, width = 25) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "IC / CE Gap",
    subtitle = "Observed minus expected",
    x = NULL, y = NULL
  )

p_cc_gap + p_ic_gap
```

```{r national-takeup-by-cause-gap-summary}
#| tbl-cap: "Recent divergence between observed and composition-predicted take-up."

recent_takeup_gap_summary |>
  mutate(
    `Average Gap Since 2023` = percent(`Average Gap Since 2023`, accuracy = 0.01),
    `Latest Gap` = percent(`Latest Gap`, accuracy = 0.01),
    `Latest Month` = format(`Latest Month`, "%Y-%m")
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

This is best read as a composition-accounting exercise rather than a causal estimate. If the fitted line tracks the broad level of take-up but the residual gap turns persistently negative in recent years, that would mean the **current mix of temporary layoffs, permanent job losers, quits, reentrants, and new entrants would have predicted more claiming than we actually observe**. That would strengthen the case for a separate post-2020 claiming-propensity shift, not just a change in who is unemployed.

## National Take-Up vs. By-Cause Levels And Shares

The share regressions above treat unemployment composition as the main object. A simpler descriptive check is to ask whether national `IC/CE` and `CC/CE` move with the **level** of each reason-for-unemployment category or with that category's **share of total unemployment**, allowing the predictor to lead take-up by one or two months.

This section uses the monthly national FRED series already loaded above. For each cause, it reports:

- the contemporaneous correlation with `CC/CE` and `IC/CE`
- the correlation using a 1-month lag of the unemployment measure
- the correlation using a 2-month lag of the unemployment measure

The plots standardize each series within panel to keep the display parsimonious. They are for visual co-movement only; the tables below report the correlations in the original monthly data. To check whether the temporary-layoff result is mostly a furlough-era artifact, the section now reports both the **full history** and a version that excludes **March 2020 through December 2021**, matching the memo's main monthly pandemic-robustness window.

```{r national-cause-correlation-setup}
cause_display_order <- c(
  "Temporary Layoff",
  "Permanent Job Losers",
  "Completed Temp Job",
  "Job Leavers",
  "Reentrants",
  "New Entrants"
)

national_cause_levels <- read_final_parquet("unemployment_fred_monthly.parquet")

share_name_map <- c(
  temp_layoff_share = "Temporary Layoff",
  perm_job_loser_share = "Permanent Job Losers",
  completed_temp_job_share = "Completed Temp Job",
  job_leaver_share = "Job Leavers",
  reentrant_share = "Reentrants",
  new_entrant_share = "New Entrants"
)

safe_standardize <- function(x) {
  s <- sd(x, na.rm = TRUE)
  m <- mean(x, na.rm = TRUE)
  if (!is.finite(s) || s == 0) {
    return(rep(0, length(x)))
  }
  (x - m) / s
}

covid_month_start <- as.Date("2020-03-01")
covid_month_end <- as.Date("2021-12-01")

cause_level_long <- national_cause_levels |>
  select(date, all_of(cause_display_order)) |>
  pivot_longer(
    cols = all_of(cause_display_order),
    names_to = "cause",
    values_to = "predictor_value"
  ) |>
  mutate(
    cause = factor(cause, levels = cause_display_order),
    measure_type = "Level"
  ) |>
  left_join(
    takeup_cause_df |>
      select(date, cc_ce, ic_ce),
    by = "date"
  )

cause_share_long <- takeup_cause_df |>
  select(date, ends_with("_share"), cc_ce, ic_ce) |>
  pivot_longer(
    cols = ends_with("_share"),
    names_to = "share_name",
    values_to = "predictor_value"
  ) |>
  mutate(
    cause = recode(share_name, !!!share_name_map),
    cause = factor(cause, levels = cause_display_order),
    measure_type = "Share"
  )

build_cause_correlation_table <- function(data, value_label) {
  data |>
    arrange(cause, date) |>
    group_by(cause) |>
    mutate(
      predictor_lag1 = lag(predictor_value, 1),
      predictor_lag2 = lag(predictor_value, 2)
    ) |>
    summarise(
      Measure = value_label,
      `CC/CE (t)` = cor(predictor_value, cc_ce, use = "complete.obs"),
      `CC/CE (t-1)` = cor(predictor_lag1, cc_ce, use = "complete.obs"),
      `CC/CE (t-2)` = cor(predictor_lag2, cc_ce, use = "complete.obs"),
      `IC/CE (t)` = cor(predictor_value, ic_ce, use = "complete.obs"),
      `IC/CE (t-1)` = cor(predictor_lag1, ic_ce, use = "complete.obs"),
      `IC/CE (t-2)` = cor(predictor_lag2, ic_ce, use = "complete.obs"),
      .groups = "drop"
    ) |>
    rename(Cause = cause)
}

level_corr_table <- build_cause_correlation_table(cause_level_long, "Level")
share_corr_table <- build_cause_correlation_table(cause_share_long, "Share")
level_corr_table_excl_covid <- build_cause_correlation_table(
  filter(cause_level_long, date < covid_month_start | date > covid_month_end),
  "Level"
)
share_corr_table_excl_covid <- build_cause_correlation_table(
  filter(cause_share_long, date < covid_month_start | date > covid_month_end),
  "Share"
)

build_cause_comovement_plot <- function(data, predictor_label) {
  data |>
    transmute(
      date,
      cause,
      !!predictor_label := predictor_value,
      `CC/CE` = cc_ce,
      `IC/CE` = ic_ce
    ) |>
    pivot_longer(
      cols = -c(date, cause),
      names_to = "series",
      values_to = "value"
    ) |>
    group_by(cause, series) |>
    mutate(z_value = safe_standardize(value)) |>
    ungroup() |>
    ggplot(aes(date, z_value, color = series)) +
    geom_hline(yintercept = 0, color = "grey85", linewidth = 0.3) +
    geom_line(linewidth = 0.7) +
    facet_wrap(~ cause, ncol = 2, scales = "free_x") +
    scale_color_manual(
      values = c(
        setNames("black", predictor_label),
        "CC/CE" = "#0073C2",
        "IC/CE" = "#EFC000"
      )
    ) +
    labs(
      x = NULL,
      y = "Standardized value",
      color = NULL
    ) +
    theme_report +
    theme(
      axis.text.x = element_text(size = 7),
      legend.position = "top"
    )
}
```

### By-Cause Levels

::: {.panel-tabset}

#### Full History

```{r fig-national-cause-level-comovement-full}
#| fig-cap: "Standardized national monthly co-movement between each by-cause unemployment level and UI take-up measures. Each panel overlays the cause level, CC/CE, and IC/CE."
#| fig-height: 9

build_cause_comovement_plot(cause_level_long, "Cause level") +
  labs(
    title = "National Take-Up vs. By-Cause Unemployment Levels",
    subtitle = "Full monthly history; each series standardized within panel"
  )
```

```{r national-cause-level-correlation-table-full}
#| tbl-cap: "Correlation of national IC/CE and CC/CE with monthly by-cause unemployment levels and their lags, full history."

level_corr_table |>
  mutate(across(-c(Cause, Measure), \(x) sprintf("%.3f", x))) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

#### Excl COVID

```{r fig-national-cause-level-comovement-excl-covid}
#| fig-cap: "Standardized national monthly co-movement between each by-cause unemployment level and UI take-up measures, excluding March 2020 through December 2021."
#| fig-height: 9

build_cause_comovement_plot(
  filter(cause_level_long, date < covid_month_start | date > covid_month_end),
  "Cause level"
) +
  labs(
    title = "National Take-Up vs. By-Cause Unemployment Levels",
    subtitle = "Excluding March 2020 through December 2021; each series standardized within panel"
  )
```

```{r national-cause-level-correlation-table-excl-covid}
#| tbl-cap: "Correlation of national IC/CE and CC/CE with monthly by-cause unemployment levels and their lags, excluding March 2020 through December 2021."

level_corr_table_excl_covid |>
  mutate(across(-c(Cause, Measure), \(x) sprintf("%.3f", x))) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

:::

### By-Cause Shares

::: {.panel-tabset}

#### Full History

```{r fig-national-cause-share-comovement-full}
#| fig-cap: "Standardized national monthly co-movement between each by-cause unemployment share and UI take-up measures. Each panel overlays the cause share, CC/CE, and IC/CE."
#| fig-height: 9

build_cause_comovement_plot(cause_share_long, "Cause share") +
  labs(
    title = "National Take-Up vs. By-Cause Unemployment Shares",
    subtitle = "Full monthly history; each series standardized within panel"
  )
```

```{r national-cause-share-correlation-table-full}
#| tbl-cap: "Correlation of national IC/CE and CC/CE with monthly by-cause unemployment shares and their lags, full history."

share_corr_table |>
  mutate(across(-c(Cause, Measure), \(x) sprintf("%.3f", x))) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

#### Excl COVID

```{r fig-national-cause-share-comovement-excl-covid}
#| fig-cap: "Standardized national monthly co-movement between each by-cause unemployment share and UI take-up measures, excluding March 2020 through December 2021."
#| fig-height: 9

build_cause_comovement_plot(
  filter(cause_share_long, date < covid_month_start | date > covid_month_end),
  "Cause share"
) +
  labs(
    title = "National Take-Up vs. By-Cause Unemployment Shares",
    subtitle = "Excluding March 2020 through December 2021; each series standardized within panel"
  )
```

```{r national-cause-share-correlation-table-excl-covid}
#| tbl-cap: "Correlation of national IC/CE and CC/CE with monthly by-cause unemployment shares and their lags, excluding March 2020 through December 2021."

share_corr_table_excl_covid |>
  mutate(across(-c(Cause, Measure), \(x) sprintf("%.3f", x))) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

:::

For placement, this is the right spot: it stays with the other **national descriptive checks** and comes before the memo transitions into the state-panel data build.

# Data Feasibility

```{r}
#| tbl-cap: "Data-source feasibility for a monthly state UI take-up study."

tibble(
  Source = c("FRED / ETA", "FRED / LAUS", "IPUMS CPS", "IPUMS CPS"),
  Measure = c(
    "State UI claims data",
    "State unemployment totals",
    "State-month new entrant composition",
    "State-month migrant composition"
  ),
  `What Exists` = c(
    "Weekly initial claims, continued claims, insured unemployment rate, and covered employment by state",
    "Monthly state unemployment rates and unemployed persons",
    "Monthly CPS microdata with WHYUNEMP and STATEFIP",
    "Monthly CPS microdata with STATEFIP plus NATIVITY and CITIZEN"
  ),
  Feasible = c("Yes", "Yes", "Yes", "Yes, with proxy choice"),
  Notes = c(
    "Can be aggregated to month for state panel construction",
    "Useful unemployment-side context, but not the preferred claims denominator",
    "Can construct weighted state-month share of unemployed who are new entrants",
    "Best monthly proxies are foreign-born or noncitizen shares; recent-migrant measures are weaker"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Source ~ px(100),
    Measure ~ px(150),
    `What Exists` ~ px(240),
    Feasible ~ px(80),
    Notes ~ px(220)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = c(Source, Feasible))
  )
```

# What FRED Can Provide

FRED appears sufficient for the **state UI outcome side** of the study. The relevant release is the **State Unemployment Insurance Weekly Claims Report**, which includes:

- **Initial claims by state**
- **Continued claims (insured unemployment) by state**
- **Insured unemployment rate by state**
- **Covered employment by state**

These are weekly series, so a monthly panel requires an aggregation choice. The memo uses the following rules:

- **Initial claims**: sum all weekly filings within the calendar month (flow measure)
- **Continued claims**: average weekly continued claims within the calendar month (stock measure)
- **Insured unemployment rate**: average weekly rate within the calendar month

Weekly observations are assigned to months via `floor_date(date, "month")`, so weeks that straddle month boundaries are assigned to the month containing their start date. End-of-month continued claims is a natural robustness alternative.

FRED also has **monthly state unemployment rates** and **monthly state unemployed persons** from LAUS, which are natural denominators for a take-up ratio.

## FRED Inputs For The Panel

```{r}
#| tbl-cap: "Recommended FRED-side inputs for panel construction."

tibble(
  Input = c(
    "Initial claims by state",
    "Continued claims by state",
    "Insured unemployment rate by state",
    "Covered employment by state",
    "State unemployed persons",
    "State unemployment rate",
    "Annual state resident population"
  ),
  Definition = c(
    "Number of first-time UI claims filed during the reference week",
    "Number of people receiving UI for a week of unemployment after an initial claim",
    "Insured unemployment as a share of covered employment",
    "Number of employees covered by state UI programs",
    "Estimated number of unemployed persons in the state",
    "State unemployment rate from LAUS",
    "Resident population in the state"
  ),
  `What It Captures` = c(
    "Flow into the UI system",
    "Stock of active UI receipt",
    "UI intensity relative to covered jobs",
    "The insured job base used to normalize claims",
    "Scale of unemployment in the state labor market",
    "Overall labor-market slack",
    "State size for regression weighting"
  ),
  Frequency = c("Weekly", "Weekly", "Weekly", "Weekly", "Monthly", "Monthly", "Annual"),
  `Statistical Utility` = c(
    "Alternative flow-style outcome; useful robustness check",
    "Preferred stock-style UI receipt measure for baseline regressions",
    "Direct take-up/intensity outcome that avoids constructing a ratio",
    "Preferred denominator for initial-claims and continued-claims normalization",
    "Useful labor-market context, but not the preferred denominator for claims normalization",
    "Control variable and descriptive labor-market context",
    "Population weight for state-level regressions"
  ),
  `Preferred Role` = c(
    "Robustness",
    "Baseline outcome",
    "Alternative baseline outcome",
    "Baseline denominator",
    "Baseline denominator",
    "Control / context",
    "Regression weight"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Input ~ px(140),
    Definition ~ px(190),
    `What It Captures` ~ px(165),
    Frequency ~ px(75),
    `Statistical Utility` ~ px(225),
    `Preferred Role` ~ px(110)
  )
```

The memo now pulls these FRED inputs for **all 50 states**. For the weekly UI series, the state-level naming pattern is consistent across the FRED release (`XXICLAIMS`, `XXCCLAIMS`, `XXINSUREDUR`, `XXCEMPLOY`). For the monthly LAUS inputs, the state unemployment rate is `XXUR` and the seasonally adjusted unemployed-persons series follows the `LASST{state_fips}0000000000004` pattern.
For annual population, the memo uses the FRED state resident population series `XXPOP`.

The table below is a **pre-pull inspection table** so you can verify the exact series IDs for every state before the live FRED pull runs.

```{r fred-all-states-pull}
#| cache: true

fred_panel_path = here("data", "final", "ui_fred_monthly_state_panel.parquet")
fred_catalog_path = here("data", "final", "ui_fred_series_catalog.parquet")

fred_catalog_export = read_final_parquet("ui_fred_series_catalog.parquet")
fred_state_month_panel = read_final_parquet("ui_fred_monthly_state_panel.parquet") |>
  mutate(date = as.Date(date))
```

```{r}
#| tbl-cap: "All-state FRED pull coverage used for the memo's UI panel."

fred_state_month_panel |>
  summarise(
    States = n_distinct(state),
    `Date Range Start` = min(date, na.rm = TRUE),
    `Date Range End` = max(date, na.rm = TRUE),
    Rows = n()
  ) |>
  mutate(
    `Date Range Start` = format(`Date Range Start`, "%Y-%m-%d"),
    `Date Range End` = format(`Date Range End`, "%Y-%m-%d")
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    States ~ px(70),
    `Date Range Start` ~ px(120),
    `Date Range End` ~ px(120),
    Rows ~ px(90)
  ) |>
  tab_source_note(
    source_note = md(paste0(
      "Saved locally to `", fred_panel_path, "` and `", fred_catalog_path, "`."
    ))
  )
```

```{r}
#| tbl-cap: "Latest month available in the saved FRED state panel."

fred_state_month_panel |>
  filter(date == max(date, na.rm = TRUE)) |>
  summarise(
    States = n_distinct(state),
    `Latest Month` = max(date, na.rm = TRUE)
  ) |>
  mutate(`Latest Month` = format(`Latest Month`, "%Y-%m-%d")) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

```{r}
#| tbl-cap: "Saved FRED panel columns."

tibble(
  Column = c(
    "date",
    "state",
    "state_abbr",
    "initial_claims",
    "continued_claims",
    "insured_unemployment_rate",
    "covered_employment",
    "annual_population",
    "unemployed_persons",
    "unemployment_rate"
  ),
  Description = c(
    "Month of observation",
    "State name",
    "Two-letter state abbreviation",
    "Sum of weekly initial claims within month",
    "Average weekly continued claims within month",
    "Average weekly insured unemployment rate within month",
    "Implied covered employment = continued claims / (insured unemployment rate / 100)",
    "Annual state resident population, repeated across months within the same year",
    "Monthly unemployed persons",
    "Monthly unemployment rate"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Column ~ px(170),
    Description ~ px(380)
  )
```

```{r fred-series-table}
#| tbl-cap: "All-state FRED reference table for the proposed UI take-up panel."

state_meta |>
  transmute(
    State = state,
    `Initial Claims` = paste0("[", abb, "ICLAIMS](https://fred.stlouisfed.org/series/", abb, "ICLAIMS)"),
    `Continued Claims` = paste0("[", abb, "CCLAIMS](https://fred.stlouisfed.org/series/", abb, "CCLAIMS)"),
    `Insured UR` = paste0("[", abb, "INSUREDUR](https://fred.stlouisfed.org/series/", abb, "INSUREDUR)"),
    `Covered Employment` = paste0("[", abb, "CEMPLOY](https://fred.stlouisfed.org/series/", abb, "CEMPLOY)"),
    `Unemployed Persons` = paste0("[LASST", fips, "0000000000004](https://fred.stlouisfed.org/series/LASST", fips, "0000000000004)"),
    `Unemployment Rate` = paste0("[", abb, "UR](https://fred.stlouisfed.org/series/", abb, "UR)")
  ) |>
  gt() |>
  fmt_markdown(columns = c(
    `Initial Claims`,
    `Continued Claims`,
    `Insured UR`,
    `Covered Employment`,
    `Unemployed Persons`,
    `Unemployment Rate`
  )) |>
  tab_options(
    table.font.size = px(10),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    State ~ px(120),
    `Initial Claims` ~ px(120),
    `Continued Claims` ~ px(130),
    `Insured UR` ~ px(110),
    `Covered Employment` ~ px(135),
    `Unemployed Persons` ~ px(150),
    `Unemployment Rate` ~ px(120)
  ) |>
  tab_source_note(
    source_note = md("These links point to the FRED series IDs relevant to the all-state UI panel above, including covered employment used to normalize claims.")
  )
```

```{r fred-panel-helpers}
fred_analysis_specs <- tibble(
  variable = c(
    "initial_claims_per_covered_employment",
    "continued_claims_per_covered_employment",
    "insured_unemployment_rate_share"
  ),
  label = c(
    "Initial Claims / Covered Employment",
    "Continued Claims / Covered Employment",
    "Insured Unemployment Rate"
  ),
  p10_color = c("#4c78a8", "#b2477c", "#2f6f4f"),
  p50_color = c("#0b3c5d", "#7a1f52", "#1b4332"),
  p90_color = c("#3f88c5", "#e56b6f", "#52b788")
)

fred_analysis_panel <- fred_state_month_panel |>
  mutate(
    initial_claims_per_covered_employment = initial_claims / covered_employment,
    continued_claims_per_covered_employment = continued_claims / covered_employment,
    insured_unemployment_rate_share = insured_unemployment_rate / 100
  ) |>
  pivot_longer(
    cols = all_of(fred_analysis_specs$variable),
    names_to = "variable",
    values_to = "value"
  ) |>
  left_join(fred_analysis_specs, by = "variable") |>
  filter(is.finite(value))

fred_analysis_summary_long <- fred_analysis_panel |>
  group_by(variable, label, date, p10_color, p50_color, p90_color) |>
  summarise(
    p10 = quantile(value, probs = 0.10, na.rm = TRUE),
    p50 = quantile(value, probs = 0.50, na.rm = TRUE),
    p90 = quantile(value, probs = 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(p10, p50, p90),
    names_to = "stat",
    values_to = "value"
  ) |>
  mutate(
    stat_label = recode(stat, p10 = "P10", p50 = "P50", p90 = "P90"),
    stat_color = case_when(
      stat == "p10" ~ p10_color,
      stat == "p50" ~ p50_color,
      stat == "p90" ~ p90_color
    )
  )

fred_analysis_annual_avg <- fred_analysis_panel |>
  mutate(year = year(date)) |>
  group_by(variable, label, year, state, state_abbr) |>
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

build_fred_analysis_time_series_plot <- function(var_id, y_range = NULL) {
  state_data <- fred_analysis_panel |>
    filter(variable == var_id)

  summary_data <- fred_analysis_summary_long |>
    filter(variable == var_id)

  label_text <- fred_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)

  p <- plot_ly()

  p <- p |>
    add_lines(
      data = state_data,
      x = ~date,
      y = ~value,
      split = ~state,
      line = list(color = "rgba(150, 150, 150, 0.22)", width = 0.6),
      hoverinfo = "skip",
      showlegend = FALSE
    )

  for (stat_name in c("p10", "p50", "p90")) {
    stat_data <- summary_data |>
      filter(stat == stat_name)

    p <- p |>
      add_lines(
        data = stat_data,
        x = ~date,
        y = ~value,
        name = stat_data$stat_label[[1]],
        line = list(
          color = stat_data$stat_color[[1]],
          width = if (stat_name == "p50") 2.6 else 1.8
        ),
        text = ~paste0(
          label_text, "<br>",
          stat_label, ": ", percent(value, accuracy = 0.1), "<br>",
          "Date: ", format(date, "%Y-%m-%d")
        ),
        hovertemplate = "%{text}<extra></extra>",
        showlegend = TRUE
      )
  }

  p |>
    layout(
      hovermode = "x unified",
      legend = list(orientation = "h", x = 0, y = 1.12),
      margin = list(l = 60, r = 20, t = 40, b = 40),
      xaxis = list(
        title = "",
        rangeslider = list(visible = TRUE)
      ),
      yaxis = list(title = label_text, tickformat = ".1%", range = y_range)
    )
}

build_fred_analysis_map_plot <- function(var_id) {
  map_data <- fred_analysis_annual_avg |>
    filter(variable == var_id)

  label_text <- fred_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)

  plot_ly(
    data = map_data,
    type = "choropleth",
    locationmode = "USA-states",
    locations = ~state_abbr,
    z = ~value,
    frame = ~year,
    colorscale = list(
      c(0.0, "#0b3c5d"),
      c(0.5, "#f3e5d8"),
      c(1.0, "#7f0000")
    ),
    colorbar = list(
      title = paste0(label_text, "<br>NA = gray"),
      tickformat = ".1%",
      x = 1.02,
      xanchor = "left",
      y = 0.5,
      yanchor = "middle",
      len = 0.6,
      thickness = 12
    ),
    marker = list(line = list(color = "white", width = 0.6)),
    text = ~paste0(
      state, "<br>",
      "Year: ", year, "<br>",
      label_text, ": ", percent(value, accuracy = 0.1)
    ),
    hovertemplate = "%{text}<extra></extra>"
  ) |>
    layout(
      margin = list(l = 0, r = 60, t = 24, b = 0),
      annotations = list(
        list(
          x = 1.02,
          y = 0.84,
          xref = "paper",
          yref = "paper",
          text = "NA = gray",
          showarrow = FALSE,
          font = list(size = 10, color = "#555555")
        )
      ),
      geo = list(
        scope = "usa",
        projection = list(type = "albers usa"),
        showlakes = FALSE,
        bgcolor = "rgba(0,0,0,0)"
      )
    )
}

build_fred_analysis_summary_table <- function(var_id) {
  label_text <- fred_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)

  fred_analysis_panel |>
    filter(variable == var_id) |>
    summarise(
      States = n_distinct(state),
      `Monthly Observations` = n(),
      `Start Month` = min(date),
      `End Month` = max(date),
      Mean = mean(value, na.rm = TRUE),
      SD = sd(value, na.rm = TRUE),
      Min = min(value, na.rm = TRUE),
      P10 = quantile(value, probs = 0.10, na.rm = TRUE),
      P50 = quantile(value, probs = 0.50, na.rm = TRUE),
      P90 = quantile(value, probs = 0.90, na.rm = TRUE),
      Max = max(value, na.rm = TRUE)
    ) |>
    gt() |>
    fmt_date(columns = c(`Start Month`, `End Month`), date_style = 10) |>
    fmt_percent(columns = c(Mean, SD, Min, P10, P50, P90, Max), decimals = 1) |>
    tab_header(title = label_text) |>
    tab_options(
      table.font.size = px(11),
      column_labels.font.weight = "bold",
      column_labels.background.color = "#2c3e50"
    ) |>
    tab_style(
      style = cell_text(color = "white"),
      locations = cells_column_labels()
    )
}
```

# FRED Panel Assembly

California was gray in the earlier initial-claims map because the state initial-claims series does not begin until **February 1986** in the saved panel, so any earlier animation year has missing values and Plotly renders those states in gray. The normalized claims sections below only use nonmissing years for the mapped variable.

The covered-employment normalization is cleaner for the administrative claims series. `Initial claims` is a **monthly flow** formed by summing weekly filings across the month, while `covered employment` is the insured job base. That ratio can still be fairly large in recessions, but it no longer mixes a monthly claims flow with a CPS/LAUS unemployment stock. `Continued claims / covered employment` is mechanically the same object as the insured unemployment rate in share form.

The covered-employment-normalized claims sections extend through the end of the weekly UI data because both `initial claims / covered employment` and `continued claims / covered employment` can be constructed from the UI release itself. The unemployment-based context series still stop at **2025-12-01** in the saved LAUS pull.

## Initial Claims Adjusted by Covered Employment

::: {.panel-tabset}

### Time Series

```{r initial-claims-adjusted-timeseries}
build_fred_analysis_time_series_plot("initial_claims_per_covered_employment")
```

### Time Series - Trim Y

```{r initial-claims-adjusted-timeseries-trim}
build_fred_analysis_time_series_plot("initial_claims_per_covered_employment", y_range = c(0, 0.1))
```

### Interactive Map

```{r initial-claims-adjusted-map}
build_fred_analysis_map_plot("initial_claims_per_covered_employment")
```

### Summary Statistics

```{r initial-claims-adjusted-summary}
build_fred_analysis_summary_table("initial_claims_per_covered_employment")
```

:::

## Continued Claims Adjusted by Covered Employment

::: {.panel-tabset}

### Time Series

```{r continued-claims-adjusted-timeseries}
build_fred_analysis_time_series_plot("continued_claims_per_covered_employment")
```

### Time Series - Trim Y

```{r continued-claims-adjusted-timeseries-trim}
build_fred_analysis_time_series_plot("continued_claims_per_covered_employment", y_range = c(0, 0.1))
```

### Interactive Map

```{r continued-claims-adjusted-map}
build_fred_analysis_map_plot("continued_claims_per_covered_employment")
```

### Summary Statistics

```{r continued-claims-adjusted-summary}
build_fred_analysis_summary_table("continued_claims_per_covered_employment")
```

:::

## Insured Unemployment Rate

::: {.panel-tabset}

### Time Series

```{r insured-ur-timeseries}
build_fred_analysis_time_series_plot("insured_unemployment_rate_share")
```

### Time Series - Trim Y

```{r insured-ur-timeseries-trim}
build_fred_analysis_time_series_plot("insured_unemployment_rate_share", y_range = c(0, 0.1))
```

### Interactive Map

```{r insured-ur-map}
build_fred_analysis_map_plot("insured_unemployment_rate_share")
```

### Summary Statistics

```{r insured-ur-summary}
build_fred_analysis_summary_table("insured_unemployment_rate_share")
```

:::

# What CPS Can Provide

Monthly CPS appears sufficient for the **state-month composition side** of the study. The survey is the natural complement to the FRED panel: FRED provides the **UI outcome side**, while CPS provides the **who-is-unemployed** side.

For this project, the monthly CPS can provide:

- **new entrant share** of the unemployed by state-month
- **foreign-born share** of the unemployed by state-month
- **noncitizen share** of the unemployed by state-month
- reason-for-unemployment composition such as temporary layoff, permanent job loser, job leaver, reentrant, and new entrant
- occupation, industry, education, age, sex, and race/ethnicity composition measures for the unemployed

The cleanest baseline composition variable is the **new entrant share of unemployment**, because monthly CPS directly observes:

- `WHYUNEMP`, which identifies **new entrants** and the other unemployment-reason categories
- `STATEFIP`, which identifies state of residence
- `WTFINL`, which allows weighted state-month aggregation

The migrant side is workable, but more limited. Monthly CPS supports:

- `NATIVITY` for a **foreign-born share** measure
- `CITIZEN` for a **noncitizen share** measure

It does **not** cleanly support a monthly state panel of recent-arrival status, since the most obvious migration-status variable for that story, `MIGRATE1`, is primarily an **ASEC** measure rather than a standard monthly CPS workhorse.

As a result, the monthly CPS should be used to construct **state-month composition measures among unemployed persons**, not direct administrative claiming outcomes. In practice, the workflow is:

1. keep monthly CPS records for unemployed individuals
2. identify state with `STATEFIP`
3. weight records with `WTFINL`
4. compute state-month composition shares
5. merge those CPS-side measures to the saved FRED panel

## CPS Inputs For The Panel

```{r}
#| tbl-cap: "Recommended CPS-side inputs for state-month composition measures."

tibble(
  Variable = c("STATEFIP", "WHYUNEMP", "NATIVITY", "CITIZEN", "WTFINL"),
  Definition = c(
    "State of residence identifier",
    "Reason for unemployment category",
    "Whether the person is native-born or foreign-born",
    "Citizenship status",
    "Final person weight"
  ),
  `What It Captures` = c(
    "Geographic location for the state-month panel",
    "Composition of unemployment by cause, including new entrants",
    "Broad migrant background",
    "Legal-status-adjacent proxy for potential UI access differences",
    "Population-representative weighting"
  ),
  `Monthly Use` = c("Yes", "Yes", "Yes", "Yes", "Yes"),
  `Recommended Construction` = c(
    "Group to state-month",
    "New entrant share among unemployed",
    "Foreign-born share among unemployed",
    "Noncitizen share among unemployed",
    "Weighted shares and totals"
  ),
  `Statistical Utility` = c(
    "Core geography for the panel merge",
    "Cleanest eligibility-composition regressor in the study",
    "Best broad migrant proxy for baseline specifications",
    "Useful alternative or robustness check to nativity",
    "Required for constructing all state-month weighted shares and totals"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Variable ~ px(90),
    Definition ~ px(180),
    `What It Captures` ~ px(180),
    `Monthly Use` ~ px(70),
    `Recommended Construction` ~ px(170),
    `Statistical Utility` ~ px(220)
  )
```

Monthly CPS does **not** provide a clean long panel of UI claiming, claims burden, or severance receipt. Those are conceptually important, but they belong outside the CPS panel-assembly step and should be treated as separate extensions or external controls.

# CPS Panel Assembly

The CPS state-month composition panel is saved locally, so the memo can load the aggregated panel directly without rereading the raw microdata each time. The saved state-month panel used below is `data/final/cps_ui_state_month_panel.parquet`, produced by `build_panels.R` from the unified CPS extract and stored alongside the other final parquet panels.



```{r cps-panel-load}
cps_panel_path = here("data", "final", "cps_ui_state_month_panel.parquet")

cps_state_month_panel = read_final_parquet("cps_ui_state_month_panel.parquet") |>
  mutate(date = as.Date(date))
```

```{r}
#| tbl-cap: "Saved CPS panel coverage used for the memo's composition panel."

cps_state_month_panel |>
  summarise(
    States = n_distinct(state),
    `Date Range Start` = min(date, na.rm = TRUE),
    `Date Range End` = max(date, na.rm = TRUE),
    Rows = n()
  ) |>
  mutate(
    `Date Range Start` = format(`Date Range Start`, "%Y-%m-%d"),
    `Date Range End` = format(`Date Range End`, "%Y-%m-%d")
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    States ~ px(70),
    `Date Range Start` ~ px(120),
    `Date Range End` ~ px(120),
    Rows ~ px(90)
  ) |>
  tab_source_note(
    source_note = md(paste0(
      "Saved locally to `", cps_panel_path, "`, generated by `build_panels.R` and read directly from `data/final`."
    ))
  )
```

White-collar and blue-collar assignments follow the discontinued BLS National Compensation Survey mapping in 2007 Table 2D, which groups occupations into white-collar, blue-collar, and service categories.[BLS Table 2D occupation group definitions](https://www.bls.gov/eci/factsheets/national-compensation-survey-classification-systems-mapping-files.htm) The memo’s `other_services_share` then absorbs everyone not classified as white-collar or blue-collar, so it includes the BLS service occupations plus residual NIU / military / no-usable-prior-occupation cases.

```{r}
#| tbl-cap: "Saved CPS panel columns."

tibble(
  Column = c(
    "date",
    "state",
    "state_abbr",
    "unemployed_total",
    "new_entrant_share",
    "foreign_born_share",
    "noncitizen_share",
    "working_age_25_54_share",
    "white_collar_share",
    "blue_collar_share",
    "other_services_share",
    "service_share",
    "other_occ_share",
    "perm_job_loser_share",
    "temp_layoff_share",
    "completed_temp_job_share",
    "job_leaver_share",
    "reentrant_share",
    "white_collar_perm_job_loser_share"
  ),
  Description = c(
    "Month of observation",
    "State name",
    "Two-letter state abbreviation",
    "Weighted total unemployed persons in the state-month",
    "Weighted share of unemployed persons who are new entrants",
    "Weighted share of unemployed persons who are foreign born",
    "Weighted share of unemployed persons who are noncitizens",
    "Weighted share of unemployed persons ages 25 to 54",
    "Weighted share of unemployed persons in BLS-style white-collar occupations: professional / technical, executive / administrative / managerial, sales, and administrative support jobs",
    "Weighted share of unemployed persons in BLS-style blue-collar occupations: precision production, craft, repair, machine operators, inspectors, assemblers, transportation / material moving, and handlers / helpers / laborers",
    "Weighted share of unemployed persons not classified as white-collar or blue-collar. This `Other Services` bucket includes BLS service occupations plus residual NIU / military / no-usable-prior-occupation cases",
    "Diagnostic component of `other_services_share`: the subset in BLS service occupations",
    "Diagnostic component of `other_services_share`: residual non-white / non-blue cases, mainly NIU / no usable prior occupation and military cases",
    "Weighted share of unemployed persons who are permanent job losers",
    "Weighted share of unemployed persons on temporary layoff",
    "Weighted share of unemployed persons whose prior temporary job ended as scheduled",
    "Weighted share of unemployed persons who voluntarily left their last job",
    "Weighted share of unemployed persons re-entering the labor force after a period out",
    "Weighted share of unemployed persons who are white-collar permanent job losers"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Column ~ px(170),
    Description ~ px(380)
  )
```

```{r cps-panel-helpers}
cps_analysis_specs <- tibble(
  variable = c(
    "new_entrant_share",
    "foreign_born_share",
    "noncitizen_share",
    "white_collar_share",
    "blue_collar_share",
    "other_services_share",
    "perm_job_loser_share",
    "temp_layoff_share",
    "white_collar_perm_job_loser_share"
  ),
  label = c(
    "Share of Unemployed Who Are New Entrants",
    "Share of Unemployed Who Are Foreign Born",
    "Share of Unemployed Who Are Noncitizens",
    "Share of Unemployed in White-Collar Occupations",
    "Share of Unemployed in Blue-Collar Occupations",
    "Share of Unemployed in Other Services",
    "Share of Unemployed Who Are Permanent Job Losers",
    "Share of Unemployed on Temporary Layoff",
    "Share of Unemployed Who Are White-Collar Permanent Job Losers"
  ),
  map_label = c(
    "New entrant share",
    "Foreign-born share",
    "Noncitizen share",
    "White-collar share",
    "Blue-collar share",
    "Other-services share",
    "Permanent job loser share",
    "Temporary layoff share",
    "White-collar permanent job loser share"
  ),
  p10_color = c("#4c78a8", "#2a9d8f", "#b2477c", "#8c6d31", "#3a7d44", "#577590", "#7f5539", "#b56576", "#457b9d"),
  p50_color = c("#0b3c5d", "#1b4332", "#7a1f52", "#5f4b08", "#1b5e20", "#355070", "#5c4033", "#7d3c4c", "#1d3557"),
  p90_color = c("#3f88c5", "#52b788", "#e56b6f", "#d4a373", "#74c69d", "#98c1d9", "#b08968", "#e5989b", "#4ea8de")
)

cps_panel_long <- cps_state_month_panel |>
  pivot_longer(
    cols = all_of(cps_analysis_specs$variable),
    names_to = "variable",
    values_to = "value"
  ) |>
  left_join(cps_analysis_specs, by = "variable") |>
  filter(is.finite(value))

cps_summary_long <- cps_panel_long |>
  group_by(variable, label, date, p10_color, p50_color, p90_color) |>
  summarise(
    p10 = quantile(value, probs = 0.10, na.rm = TRUE),
    p50 = quantile(value, probs = 0.50, na.rm = TRUE),
    p90 = quantile(value, probs = 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  pivot_longer(
    cols = c(p10, p50, p90),
    names_to = "stat",
    values_to = "value"
  ) |>
  mutate(
    stat_label = recode(stat, p10 = "P10", p50 = "P50", p90 = "P90"),
    stat_color = case_when(
      stat == "p10" ~ p10_color,
      stat == "p50" ~ p50_color,
      stat == "p90" ~ p90_color
    )
  )

cps_annual_avg <- cps_panel_long |>
  mutate(year = year(date)) |>
  group_by(variable, label, year, state, state_abbr) |>
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop")

build_cps_time_series_plot <- function(var_id) {
  state_data <- cps_panel_long |>
    filter(variable == var_id)

  summary_data <- cps_summary_long |>
    filter(variable == var_id)

  label_text <- cps_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)

  p <- plot_ly()

  p <- p |>
    add_lines(
      data = state_data,
      x = ~date,
      y = ~value,
      split = ~state,
      line = list(color = "rgba(150, 150, 150, 0.22)", width = 0.6),
      hoverinfo = "skip",
      showlegend = FALSE
    )

  for (stat_name in c("p10", "p50", "p90")) {
    stat_data <- summary_data |>
      filter(stat == stat_name)

    p <- p |>
      add_lines(
        data = stat_data,
        x = ~date,
        y = ~value,
        name = stat_data$stat_label[[1]],
        line = list(
          color = stat_data$stat_color[[1]],
          width = if (stat_name == "p50") 2.6 else 1.8
        ),
        text = ~paste0(
          label_text, "<br>",
          stat_label, ": ", percent(value, accuracy = 0.1), "<br>",
          "Date: ", format(date, "%Y-%m-%d")
        ),
        hovertemplate = "%{text}<extra></extra>",
        showlegend = TRUE
      )
  }

  p |>
    layout(
      hovermode = "x unified",
      legend = list(orientation = "h", x = 0, y = 1.12),
      margin = list(l = 60, r = 20, t = 40, b = 40),
      xaxis = list(
        title = "",
        rangeslider = list(visible = TRUE)
      ),
      yaxis = list(title = label_text, tickformat = ".1%")
    )
}

build_cps_map_plot <- function(var_id) {
  map_data <- cps_annual_avg |>
    filter(variable == var_id)

  label_text <- cps_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)
  map_label_text <- cps_analysis_specs |>
    filter(variable == var_id) |>
    pull(map_label)

  plot_ly(
    data = map_data,
    type = "choropleth",
    locationmode = "USA-states",
    locations = ~state_abbr,
    z = ~value,
    frame = ~year,
    colorscale = list(
      c(0.0, "#0b3c5d"),
      c(0.5, "#f3e5d8"),
      c(1.0, "#7f0000")
    ),
    colorbar = list(
      title = paste0(map_label_text, "<br>NA = gray"),
      tickformat = ".1%",
      x = 1.02,
      xanchor = "left",
      y = 0.5,
      yanchor = "middle",
      len = 0.6,
      thickness = 12
    ),
    marker = list(line = list(color = "white", width = 0.6)),
    text = ~paste0(
      state, "<br>",
      "Year: ", year, "<br>",
      label_text, ": ", percent(value, accuracy = 0.1)
    ),
    hovertemplate = "%{text}<extra></extra>"
  ) |>
    layout(
      margin = list(l = 0, r = 60, t = 24, b = 0),
      annotations = list(
        list(
          x = 1.02,
          y = 0.84,
          xref = "paper",
          yref = "paper",
          text = "NA = gray",
          showarrow = FALSE,
          font = list(size = 10, color = "#555555")
        )
      ),
      geo = list(
        scope = "usa",
        projection = list(type = "albers usa"),
        showlakes = FALSE,
        bgcolor = "rgba(0,0,0,0)"
      )
    )
}

build_cps_summary_table <- function(var_id) {
  label_text <- cps_analysis_specs |>
    filter(variable == var_id) |>
    pull(label)

  cps_panel_long |>
    filter(variable == var_id) |>
    summarise(
      States = n_distinct(state),
      `Monthly Observations` = n(),
      `Start Month` = min(date),
      `End Month` = max(date),
      Mean = mean(value, na.rm = TRUE),
      SD = sd(value, na.rm = TRUE),
      Min = min(value, na.rm = TRUE),
      P10 = quantile(value, probs = 0.10, na.rm = TRUE),
      P50 = quantile(value, probs = 0.50, na.rm = TRUE),
      P90 = quantile(value, probs = 0.90, na.rm = TRUE),
      Max = max(value, na.rm = TRUE),
      .groups = "drop"
    ) |>
    gt() |>
    fmt_date(columns = c(`Start Month`, `End Month`), date_style = 10) |>
    fmt_percent(columns = c(Mean, SD, Min, P10, P50, P90, Max), decimals = 1) |>
    tab_header(title = label_text) |>
    tab_options(
      table.font.size = px(11),
      column_labels.font.weight = "bold",
      column_labels.background.color = "#2c3e50"
    ) |>
    tab_style(
      style = cell_text(color = "white"),
      locations = cells_column_labels()
    )
}
```

```{r cps-occupation-partition-check}
#| tbl-cap: "Check that the three broad occupation groups are mutually exclusive and exhaustive in the saved CPS panel."

occupation_partition_check <- cps_state_month_panel |>
  transmute(
    occupation_share_total = white_collar_share + blue_collar_share + other_services_share
  ) |>
  summarise(
    `Minimum Sum` = min(occupation_share_total, na.rm = TRUE),
    `P1 Sum` = quantile(occupation_share_total, 0.01, na.rm = TRUE),
    `Median Sum` = quantile(occupation_share_total, 0.50, na.rm = TRUE),
    `P99 Sum` = quantile(occupation_share_total, 0.99, na.rm = TRUE),
    `Maximum Sum` = max(occupation_share_total, na.rm = TRUE)
  ) |>
  mutate(
    `Minimum Sum` = percent(`Minimum Sum`, accuracy = 0.1),
    `P1 Sum` = percent(`P1 Sum`, accuracy = 0.1),
    `Median Sum` = percent(`Median Sum`, accuracy = 0.1),
    `P99 Sum` = percent(`P99 Sum`, accuracy = 0.1),
    `Maximum Sum` = percent(`Maximum Sum`, accuracy = 0.1)
  )

occupation_partition_check |>
  gt() |>
  tab_source_note(
    source_note = md("The build pipeline enforces a row-level integrity check before aggregation: each unemployed record must fall into exactly one of `white_collar`, `blue_collar`, or `other_services`, where `other_services = service + other_occ`.")
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## New Entrant Share

::: {.panel-tabset}

### Time Series

```{r cps-new-entrant-timeseries}
build_cps_time_series_plot("new_entrant_share")
```

### Interactive Map

```{r cps-new-entrant-map}
build_cps_map_plot("new_entrant_share")
```

### Summary Statistics

```{r cps-new-entrant-summary}
build_cps_summary_table("new_entrant_share")
```

:::

## Foreign-Born Share

::: {.panel-tabset}

### Time Series

```{r cps-foreign-born-timeseries}
build_cps_time_series_plot("foreign_born_share")
```

### Interactive Map

```{r cps-foreign-born-map}
build_cps_map_plot("foreign_born_share")
```

### Summary Statistics

```{r cps-foreign-born-summary}
build_cps_summary_table("foreign_born_share")
```

:::

## Noncitizen Share

::: {.panel-tabset}

### Time Series

```{r cps-noncitizen-timeseries}
build_cps_time_series_plot("noncitizen_share")
```

### Interactive Map

```{r cps-noncitizen-map}
build_cps_map_plot("noncitizen_share")
```

### Summary Statistics

```{r cps-noncitizen-summary}
build_cps_summary_table("noncitizen_share")
```

:::

## White-Collar Share

::: {.panel-tabset}

### Time Series

```{r cps-white-collar-timeseries}
build_cps_time_series_plot("white_collar_share")
```

### Interactive Map

```{r cps-white-collar-map}
build_cps_map_plot("white_collar_share")
```

### Summary Statistics

```{r cps-white-collar-summary}
build_cps_summary_table("white_collar_share")
```

:::

## Other-Services Share

::: {.panel-tabset}

### Time Series

```{r cps-other-services-timeseries}
build_cps_time_series_plot("other_services_share")
```

### Interactive Map

```{r cps-other-services-map}
build_cps_map_plot("other_services_share")
```

### Summary Statistics

```{r cps-other-services-summary}
build_cps_summary_table("other_services_share")
```

:::

## Blue-Collar Share

::: {.panel-tabset}

### Time Series

```{r cps-blue-collar-timeseries}
build_cps_time_series_plot("blue_collar_share")
```

### Interactive Map

```{r cps-blue-collar-map}
build_cps_map_plot("blue_collar_share")
```

### Summary Statistics

```{r cps-blue-collar-summary}
build_cps_summary_table("blue_collar_share")
```

:::

## Permanent Job Loser Share

::: {.panel-tabset}

### Time Series

```{r cps-perm-job-loser-timeseries}
build_cps_time_series_plot("perm_job_loser_share")
```

### Interactive Map

```{r cps-perm-job-loser-map}
build_cps_map_plot("perm_job_loser_share")
```

### Summary Statistics

```{r cps-perm-job-loser-summary}
build_cps_summary_table("perm_job_loser_share")
```

:::

## Temporary Layoff Share

::: {.panel-tabset}

### Time Series

```{r cps-temp-layoff-timeseries}
build_cps_time_series_plot("temp_layoff_share")
```

### Interactive Map

```{r cps-temp-layoff-map}
build_cps_map_plot("temp_layoff_share")
```

### Summary Statistics

```{r cps-temp-layoff-summary}
build_cps_summary_table("temp_layoff_share")
```

:::

## White-Collar Permanent Job Loser Share

::: {.panel-tabset}

### Time Series

```{r cps-wc-perm-job-loser-timeseries}
build_cps_time_series_plot("white_collar_perm_job_loser_share")
```

### Interactive Map

```{r cps-wc-perm-job-loser-map}
build_cps_map_plot("white_collar_perm_job_loser_share")
```

### Summary Statistics

```{r cps-wc-perm-job-loser-summary}
build_cps_summary_table("white_collar_perm_job_loser_share")
```

:::

{{< include _ui_cps_profiles.qmd >}}

# Final Merged Panel

The final merged panel combines the saved FRED outcome-side panel with the saved CPS composition-side panel. The purpose is to produce a single analysis-ready dataset that can be estimated at multiple time aggregations depending on the tradeoff between identification and precision.

The monthly panel preserves the most within-state time variation, the quarterly panel provides a middle ground, and the yearly panel substantially improves the precision of the CPS composition estimates. All three are saved under `data/final`, and this memo now refreshes them during render so the downstream analysis stays synchronized with the current source panels.

The aggregation procedure differs by data source because the inputs have different statistical meaning:

- **CPS composition variables** are rebuilt directly from the microdata at each frequency. Monthly panels group by `YEAR, MONTH`; quarterly panels regroup the underlying person records by `YEAR, quarter`; annual panels regroup by `YEAR`. So quarterly and annual shares are recomputed from pooled unemployed persons, not formed by averaging monthly shares.
- **FRED outcome variables** start from the saved monthly state panel. When moving from month to quarter or year, `initial_claims` is treated as a **flow** and therefore summed across months within the period. By contrast, `continued_claims`, `insured_unemployment_rate`, `covered_employment`, `unemployed_persons`, and `unemployment_rate` are treated as **stock/rate-style monthly measures** and are averaged across months within the period. `annual_population` is carried from the annual state population series and repeated within year for monthly and quarterly analysis.
- The merged quarterly and annual panels are created only after those frequency-specific CPS and FRED aggregations are complete. Derived outcomes such as `initial_claims_per_covered_employment` and `continued_claims_per_covered_employment` are then computed from the aggregated ingredients.

## Final Panel Outputs

```{r rebuild-final-panels}
#| echo: false

invisible(NULL)
```

```{r}
#| tbl-cap: "Saved final merged panel outputs."

tibble(
  File = c(
    "ui_takeup_panel_state_month.parquet",
    "ui_takeup_panel_state_month_analysis_ready.parquet",
    "ui_takeup_panel_state_quarter.parquet",
    "ui_takeup_panel_state_quarter_analysis_ready.parquet",
    "ui_takeup_panel_state_year.parquet",
    "ui_takeup_panel_state_year_analysis_ready.parquet",
    "cps_month_precision.parquet",
    "cps_quarter_precision.parquet",
    "cps_year_precision.parquet",
    "cps_precision_comparison.parquet"
  ),
  Purpose = c(
    "Full merged state-month panel",
    "State-month panel restricted to overlap observations with all key variables present",
    "Full merged state-quarter panel",
    "State-quarter panel restricted to overlap observations with all key variables present",
    "Full merged state-year panel",
    "State-year panel restricted to overlap observations with all key variables present",
    "State-month CPS precision diagnostics",
    "State-quarter CPS precision diagnostics",
    "State-year CPS precision diagnostics",
    "Cross-frequency comparison of CPS precision"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    File ~ px(260),
    Purpose ~ px(430)
  ) |>
  tab_source_note(
    source_note = md("All files are saved in the local `data/final` directory.")
  )
```

## Dataset Previews

The tables below show the first few rows of each saved panel so the reader can see the exact structure that is being carried forward into estimation.

```{r final-panel-preview-helpers}
preview_final_dataset = function(filename) {
  read_final_parquet(filename) |>
    head() |>
    DT::datatable(
      rownames = FALSE,
      options = list(
        pageLength = 6,
        dom = "tip",
        scrollX = TRUE
      )
    )
}
```

::: {.panel-tabset}

### State-Month Full

```{r final-preview-month-full}
preview_final_dataset("ui_takeup_panel_state_month.parquet")
```

### State-Month Analysis Ready

```{r final-preview-month-ar}
preview_final_dataset("ui_takeup_panel_state_month_analysis_ready.parquet")
```

### State-Quarter Full

```{r final-preview-quarter-full}
preview_final_dataset("ui_takeup_panel_state_quarter.parquet")
```

### State-Quarter Analysis Ready

```{r final-preview-quarter-ar}
preview_final_dataset("ui_takeup_panel_state_quarter_analysis_ready.parquet")
```

### State-Year Full

```{r final-preview-year-full}
preview_final_dataset("ui_takeup_panel_state_year.parquet")
```

### State-Year Analysis Ready

```{r final-preview-year-ar}
preview_final_dataset("ui_takeup_panel_state_year_analysis_ready.parquet")
```

:::

## CPS Statistical Power

The key source of statistical noise in the merged panel is the CPS composition side. For a weighted state-period share $\hat{p}_{st}$, the memo uses the Kish-style effective sample size

$$
n^{\text{eff}}_{st} = \frac{\left(\sum_{i \in st} w_i\right)^2}{\sum_{i \in st} w_i^2},
$$

and then approximates sampling precision with

$$
\text{SE}(\hat{p}_{st}) \approx \sqrt{\frac{\hat{p}_{st}(1-\hat{p}_{st})}{n^{\text{eff}}_{st}}}
\qquad \text{and} \qquad
\text{MOE}_{95}(\hat{p}_{st}) \approx 1.96 \times \text{SE}(\hat{p}_{st}).
$$

This makes the panel tradeoff transparent: moving from month to quarter or year raises $n^{\text{eff}}_{st}$ by pooling more CPS records, and the sampling noise falls roughly at the rate $1 / \sqrt{n^{\text{eff}}_{st}}$.

```{r}
#| tbl-cap: "Approximate CPS precision by aggregation frequency."

read_final_parquet("cps_precision_comparison.parquet") |>
  mutate(
    Measure = recode(
      measure,
      new_entrant_share = "New entrant share",
      foreign_born_share = "Foreign-born share",
      noncitizen_share = "Noncitizen share"
    ),
    Frequency = str_to_title(frequency)
  ) |>
  select(
    Frequency,
    Measure,
    Periods = periods,
    `Median Eff. N` = median_effective_n,
    `P10 Eff. N` = p10_effective_n,
    `Median MOE (95%)` = median_moe95,
    `P90 MOE (95%)` = p90_moe95,
    `Median Unweighted Numerator N` = median_unweighted_numerator,
    `P10 Unweighted Numerator N` = p10_unweighted_numerator
  ) |>
  gt() |>
  fmt_number(
    columns = c(`Median Eff. N`, `P10 Eff. N`, `Median Unweighted Numerator N`, `P10 Unweighted Numerator N`),
    decimals = 1
  ) |>
  fmt_percent(
    columns = c(`Median MOE (95%)`, `P90 MOE (95%)`),
    decimals = 1
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Frequency ~ px(90),
    Measure ~ px(150),
    Periods ~ px(80),
    `Median Eff. N` ~ px(95),
    `P10 Eff. N` ~ px(90),
    `Median MOE (95%)` ~ px(110),
    `P90 MOE (95%)` ~ px(110),
    `Median Unweighted Numerator N` ~ px(135),
    `P10 Unweighted Numerator N` ~ px(130)
  )
```

```{r}
#| tbl-cap: "Frequency-level tradeoff summary using the CPS precision diagnostics."

read_final_parquet("cps_precision_comparison.parquet") |>
  group_by(frequency) |>
  summarise(
    `Median Eff. N` = median(median_effective_n),
    `Median MOE (95%)` = median(median_moe95),
    .groups = "drop"
  ) |>
  mutate(
    Frequency = str_to_title(frequency),
    month_eff_n = `Median Eff. N`[frequency == "month"],
    `SE Relative To Month` = sqrt(month_eff_n / `Median Eff. N`)
  ) |>
  select(Frequency, `Median Eff. N`, `Median MOE (95%)`, `SE Relative To Month`) |>
  filter(!is.na(`SE Relative To Month`)) |>
  gt() |>
  fmt_number(columns = `Median Eff. N`, decimals = 1) |>
  fmt_percent(columns = c(`Median MOE (95%)`, `SE Relative To Month`), decimals = 1) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

In practical terms, the monthly CPS panel is usable but noisy, especially in smaller states. The quarter panel raises the typical effective sample size by about a factor of three, so the sampling error falls to roughly $\sqrt{45/135} \approx 0.58$ of the monthly level. The year panel raises the typical effective sample size to a bit above 500, so the sampling error falls to roughly $\sqrt{45/526} \approx 0.29$ of the monthly level. That is why the quarterly and yearly panels are valuable robustness versions even if the monthly panel remains attractive for identification.

# Methodology

The merged panel combines the FRED state series with CPS-based composition measures at monthly, quarterly, and yearly frequencies. The CPS side is the part of the design where finite-sample precision matters most, because each state-period estimate is built from unemployed CPS respondents rather than from administrative census-style counts.

The `effective_n` variable is therefore not the literal number of sampled unemployed respondents. Instead, it is the **equal-weight sample size that would deliver the same precision as the weighted CPS cell actually observed**:

$$
n^{\text{eff}}_{st} = \frac{\left(\sum_{i \in st} w_i\right)^2}{\sum_{i \in st} w_i^2}.
$$

If all survey weights were equal, then $n^{\text{eff}}_{st}$ would equal the raw respondent count. When weights are unequal, $n^{\text{eff}}_{st}$ is smaller than the raw count because highly unequal weights reduce precision.

For a CPS share such as the new-entrant share, noncitizen share, blue-collar share, or working-age 25-54 share, the memo uses

$$
\hat{p}_{st} = \frac{\sum_{i \in st} w_i \cdot 1\{X_i = 1\}}{\sum_{i \in st} w_i}
$$

and then approximates uncertainty with

$$
\text{SE}(\hat{p}_{st}) \approx \sqrt{\frac{\hat{p}_{st}(1-\hat{p}_{st})}{n^{\text{eff}}_{st}}},
\qquad
\text{MOE}_{95}(\hat{p}_{st}) \approx 1.96 \times \text{SE}(\hat{p}_{st}).
$$

These are approximation formulas for relative precision, not design-based CPS standard errors. They account for weight inequality via the Kish effective sample size but do not capture the full CPS variance structure from stratification and clustering, so they may be somewhat optimistic. They are still informative about the tradeoff across panel frequencies. Aggregating from state-month to state-quarter to state-year increases $n^{\text{eff}}_{st}$ and shrinks the expected noise roughly in proportion to $1 / \sqrt{n^{\text{eff}}_{st}}$. That is the core reason the yearly panel is much more stable than the monthly panel, even though it sacrifices short-run time variation.

# Best Version Of The Migrant Story

At monthly frequency, the strongest available versions of the migrant channel are:

1. **Foreign-born share of the unemployed**
2. **Noncitizen share of the unemployed**

These are not perfect measures of UI ineligibility, but they are plausible and measurable. By contrast, a strict "recent arrival from abroad" story is much harder to implement cleanly in a monthly state panel using CPS alone.

That means the best empirical framing is probably:

- **primary composition channel**: new entrants
- **baseline migrant control**: noncitizen unemployment share
- **later robustness version**: foreign-born unemployment share

This is a stronger design than trying to force a monthly "recent migrant" variable that the data do not naturally support, and it matches the current regression code, which uses **noncitizen share** in the baseline specification.

# Recommended Baseline Panel

```{r}
#| tbl-cap: "Recommended baseline panel design."

tibble(
  Component = c(
    "Unit of observation",
    "Time frequency",
    "Outcomes",
    "Baseline control 1",
    "Baseline control 2",
    "Baseline control 3",
    "Baseline control 4",
    "Fixed effects",
    "Weighting",
    "Near-term next step"
  ),
  Choice = c(
    "State x year-month",
    "Monthly (with quarterly and yearly robustness panels)",
    "CC/CE, IC/CE, and IUR",
    "New entrant share of unemployed",
    "Noncitizen share of unemployed",
    "Blue-collar share of unemployed",
    "Working-age (25-54) share of unemployed",
    "State FE + date FE; SEs clustered by state",
    "Annual state population, repeated within year for monthly and quarterly panels",
    "Pandemic-window robustness using the same baseline controls"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Component)
  )
```

# Interpretation And Limits

A regression of UI take-up on the current baseline controls would be informative, but it should be interpreted carefully.

What it can say:

- whether states/months with more **new entrants** have lower measured UI take-up
- whether states/months with larger **noncitizen** unemployment shares also have lower measured take-up
- whether states/months with more **blue-collar unemployment** differ systematically in claims intensity
- whether states/months with a larger **prime-age (25-54) unemployment share** differ systematically in claims intensity

What it cannot say by itself:

- whether lower take-up is caused by **severance packages**
- whether low claiming among white-collar workers is the dominant mechanism

Those stories are still plausible, but they likely require either:

- occupation/industry composition interactions, or
- separate evidence on severance, claim hassle, or expected duration

**Measurement error and attenuation bias.** Because the CPS composition regressors in the current baseline (new entrant share, noncitizen share, blue-collar share, and working-age 25-54 share) are estimated from survey microdata rather than observed from administrative records, they are measured with nontrivial sampling noise. This creates a classical-style errors-in-variables problem that will tend to **attenuate slope estimates toward zero**, with the bias most severe in small states and at monthly frequency where CPS cell sizes are smallest. If coefficients are larger and more stable at quarterly or yearly frequency, that pattern is *consistent with* reduced measurement error rather than a different underlying relationship. The quarterly and yearly panels are therefore not merely cosmetic robustness checks — they are an important diagnostic for whether monthly noise is washing out real associations. Formal measurement-error corrections (IV, SIMEX) are deferred to the estimation stage; the design memo treats multi-frequency comparison as the primary guard against attenuation.

**Pandemic-era comparability.** The sample covers March 2020 onward, a period of temporary federal UI programs (PUA, PEUC, FPUC), eligibility expansions, administrative disruptions, and unusual CPS measurement conditions. These distortions affect both the outcome (claims ratios) and the composition regressors. The baseline analysis therefore reports a pandemic-robustness comparison that excludes March 2020 through December 2021 in the monthly panel, 2020 Q1 through 2021 Q4 in the quarterly panel, and calendar years 2020-2021 in the annual panel.

# Claims Hassle And Administrative Burden

The **claims-hassle** story is conceptually plausible, but the main challenge is measurement. FRED and IPUMS CPS do not appear to provide a clean, long monthly state panel of administrative burden.

That means the burden channel is best treated in one of three ways:

1. **State-level burden index**
   Construct or import a cross-state measure of UI administrative burden, such as:
   - online-only versus multi-channel application
   - availability of phone or in-person filing
   - language accessibility
   - waiting week
   - weekly recertification complexity
   - identity-verification burden

2. **Time-varying policy variables**
   If available, add state-by-year or state-by-month variation in:
   - waiting weeks
   - benefit duration
   - replacement rates
   - verification requirements

3. **Interpretive channel rather than core regressor**
   Treat hassle as a background mechanism and focus the baseline regressions on measurable composition channels, then discuss administrative burden as an omitted but important driver.

The most realistic first-pass empirical strategy is therefore:

- use **state fixed effects** to absorb persistent differences in administrative burden
- add any time-varying UI policy variables if they can be assembled
- discuss claims hassle as an important but imperfectly measured mechanism

## Candidate Burden Measures

```{r}
#| tbl-cap: "Practical ways to incorporate claims hassle into the design."

tibble(
  Approach = c(
    "State burden index",
    "Time-varying UI policy controls",
    "Administrative outcomes",
    "Narrative / interpretation"
  ),
  `What It Captures` = c(
    "Persistent cross-state differences in application difficulty",
    "Changes in formal rules affecting claiming incentives or burden",
    "Drop-off, denials, first-payment conversion, processing speed",
    "Qualitative explanation for residual low take-up"
  ),
  `Monthly Panel Feasible` = c(
    "Partly",
    "Possibly",
    "Usually no, not from standard FRED/IPUMS inputs",
    "Yes"
  ),
  `Recommended Use` = c(
    "Secondary regressor or interaction term",
    "Preferred if reliable state-time panel can be built",
    "Future extension if better administrative data are obtained",
    "Include in interpretation even if not directly measured"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

# Occupation Composition And Interaction Terms

The white-collar severance story is better tested through **occupation-composition controls and interactions** than through verbal interpretation alone.

The logic is straightforward:

- **new entrant share** is a proxy for weak UI eligibility
- **migrant share** is a proxy for potentially weaker attachment or eligibility
- **occupation composition** is a proxy for claiming behavior conditional on job loss

If white-collar workers are less likely to claim UI because of severance, expected rapid re-employment, or lower willingness to navigate the claiming process, then take-up should be lower in states/months where unemployment is more concentrated in professional and managerial occupations.

IPUMS CPS gives a workable path to this because occupation and class-of-worker variables are available in the monthly microdata and can be aggregated to state-month composition measures.

## Occupation Group Definitions

For the broad occupation controls, the key identification point is simple: the panel now uses a three-way occupation partition:

- `white_collar_share`
- `blue_collar_share`
- `other_services_share`

White-collar and blue-collar are defined to match the discontinued BLS National Compensation Survey Table 2D groupings as closely as possible in `OCC1990`: white collar covers professional / technical, executive / administrative / managerial, sales, and administrative support occupations; blue collar covers precision production, craft, repair, machine operators / inspectors / assemblers, transportation / material moving, and handlers / helpers / laborers. `other_services_share` is then defined as **everyone else**. These three groups are constructed to be **mutually exclusive and exhaustive** for the unemployed sample. In the fuller occupation-control regressions below, **Other Services is the omitted reference category**.

```{r}
#| tbl-cap: "Plain-language definitions for the broad occupation groups used in the regressions."

tibble(
  `Regression Variable` = c(
    "`white_collar_share`",
    "`blue_collar_share`",
    "`other_services_share`"
  ),
  `Broad Occupation Types Included` = c(
    "Professional / technical, executive / administrative / managerial, sales, and administrative support occupations",
    "Precision production, craft, repair, machine operators / inspectors / assemblers, transportation / material moving, and handlers / helpers / laborers",
    "Everyone else: BLS service occupations plus residual NIU / military / no-usable-prior-occupation cases"
  ),
  Examples = c(
    "Managers, accountants, lawyers, engineers, teachers, computer workers, technicians, sales workers, clerks, and administrative assistants",
    "Electricians, carpenters, mechanics, machinists, assemblers, truck drivers, warehouse movers, equipment cleaners, helpers, and laborers",
    "Police and firefighters, cooks and servers, janitors, nursing aides, childcare workers, hairstylists, household workers, NIU-coded cases, and military occupations"
  ),
  `Regression Role` = c(
    "Estimated relative to Other Services",
    "Estimated relative to Other Services",
    "Omitted reference category"
  )
) |>
  gt() |>
  tab_source_note(
    source_note = md("White-collar and blue-collar definitions follow the discontinued BLS National Compensation Survey Table 2D groupings, with `other_services_share` defined as the residual category: [BLS Table 2D](https://www.bls.gov/eci/factsheets/national-compensation-survey-classification-systems-mapping-files.htm).")
  ) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    `Regression Variable` ~ px(140),
    `Broad Occupation Types Included` ~ px(300),
    Examples ~ px(260),
    `Regression Role` ~ px(140)
  )
```

```{r}
#| tbl-cap: "Exact OCC1990-to-broad-occupation crosswalk used in the memo."

cps_occ_codebook <- read_final_parquet("cps_occ1990_broad_codebook.parquet")

datatable(
  cps_occ_codebook,
  rownames = FALSE,
  filter = "top",
  options = list(pageLength = 20, scrollX = TRUE),
  caption = htmltools::tags$caption(
    style = "caption-side: top; text-align: left;",
    "Each OCC1990 code, its IPUMS occupation label, and the broad category assignment used in the memo."
  )
)
```

## Useful Occupation And Work-Background Measures

```{r}
#| tbl-cap: "Candidate occupation and work-composition measures from monthly CPS."

tibble(
  Measure = c(
    "Blue-collar share of unemployed",
    "White-collar share of unemployed",
    "Other-services share of unemployed",
    "Private-sector share of unemployed with prior job",
    "Professional / managerial share of permanent job losers",
    "Professional / managerial share of temporary layoffs"
  ),
  `Why It Helps` = c(
    "Captures construction / manufacturing / production exposure",
    "Captures professional, technical, managerial, sales, and administrative-support unemployment concentration",
    "Provides the omitted comparison group for the broad occupation split",
    "Separates private-sector from other prior-job backgrounds",
    "Targets the severance/claim-propensity story more directly",
    "Helps distinguish furlough-style layoffs from permanent separations"
  ),
  `Use In Regression` = c(
    "Control",
    "Preferred next extension",
    "Omitted reference category",
    "Control",
    "Preferred interaction variable",
    "Robustness / mechanism"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Why Interactions Matter

A main-effect occupation variable tells you whether take-up is systematically lower when unemployment is more white-collar. An interaction term goes one step further and asks whether the effect of one composition channel depends on the occupational mix.

Examples:

- If the **migrant share** coefficient becomes more negative when white-collar unemployment is high, that suggests low claiming among migrant-heavy unemployment is especially pronounced in states/months where displaced workers are more white-collar.
- If the **permanent-job-loser share** becomes less predictive of claims when professional unemployment is high, that is consistent with a severance or lower-claim-propensity mechanism among otherwise job-attached workers.

The cleanest interaction is probably not with all unemployed persons, but with **job-loser composition**. That keeps the interpretation closer to the set of people who are plausibly UI-eligible.

# Preferred Regression Extensions

The implemented baseline model is:

$$
\text{takeup}_{st} =
\beta_1 \text{NewEntrantShare}_{st} +
\beta_2 \text{NoncitizenShare}_{st} +
\beta_3 \text{BlueCollarShare}_{st} +
\beta_4 \text{WorkingAgeShare}_{st} +
\alpha_s + \gamma_t + \varepsilon_{st}
$$

This is a deliberately conservative first step:

- `new_entrant_share` gets at places with more first-time job seekers and turnover near the edge of UI eligibility
- `noncitizen_share` gets at places with larger immigrant unemployment shares and potentially larger access or eligibility frictions
- `blue_collar_share` gets at shifts in industry and occupation mix
- `working_age_25_54_share` gets at age composition in the unemployed pool

All specifications are estimated with **annual state population weights**, repeated within year for monthly and quarterly panels.

The next useful extension, once this baseline has been digested, is:

$$
\text{takeup}_{st} =
\beta_1 \text{NewEntrantShare}_{st} +
\beta_2 \text{NoncitizenShare}_{st} +
\beta_3 \text{BlueCollarShare}_{st} +
\beta_4 \text{WorkingAgeShare}_{st} +
\beta_5 \text{WhiteCollarShare}_{st} +
\alpha_s + \gamma_t + \varepsilon_{st}
$$

where `WorkingAgeShare` refers to the unemployed share ages **25-54**. This fuller occupation split omits `OtherServicesShare`, so the coefficients on `BlueCollarShare` and `WhiteCollarShare` are both interpreted relative to the residual Other Services group.

## Empirical Priority Order

```{r}
#| tbl-cap: "Suggested sequence for moving from baseline to mechanism tests."

tibble(
  Step = c(
    "1. Baseline composition model",
    "2. Pandemic robustness",
    "3. Add white-collar share",
    "4. Add job-loser mechanism variables",
    "5. Add UI policy / burden controls"
  ),
  Purpose = c(
    "Estimate the weighted four-control baseline: new entrants, noncitizens, blue-collar share, and prime-age share",
    "Check whether the baseline survives excluding the pandemic-distortion window",
    "Then estimate the fuller three-way occupation split by adding white-collar share, so that white-collar and blue-collar unemployment are both interpreted relative to omitted Other Services",
    "Then probe severance / claim-propensity channels through permanent job losers, temporary layoffs, and white-collar permanent job losers",
    "Only after that, separate composition effects from administrative-friction effects"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

# Bottom Line

The project is empirically feasible with currently available data.

The cleanest version of the design is:

- build a **monthly state panel**
- use a **covered-employment-normalized claims outcome** as the baseline administrative scale
- construct **new entrant share** from `WHYUNEMP`
- use **noncitizen share** as the baseline migrant proxy
- control for **blue-collar share** and **working-age (25-54) share**
- estimate models with **state FE and date FE**, clustering standard errors by state and weighting by annual state population

The strongest immediate next step is therefore to move from conceptual discussion to data assembly and estimation — which follows below.

# Regression Results

This section now focuses on a single baseline specification. All regressions use **state fixed effects** and **date fixed effects** with **standard errors clustered by state**. The two reported outcomes are:

- **CC/CE**: continued claims / covered employment
- **IC/CE**: initial claims / covered employment

`IUR` is omitted from the regression tables because it is redundant with `CC/CE` in this setup. Each reported outcome is estimated at **monthly**, **quarterly**, and **annual** frequency. The estimation sample is restricted to observations through **December 2025** and then further restricted to **balanced panels** with all 50 states present in every included month, quarter, or year.

```{r regression-setup}
library(fixest)
library(modelsummary)

panel_month_raw = read_final_parquet("ui_takeup_panel_state_month_analysis_ready.parquet") |>
  mutate(
    date = as.Date(period_start),
    year_month = paste0(year, "-", sprintf("%02d", month)),
    post_2020 = as.integer(year >= 2021L),
    post_2022 = as.integer(year >= 2023L)
  )

panel_quarter_raw = read_final_parquet("ui_takeup_panel_state_quarter_analysis_ready.parquet") |>
  mutate(
    date = as.Date(period_start),
    year_quarter = paste0(year, "-Q", quarter),
    post_2020 = as.integer(year >= 2021L),
    post_2022 = as.integer(year >= 2023L)
  )

panel_year_raw = read_final_parquet("ui_takeup_panel_state_year_analysis_ready.parquet") |>
  mutate(
    date = as.Date(period_start),
    post_2020 = as.integer(year >= 2021L),
    post_2022 = as.integer(year >= 2023L)
  )

sample_end_date = as.Date("2025-12-31")
expected_states = nrow(state_meta)

keep_balanced_periods = function(data, period_var) {
  balanced_periods = data |>
    count(.data[[period_var]], name = "state_count") |>
    filter(state_count == expected_states) |>
    pull(.data[[period_var]])

  data |>
    filter(.data[[period_var]] %in% balanced_periods)
}

panel_month_cutoff = panel_month_raw |>
  filter(date <= sample_end_date)

panel_quarter_cutoff = panel_quarter_raw |>
  filter(date <= sample_end_date)

panel_year_cutoff = panel_year_raw |>
  filter(date <= sample_end_date)

panel_month = panel_month_cutoff |>
  keep_balanced_periods("year_month")

panel_quarter = panel_quarter_cutoff |>
  keep_balanced_periods("year_quarter")

panel_year = panel_year_cutoff |>
  keep_balanced_periods("year")

baseline_rhs = "new_entrant_share + noncitizen_share + blue_collar_share + working_age_25_54_share"

fit_baseline = function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ",
        baseline_rhs,
        " | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

msummary_styled = function(models, title, notes) {
  msummary(
    models,
    stars = c("*" = 0.10, "**" = 0.05, "***" = 0.01),
    gof_map = c("nobs", "r.squared", "adj.r.squared", "within.r.squared"),
    title = title,
    notes = notes,
    output = "gt"
  ) |>
    tab_options(table.font.size = px(11))
}
```

## Baseline Controls

The baseline control set is intentionally simple and maps directly to the mechanisms discussed above.

```{r}
#| tbl-cap: "Baseline controls carried in every regression."

tibble(
  Control = c(
    "`new_entrant_share`",
    "`noncitizen_share`",
    "`blue_collar_share`",
    "`working_age_25_54_share`"
  ),
  `What It Proxies For` = c(
    "Places with more first-time job seekers and more labor-market turnover at the margin of UI eligibility",
    "Places with larger immigrant unemployment shares and potentially larger access / eligibility frictions",
    "Changes in industry / occupation mix toward blue-collar unemployment",
    "Age composition of unemployment, captured as the share of unemployed workers ages 25 to 54"
  )
) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Control ~ px(160),
    `What It Proxies For` ~ px(520)
  )
```

## Panel Balance Checks

```{r regression-balance-checks}
balance_summary <- bind_rows(
  tibble(
    Frequency = "Monthly",
    Sample = c("Through 2025 cutoff", "Balanced through 2025"),
    States = c(n_distinct(panel_month_cutoff$state), n_distinct(panel_month$state)),
    Periods = c(n_distinct(panel_month_cutoff$year_month), n_distinct(panel_month$year_month)),
    Rows = c(nrow(panel_month_cutoff), nrow(panel_month)),
    `Expected Rows` = c(
      n_distinct(panel_month_cutoff$year_month) * expected_states,
      n_distinct(panel_month$year_month) * expected_states
    )
  ),
  tibble(
    Frequency = "Quarterly",
    Sample = c("Through 2025 cutoff", "Balanced through 2025"),
    States = c(n_distinct(panel_quarter_cutoff$state), n_distinct(panel_quarter$state)),
    Periods = c(n_distinct(panel_quarter_cutoff$year_quarter), n_distinct(panel_quarter$year_quarter)),
    Rows = c(nrow(panel_quarter_cutoff), nrow(panel_quarter)),
    `Expected Rows` = c(
      n_distinct(panel_quarter_cutoff$year_quarter) * expected_states,
      n_distinct(panel_quarter$year_quarter) * expected_states
    )
  ),
  tibble(
    Frequency = "Annual",
    Sample = c("Through 2025 cutoff", "Balanced through 2025"),
    States = c(n_distinct(panel_year_cutoff$state), n_distinct(panel_year$state)),
    Periods = c(n_distinct(panel_year_cutoff$year), n_distinct(panel_year$year)),
    Rows = c(nrow(panel_year_cutoff), nrow(panel_year)),
    `Expected Rows` = c(
      n_distinct(panel_year_cutoff$year) * expected_states,
      n_distinct(panel_year$year) * expected_states
    )
  )
) |>
  mutate(
    Balanced = if_else(Rows == `Expected Rows`, "Yes", "No"),
    `Rows Lost vs Cutoff` = c(
      0L,
      nrow(panel_month_cutoff) - nrow(panel_month),
      0L,
      nrow(panel_quarter_cutoff) - nrow(panel_quarter),
      0L,
      nrow(panel_year_cutoff) - nrow(panel_year)
    )
  )

balance_summary |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Baseline Estimates

```{r baseline-models}
ccu_mo_base <- fit_baseline("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_base <- fit_baseline("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_base <- fit_baseline("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_base <- fit_baseline("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_base <- fit_baseline("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_base <- fit_baseline("initial_claims_per_covered_employment", panel_year, "year")
```

The tables below use the full-sample versus pandemic-exclusion comparison as the main presentation of the baseline estimates, since that contains the baseline coefficients and the robustness check in one place. The pandemic period involved extraordinary temporary UI expansions and administrative disruption, so the second column for each frequency excludes the main pandemic-distortion window.

```{r pandemic-samples}
mo_nopan <- panel_month |>
  filter(date < as.Date("2020-03-01") | date > as.Date("2021-12-01")) |>
  keep_balanced_periods("year_month")

qr_nopan <- panel_quarter |>
  filter(date < as.Date("2020-01-01") | date > as.Date("2022-01-01")) |>
  keep_balanced_periods("year_quarter")

yr_nopan <- panel_year |>
  filter(!year %in% c(2020L, 2021L)) |>
  keep_balanced_periods("year")

ccu_nopan_mo <- fit_baseline("continued_claims_per_covered_employment", mo_nopan, "year_month")
ccu_nopan_qr <- fit_baseline("continued_claims_per_covered_employment", qr_nopan, "year_quarter")
ccu_nopan_yr <- fit_baseline("continued_claims_per_covered_employment", yr_nopan, "year")

icu_nopan_mo <- fit_baseline("initial_claims_per_covered_employment", mo_nopan, "year_month")
icu_nopan_qr <- fit_baseline("initial_claims_per_covered_employment", qr_nopan, "year_quarter")
icu_nopan_yr <- fit_baseline("initial_claims_per_covered_employment", yr_nopan, "year")
```

::: {.panel-tabset}

### CC/CE

```{r reg-pandemic-ccu}
msummary_styled(
  list(
    "Monthly Full" = ccu_mo_base,
    "Monthly Excl. Pandemic" = ccu_nopan_mo,
    "Quarterly Full" = ccu_qr_base,
    "Quarterly Excl. Pandemic" = ccu_nopan_qr,
    "Annual Full" = ccu_yr_base,
    "Annual Excl. Pandemic" = ccu_nopan_yr
  ),
  title = "Baseline Estimates and Pandemic Robustness: Continued Claims / Covered Employment",
  notes = "Monthly robustness excludes March 2020 through December 2021. Quarterly robustness excludes 2020 Q1 through 2021 Q4. Annual robustness excludes 2020 and 2021. The control set is unchanged, and all regressions remain weighted by annual state population."
)
```

### IC/CE

```{r reg-pandemic-icu}
msummary_styled(
  list(
    "Monthly Full" = icu_mo_base,
    "Monthly Excl. Pandemic" = icu_nopan_mo,
    "Quarterly Full" = icu_qr_base,
    "Quarterly Excl. Pandemic" = icu_nopan_qr,
    "Annual Full" = icu_yr_base,
    "Annual Excl. Pandemic" = icu_nopan_yr
  ),
  title = "Baseline Estimates and Pandemic Robustness: Initial Claims / Covered Employment",
  notes = "Monthly robustness excludes March 2020 through December 2021. Quarterly robustness excludes 2020 Q1 through 2021 Q4. Annual robustness excludes 2020 and 2021. The control set is unchanged, and all regressions remain weighted by annual state population."
)
```

:::

# Expected UI Takeup by Cause

## Annual By-Cause Composition Extension

Before moving to occupation mix and interaction mechanisms, it is useful to ask a more direct question: in the **state-year panel**, does UI take-up vary systematically with the **reason-for-unemployment composition** itself? The specification below uses the annual panel only, because the full six-way by-cause split is most credible after pooling CPS observations within year. `perm_job_loser_share` is omitted, so each coefficient is interpreted relative to a higher permanent-job-loser share.

```{r annual-bycause-extension-models}
annual_bycause_rhs <- paste(
  "temp_layoff_share",
  "completed_temp_job_share",
  "job_leaver_share",
  "reentrant_share",
  "new_entrant_share",
  sep = " + "
)

fit_annual_bycause_extension <- function(outcome, data) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ",
        annual_bycause_rhs,
        " | state + year"
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_yr_bycause <- fit_annual_bycause_extension("continued_claims_per_covered_employment", panel_year)
icu_yr_bycause <- fit_annual_bycause_extension("initial_claims_per_covered_employment", panel_year)

ccu_yr_bycause_nopan <- fit_annual_bycause_extension("continued_claims_per_covered_employment", yr_nopan)
icu_yr_bycause_nopan <- fit_annual_bycause_extension("initial_claims_per_covered_employment", yr_nopan)
```

::: {.panel-tabset}

### CC/CE

```{r annual-bycause-extension-ccce}
msummary_styled(
  list(
    "Annual Full" = ccu_yr_bycause,
    "Annual Excl. Pandemic" = ccu_yr_bycause_nopan
  ),
  title = "Annual By-Cause Composition Extension: Continued Claims / Covered Employment",
  notes = "Annual state-year regressions with state and year fixed effects, state-clustered standard errors, and annual state population weights. Included regressors are temporary layoff, completed temporary job, job leaver, reentrant, and new-entrant shares. `perm_job_loser_share` is omitted, so coefficients are interpreted relative to permanent job losers."
)
```

### IC/CE

```{r annual-bycause-extension-icce}
msummary_styled(
  list(
    "Annual Full" = icu_yr_bycause,
    "Annual Excl. Pandemic" = icu_yr_bycause_nopan
  ),
  title = "Annual By-Cause Composition Extension: Initial Claims / Covered Employment",
  notes = "Annual state-year regressions with state and year fixed effects, state-clustered standard errors, and annual state population weights. Included regressors are temporary layoff, completed temporary job, job leaver, reentrant, and new-entrant shares. `perm_job_loser_share` is omitted, so coefficients are interpreted relative to permanent job losers."
)
```

:::

This annual by-cause extension is the closest state-panel analogue to the national descriptive regression. If `job_leaver_share`, `reentrant_share`, or `new_entrant_share` come in well below the omitted permanent-job-loser category, that is consistent with lower claims intensity in unemployment pools that are less tightly connected to prior job attachment. If `temp_layoff_share` is positive relative to permanent job losers, that would indicate stronger claims intensity in recall-style unemployment than in permanent separations.

## Residual And Robustness Checks

Because `perm_job_loser_share` is omitted, the coefficients above are **conditional contrasts relative to permanent job losers**, not raw unconditional correlations. A negative coefficient therefore means "lower take-up than permanent job losers, holding the other by-cause shares fixed," not necessarily that the raw relationship with take-up is negative. The checks below ask four concrete questions:

1. Do the residuals show obvious misspecification or a few extreme state-year observations?
2. Do the signs change sharply when the fixed-effects structure changes?
3. Are the coefficients driven by one or two influential states?
4. How tightly do the included by-cause shares move together in the annual panel?

```{r annual-bycause-diagnostic-setup}
annual_bycause_terms <- c(
  "temp_layoff_share",
  "completed_temp_job_share",
  "job_leaver_share",
  "reentrant_share",
  "new_entrant_share"
)

annual_bycause_term_labels <- c(
  temp_layoff_share = "Temporary layoff share",
  completed_temp_job_share = "Completed temp-job share",
  job_leaver_share = "Job leaver share",
  reentrant_share = "Reentrant share",
  new_entrant_share = "New entrant share"
)

safe_sd <- function(x) {
  s <- sd(x, na.rm = TRUE)
  if (!is.finite(s) || s == 0) 1 else s
}

build_annual_bycause_diag_df <- function(model, data, outcome_var, outcome_label, sample_label) {
  fitted_vals <- predict(model, newdata = data)
  observed_vals <- data[[outcome_var]]

  tibble(
    sample = sample_label,
    outcome = outcome_label,
    state = data$state,
    year = data$year,
    annual_population = data$annual_population,
    observed = observed_vals,
    fitted = fitted_vals,
    residual = observed_vals - fitted_vals
  ) |>
    mutate(
      std_residual = residual / safe_sd(residual),
      abs_residual = abs(residual)
    )
}

annual_bycause_diag_df <- bind_rows(
  build_annual_bycause_diag_df(
    ccu_yr_bycause,
    panel_year,
    "continued_claims_per_covered_employment",
    "CC/CE",
    "Annual Full"
  ),
  build_annual_bycause_diag_df(
    ccu_yr_bycause_nopan,
    yr_nopan,
    "continued_claims_per_covered_employment",
    "CC/CE",
    "Annual Excl. Pandemic"
  ),
  build_annual_bycause_diag_df(
    icu_yr_bycause,
    panel_year,
    "initial_claims_per_covered_employment",
    "IC/CE",
    "Annual Full"
  ),
  build_annual_bycause_diag_df(
    icu_yr_bycause_nopan,
    yr_nopan,
    "initial_claims_per_covered_employment",
    "IC/CE",
    "Annual Excl. Pandemic"
  )
)

largest_resid_points <- annual_bycause_diag_df |>
  group_by(sample, outcome) |>
  slice_max(order_by = abs_residual, n = 1, with_ties = FALSE) |>
  ungroup() |>
  transmute(
    sample,
    outcome,
    `Largest Residual State-Year` = paste0(state, " ", year),
    `Largest Abs Residual` = abs_residual
  )

annual_bycause_resid_summary <- annual_bycause_diag_df |>
  group_by(sample, outcome) |>
  summarise(
    `Weighted RMSE` = sqrt(weighted.mean(residual^2, annual_population, na.rm = TRUE)),
    `Weighted MAE` = weighted.mean(abs_residual, annual_population, na.rm = TRUE),
    `Corr(Observed, Fitted)` = cor(observed, fitted, use = "complete.obs"),
    `Mean Residual` = weighted.mean(residual, annual_population, na.rm = TRUE),
    `P90 Abs Residual` = quantile(abs_residual, 0.90, na.rm = TRUE),
    .groups = "drop"
  ) |>
  left_join(largest_resid_points, by = c("sample", "outcome"))

fit_annual_bycause_extension_fe_variant <- function(outcome, data, fe_variant) {
  formula_text <- switch(
    fe_variant,
    "No FE" = paste0(outcome, " ~ ", annual_bycause_rhs),
    "State FE only" = paste0(outcome, " ~ ", annual_bycause_rhs, " | state"),
    "Year FE only" = paste0(outcome, " ~ ", annual_bycause_rhs, " | year"),
    "State + Year FE" = paste0(outcome, " ~ ", annual_bycause_rhs, " | state + year")
  )

  feols(
    as.formula(formula_text),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_yr_bycause_no_fe <- fit_annual_bycause_extension_fe_variant(
  "continued_claims_per_covered_employment",
  panel_year,
  "No FE"
)
ccu_yr_bycause_state_fe <- fit_annual_bycause_extension_fe_variant(
  "continued_claims_per_covered_employment",
  panel_year,
  "State FE only"
)
ccu_yr_bycause_year_fe <- fit_annual_bycause_extension_fe_variant(
  "continued_claims_per_covered_employment",
  panel_year,
  "Year FE only"
)

icu_yr_bycause_no_fe <- fit_annual_bycause_extension_fe_variant(
  "initial_claims_per_covered_employment",
  panel_year,
  "No FE"
)
icu_yr_bycause_state_fe <- fit_annual_bycause_extension_fe_variant(
  "initial_claims_per_covered_employment",
  panel_year,
  "State FE only"
)
icu_yr_bycause_year_fe <- fit_annual_bycause_extension_fe_variant(
  "initial_claims_per_covered_employment",
  panel_year,
  "Year FE only"
)

extract_annual_bycause_coefs <- function(model) {
  tibble(
    term = annual_bycause_terms,
    estimate = unname(coef(model)[annual_bycause_terms])
  ) |>
    mutate(Term = recode(term, !!!annual_bycause_term_labels)) |>
    select(term, Term, estimate)
}

summarise_leave_one_state_out <- function(outcome_var, data, baseline_model, outcome_label, sample_label) {
  baseline_tbl <- extract_annual_bycause_coefs(baseline_model) |>
    rename(Baseline = estimate)

  loo_tbl <- map_dfr(sort(unique(data$state)), \(st) {
    model_drop <- fit_annual_bycause_extension(outcome_var, filter(data, state != st))
    extract_annual_bycause_coefs(model_drop) |>
      rename(estimate = estimate) |>
      mutate(`Dropped State` = st)
  }) |>
    left_join(baseline_tbl, by = c("term", "Term")) |>
    mutate(abs_shift = abs(estimate - Baseline))

  loo_tbl |>
    group_by(Term) |>
    summarise(
      Outcome = outcome_label,
      Sample = sample_label,
      Baseline = first(Baseline),
      Min = min(estimate, na.rm = TRUE),
      Median = median(estimate, na.rm = TRUE),
      Max = max(estimate, na.rm = TRUE),
      `Share Same Sign` = mean(sign(estimate) == sign(first(Baseline)), na.rm = TRUE),
      `Largest Abs Shift` = max(abs_shift, na.rm = TRUE),
      `Dropped State w/ Largest Shift` = `Dropped State`[which.max(abs_shift)],
      .groups = "drop"
    ) |>
    select(
      Outcome, Sample, Term, Baseline, Min, Median, Max,
      `Share Same Sign`, `Largest Abs Shift`, `Dropped State w/ Largest Shift`
    )
}

annual_bycause_loo_summary <- bind_rows(
  summarise_leave_one_state_out(
    "continued_claims_per_covered_employment",
    panel_year,
    ccu_yr_bycause,
    "CC/CE",
    "Annual Full"
  ),
  summarise_leave_one_state_out(
    "continued_claims_per_covered_employment",
    yr_nopan,
    ccu_yr_bycause_nopan,
    "CC/CE",
    "Annual Excl. Pandemic"
  ),
  summarise_leave_one_state_out(
    "initial_claims_per_covered_employment",
    panel_year,
    icu_yr_bycause,
    "IC/CE",
    "Annual Full"
  ),
  summarise_leave_one_state_out(
    "initial_claims_per_covered_employment",
    yr_nopan,
    icu_yr_bycause_nopan,
    "IC/CE",
    "Annual Excl. Pandemic"
  )
)

build_pairwise_corr_table <- function(data, sample_label) {
  pair_list <- combn(annual_bycause_terms, 2, simplify = FALSE)

  map_dfr(pair_list, \(vars) {
    tibble(
      Sample = sample_label,
      `Share Pair` = paste(
        annual_bycause_term_labels[[vars[[1]]]],
        "vs.",
        annual_bycause_term_labels[[vars[[2]]]]
      ),
      Correlation = cor(data[[vars[[1]]]], data[[vars[[2]]]], use = "complete.obs")
    )
  }) |>
    arrange(desc(abs(Correlation)))
}

annual_bycause_pairwise_corr <- bind_rows(
  build_pairwise_corr_table(panel_year, "Annual Full"),
  build_pairwise_corr_table(yr_nopan, "Annual Excl. Pandemic")
)

estimatr_available <- requireNamespace("estimatr", quietly = TRUE)

fit_annual_bycause_rlm <- function(outcome, data) {
  MASS::rlm(
    as.formula(
      paste0(
        outcome,
        " ~ ",
        annual_bycause_rhs,
        " + factor(state) + factor(year)"
      )
    ),
    data = data,
    weights = annual_population,
    maxit = 100
  )
}

extract_annual_bycause_rlm <- function(model) {
  coef_mat <- summary(model)$coefficients

  tibble(
    term = rownames(coef_mat),
    estimate = coef_mat[, "Value"],
    std_error = coef_mat[, "Std. Error"],
    t_value = coef_mat[, "t value"]
  ) |>
    filter(term %in% annual_bycause_terms) |>
    mutate(Term = recode(term, !!!annual_bycause_term_labels)) |>
    select(term, Term, estimate, std_error, t_value)
}

ccu_yr_bycause_rlm <- fit_annual_bycause_rlm(
  "continued_claims_per_covered_employment",
  panel_year
)
ccu_yr_bycause_nopan_rlm <- fit_annual_bycause_rlm(
  "continued_claims_per_covered_employment",
  yr_nopan
)
icu_yr_bycause_rlm <- fit_annual_bycause_rlm(
  "initial_claims_per_covered_employment",
  panel_year
)
icu_yr_bycause_nopan_rlm <- fit_annual_bycause_rlm(
  "initial_claims_per_covered_employment",
  yr_nopan
)

build_annual_bycause_outlier_compare <- function(feols_model, rlm_model, outcome_label, sample_label) {
  fe_tbl <- summary(feols_model)$coeftable |>
    as.data.frame() |>
    rownames_to_column("term") |>
    as_tibble() |>
    filter(term %in% annual_bycause_terms) |>
    transmute(
      term,
      `FEOLS Estimate` = Estimate,
      `FEOLS p-value` = `Pr(>|t|)`
    )

  rlm_tbl <- extract_annual_bycause_rlm(rlm_model) |>
    transmute(
      term,
      `RLM Estimate` = estimate,
      `RLM t-stat` = t_value
    )

  fe_tbl |>
    left_join(rlm_tbl, by = "term") |>
    mutate(
      Outcome = outcome_label,
      Sample = sample_label,
      Term = recode(term, !!!annual_bycause_term_labels),
      `Same Sign` = sign(`FEOLS Estimate`) == sign(`RLM Estimate`)
    ) |>
    select(
      Outcome, Sample, Term,
      `FEOLS Estimate`, `FEOLS p-value`,
      `RLM Estimate`, `RLM t-stat`,
      `Same Sign`
    )
}

annual_bycause_outlier_compare <- bind_rows(
  build_annual_bycause_outlier_compare(
    ccu_yr_bycause,
    ccu_yr_bycause_rlm,
    "CC/CE",
    "Annual Full"
  ),
  build_annual_bycause_outlier_compare(
    ccu_yr_bycause_nopan,
    ccu_yr_bycause_nopan_rlm,
    "CC/CE",
    "Annual Excl. Pandemic"
  ),
  build_annual_bycause_outlier_compare(
    icu_yr_bycause,
    icu_yr_bycause_rlm,
    "IC/CE",
    "Annual Full"
  ),
  build_annual_bycause_outlier_compare(
    icu_yr_bycause_nopan,
    icu_yr_bycause_nopan_rlm,
    "IC/CE",
    "Annual Excl. Pandemic"
  )
)
```

### Residual Diagnostics

```{r fig-annual-bycause-residuals}
#| fig-cap: "Residual-versus-fitted checks for the annual by-cause extension. Point size is proportional to annual state population."
#| fig-height: 6.5

ggplot(annual_bycause_diag_df, aes(fitted, residual, size = annual_population)) +
  geom_hline(yintercept = 0, color = "grey65", linewidth = 0.4) +
  geom_point(alpha = 0.35, color = "#0073C2") +
  geom_smooth(method = "loess", se = FALSE, color = "black", linewidth = 0.6) +
  facet_grid(outcome ~ sample, scales = "free") +
  scale_x_continuous(labels = percent_format(accuracy = 0.01)) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  labs(
    title = "Annual By-Cause Extension: Residuals vs. Fitted",
    subtitle = "A flat smooth around zero is the main visual check",
    x = "Fitted value",
    y = "Residual"
  ) +
  theme_report +
  theme(legend.position = "none")
```

```{r annual-bycause-residual-summary}
#| tbl-cap: "Residual summary for the annual by-cause extension."

annual_bycause_resid_summary |>
  mutate(
    across(c(`Weighted RMSE`, `Weighted MAE`, `Mean Residual`, `P90 Abs Residual`, `Largest Abs Residual`),
           \(x) percent(x, accuracy = 0.01)),
    `Corr(Observed, Fitted)` = sprintf("%.3f", `Corr(Observed, Fitted)`)
  ) |>
  rename(Sample = sample, Outcome = outcome) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

### Fixed-Effects Sensitivity

If the negative coefficients are being created by the fixed-effects structure rather than by a stable conditional pattern, they should move sharply when the specification changes from **no FE** to **state FE only**, **year FE only**, and **state + year FE**.

::: {.panel-tabset}

#### CC/CE

```{r annual-bycause-fe-sensitivity-ccce}
msummary_styled(
  list(
    "No FE" = ccu_yr_bycause_no_fe,
    "State FE only" = ccu_yr_bycause_state_fe,
    "Year FE only" = ccu_yr_bycause_year_fe,
    "State + Year FE" = ccu_yr_bycause
  ),
  title = "Fixed-Effects Sensitivity: Continued Claims / Covered Employment",
  notes = "Annual full-sample state-year regressions with annual state population weights and state-clustered standard errors. Comparing these columns shows whether the by-cause coefficients are coming from raw cross-state differences, common year shifts, or within-state within-year variation."
)
```

#### IC/CE

```{r annual-bycause-fe-sensitivity-icce}
msummary_styled(
  list(
    "No FE" = icu_yr_bycause_no_fe,
    "State FE only" = icu_yr_bycause_state_fe,
    "Year FE only" = icu_yr_bycause_year_fe,
    "State + Year FE" = icu_yr_bycause
  ),
  title = "Fixed-Effects Sensitivity: Initial Claims / Covered Employment",
  notes = "Annual full-sample state-year regressions with annual state population weights and state-clustered standard errors. Comparing these columns shows whether the by-cause coefficients are coming from raw cross-state differences, common year shifts, or within-state within-year variation."
)
```

:::

### Outlier-Robust Re-Estimation

The annual residual checks suggest that a few state-year observations, especially around the pandemic and a small number of non-pandemic episodes, deserve explicit robustness treatment. The table below compares the baseline FEOLS estimates to a **Huber-type robust regression** estimated with `MASS::rlm`, using the same by-cause regressors plus explicit state and year dummies.

`estimatr::lm_robust` is not available in the current local R environment, so the outlier-robust check shown here uses `MASS::rlm`.

```{r annual-bycause-outlier-robust-table}
#| tbl-cap: "Annual by-cause coefficients: baseline FEOLS versus outlier-robust `MASS::rlm`."

annual_bycause_outlier_compare |>
  mutate(
    `FEOLS Estimate` = percent(`FEOLS Estimate`, accuracy = 0.01),
    `FEOLS p-value` = sprintf("%.3f", `FEOLS p-value`),
    `RLM Estimate` = percent(`RLM Estimate`, accuracy = 0.01),
    `RLM t-stat` = sprintf("%.2f", `RLM t-stat`),
    `Same Sign` = if_else(`Same Sign`, "Yes", "No")
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(10.5),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

### Leave-One-State-Out Stability

```{r annual-bycause-leave-one-state-out}
#| tbl-cap: "Leave-one-state-out stability of the annual by-cause coefficients."

annual_bycause_loo_summary |>
  mutate(
    across(c(Baseline, Min, Median, Max, `Largest Abs Shift`), \(x) percent(x, accuracy = 0.01)),
    `Share Same Sign` = percent(`Share Same Sign`, accuracy = 1)
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(10.5),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

### Correlation Among Included By-Cause Shares

```{r annual-bycause-share-correlation-table}
#| tbl-cap: "Pairwise correlation among the included annual by-cause shares."

annual_bycause_pairwise_corr |>
  mutate(Correlation = sprintf("%.3f", Correlation)) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

## Annual By-Cause Gap Over Time

The coefficient tables above use **state and year fixed effects**, which is the right specification for asking whether states with different by-cause mixes have different take-up **within the same year**. But that is not the right object for a gap-over-time plot. If year fixed effects are included, the model is allowed to absorb each year's common national take-up shift with a separate year intercept, so the fitted values are no longer forcing the by-cause composition to explain the aggregate time path.

For the gap-over-time exercise, the memo therefore uses a deliberately more descriptive specification:

$$
\text{takeup}_{st} =
\beta_1 \text{TempLayoffShare}_{st} +
\beta_2 \text{CompletedTempJobShare}_{st} +
\beta_3 \text{JobLeaverShare}_{st} +
\beta_4 \text{ReentrantShare}_{st} +
\beta_5 \text{NewEntrantShare}_{st} +
\alpha_s + \varepsilon_{st},
$$

estimated on the **annual state-year panel** with **state fixed effects** and **annual state population weights**, but **without year fixed effects**. Permanent job losers remain the omitted reference category.

The yearly gap is then constructed in three steps:

1. Estimate the state-year model above and compute a fitted value for every state-year.
2. Within each calendar year, take the **population-weighted mean** of the observed state take-up values and the **population-weighted mean** of the fitted state take-up values.
3. Define the gap as **observed minus expected**. A negative gap means the by-cause composition is **overpredicting** take-up in that year; a positive gap means it is **underpredicting** take-up.

```{r annual-bycause-gap-models}
fit_annual_bycause_gap_model <- function(outcome, data) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ",
        annual_bycause_rhs,
        " | state"
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_yr_bycause_gap_model <- fit_annual_bycause_gap_model(
  "continued_claims_per_covered_employment",
  panel_year
)
icu_yr_bycause_gap_model <- fit_annual_bycause_gap_model(
  "initial_claims_per_covered_employment",
  panel_year
)

panel_year_gap <- panel_year |>
  mutate(
    cc_ce_hat_bycause = predict(ccu_yr_bycause_gap_model, newdata = panel_year),
    ic_ce_hat_bycause = predict(icu_yr_bycause_gap_model, newdata = panel_year)
  )

annual_gap_summary <- panel_year_gap |>
  group_by(year) |>
  summarise(
    observed_cc_ce = weighted.mean(continued_claims_per_covered_employment, annual_population, na.rm = TRUE),
    expected_cc_ce = weighted.mean(cc_ce_hat_bycause, annual_population, na.rm = TRUE),
    cc_ce_gap = observed_cc_ce - expected_cc_ce,
    observed_ic_ce = weighted.mean(initial_claims_per_covered_employment, annual_population, na.rm = TRUE),
    expected_ic_ce = weighted.mean(ic_ce_hat_bycause, annual_population, na.rm = TRUE),
    ic_ce_gap = observed_ic_ce - expected_ic_ce,
    .groups = "drop"
  ) |>
  mutate(date = as.Date(paste0(year, "-01-01")))

recent_annual_gap_summary <- annual_gap_summary |>
  filter(year >= 2023L) |>
  summarise(
    `Average CC/CE Gap Since 2023` = mean(cc_ce_gap, na.rm = TRUE),
    `Average IC/CE Gap Since 2023` = mean(ic_ce_gap, na.rm = TRUE),
    `Latest Year` = max(year),
    `Latest CC/CE Gap` = cc_ce_gap[which.max(year)],
    `Latest IC/CE Gap` = ic_ce_gap[which.max(year)]
  )
```

```{r fig-annual-bycause-gap-fit}
#| fig-height: 6.5
#| fig-cap: "Population-weighted annual observed versus by-cause-predicted take-up from a state-year model with state fixed effects only. Because year fixed effects are omitted here, the fitted line is forced to track the time path through changes in state unemployment composition rather than through free year-specific intercepts."

p_cc_annual_gap_fit <- ggplot(annual_gap_summary, aes(date)) +
  geom_line(aes(y = observed_cc_ce, color = "Observed"), linewidth = 0.9) +
  geom_line(aes(y = expected_cc_ce, color = "Expected from by-cause mix"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Observed" = "#0073C2", "Expected from by-cause mix" = "black")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "CC/CE",
    subtitle = "Population-weighted state-year average",
    x = NULL, y = NULL
  )

p_ic_annual_gap_fit <- ggplot(annual_gap_summary, aes(date)) +
  geom_line(aes(y = observed_ic_ce, color = "Observed"), linewidth = 0.9) +
  geom_line(aes(y = expected_ic_ce, color = "Expected from by-cause mix"), linewidth = 0.8, linetype = "dashed") +
  scale_color_manual(values = c("Observed" = "#EFC000", "Expected from by-cause mix" = "black")) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "IC/CE",
    subtitle = "Population-weighted state-year average",
    x = NULL, y = NULL
  )

p_cc_annual_gap_fit + p_ic_annual_gap_fit +
  plot_layout(guides = "collect") &
  theme(legend.position = "top")
```

```{r fig-annual-bycause-gap-bars}
#| fig-height: 6.5
#| fig-cap: "Population-weighted annual gap between observed and by-cause-predicted take-up. Negative values mean the by-cause composition is overpredicting take-up; positive values mean it is underpredicting take-up."

p_cc_annual_gap <- ggplot(annual_gap_summary, aes(date, cc_ce_gap)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_col(fill = "#0073C2", alpha = 0.85, width = 300) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "CC/CE Gap",
    subtitle = "Observed minus expected from by-cause mix",
    x = NULL, y = NULL
  )

p_ic_annual_gap <- ggplot(annual_gap_summary, aes(date, ic_ce_gap)) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.5) +
  geom_col(fill = "#EFC000", alpha = 0.9, width = 300) +
  scale_y_continuous(labels = percent_format(accuracy = 0.01)) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "IC/CE Gap",
    subtitle = "Observed minus expected from by-cause mix",
    x = NULL, y = NULL
  )

p_cc_annual_gap + p_ic_annual_gap
```

```{r annual-bycause-gap-summary-table}
#| tbl-cap: "Recent annual gap between observed and by-cause-predicted take-up from the state-year descriptive model."

recent_annual_gap_summary |>
  mutate(
    `Average CC/CE Gap Since 2023` = percent(`Average CC/CE Gap Since 2023`, accuracy = 0.01),
    `Average IC/CE Gap Since 2023` = percent(`Average IC/CE Gap Since 2023`, accuracy = 0.01),
    `Latest CC/CE Gap` = percent(`Latest CC/CE Gap`, accuracy = 0.01),
    `Latest IC/CE Gap` = percent(`Latest IC/CE Gap`, accuracy = 0.01)
  ) |>
  gt() |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

These plots should be read as a **descriptive accounting exercise**, not the main causal specification. The identifying question is simple: if you feed the changing state-year unemployment composition into a model that only adjusts for persistent state differences, does that composition path predict more or less take-up than the data actually show? Persistent negative gaps would mean the by-cause mix is overpredicting take-up over time, which is exactly the pattern one would expect if a separate claiming-propensity shortfall has opened up.

# Regression Extensions

## Occupation Extension

This extension adds `white_collar_share` to the baseline while keeping `blue_collar_share` in the model. The omitted occupation category is therefore `other_services_share`, so the white-collar and blue-collar coefficients are both interpreted relative to Other Services.

```{r occupation-extension-models}
fit_occupation_extension <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + white_collar_share | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_occ <- fit_occupation_extension("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_occ <- fit_occupation_extension("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_occ <- fit_occupation_extension("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_occ <- fit_occupation_extension("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_occ <- fit_occupation_extension("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_occ <- fit_occupation_extension("initial_claims_per_covered_employment", panel_year, "year")
```

::: {.panel-tabset}

### CC/CE

```{r occupation-extension-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_occ, "Quarterly" = ccu_qr_occ, "Annual" = ccu_yr_occ),
  title = "Occupation Extension with Other Services Omitted: Continued Claims / Covered Employment",
  notes = "All models include state and date fixed effects with state-clustered standard errors and annual state population weights. `blue_collar_share` remains in the model, `white_collar_share` is added, and the omitted occupation category is `other_services_share`."
)
```

- `new_entrant_share` is negative at monthly, quarterly, and annual frequency, with the quarterly estimate especially informative because pooling months likely reduces CPS measurement noise.
- `white_collar_share` is also negative and statistically significant across all three frequencies, indicating lower continuing-claims take-up where unemployment is more white collar, relative to the omitted Other Services category.
- `working_age_25_54_share` is positive, which is consistent with higher take-up where the unemployed pool contains more prime-age workers who are more plausibly attached to the labor market and therefore more likely to claim UI.
- `noncitizen_share` remains close to zero and statistically insignificant, suggesting little evidence that take-up variation is being driven by the migrant-eligibility channel, at least as captured by this proxy.
- `blue_collar_share` is not precisely different from Other Services once `white_collar_share` is included, suggesting the main occupation result is really about lower take-up in white-collar unemployment rather than unusually high take-up in blue-collar unemployment.

### IC/CE

```{r occupation-extension-icce}
msummary_styled(
  list("Monthly" = icu_mo_occ, "Quarterly" = icu_qr_occ, "Annual" = icu_yr_occ),
  title = "Occupation Extension with Other Services Omitted: Initial Claims / Covered Employment",
  notes = "All models include state and date fixed effects with state-clustered standard errors and annual state population weights. `blue_collar_share` remains in the model, `white_collar_share` is added, and the omitted occupation category is `other_services_share`."
)
```

- `new_entrant_share` is much smaller for `IC/CE` and is not a stable statistical predictor across frequencies, which complicates a simple work-history ineligibility story for initial claims.
- `noncitizen_share` remains close to zero and statistically insignificant, suggesting little evidence that migrant-related eligibility frictions are a major driver of variation in initial claims take-up in this specification.
- `blue_collar_share` is also muted once `white_collar_share` is included, indicating that the occupation result is not really about unusually high initial claiming in blue-collar unemployment.
- `working_age_25_54_share` weakens as the panel is aggregated, which is striking because it suggests that larger prime-age shares of the unemployed do not robustly predict more initial claiming.
- One possible interpretation is that, because this outcome is **initial claims**, prime-age workers with relatively strong labor-market attachment may still delay or forgo filing if they expect to find work quickly.
- `white_collar_share` is the clearest negative signal, especially at annual frequency, which is consistent with a low-takeup story rooted in severance, lower filing urgency, or easier expected reemployment among white-collar workers.

:::

## Interactions: Claim-Propensity Story

The claim-propensity story focuses on unemployed workers with clearer prior attachment. `perm_job_loser_share` is useful here because permanent job losers are the group most plausibly positioned to claim UI if they choose to file. The interaction below asks whether the white-collar shortfall is especially pronounced when unemployment is more concentrated among permanent job losers. A **negative** interaction would be consistent with lower filing propensity, severance, or faster expected reemployment among otherwise attached white-collar job losers.

```{r claim-propensity-interaction-models}
fit_claim_propensity_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + white_collar_share + perm_job_loser_share + perm_job_loser_share:white_collar_share | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_claim <- fit_claim_propensity_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_claim <- fit_claim_propensity_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_claim <- fit_claim_propensity_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_claim <- fit_claim_propensity_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_claim <- fit_claim_propensity_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_claim <- fit_claim_propensity_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

::: {.panel-tabset}

### CC/CE

```{r claim-propensity-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_claim, "Quarterly" = ccu_qr_claim, "Annual" = ccu_yr_claim),
  title = "Claim-Propensity Interaction: Continued Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `perm_job_loser_share`, and `perm_job_loser_share × white_collar_share` to the weighted baseline specification. A negative interaction indicates that the relationship between permanent job losers and take-up is weaker in whiter-collar unemployment pools."
)
```

In the **annual CC/CE** regression, `perm_job_loser_share` is positive and statistically significant, while the interaction `white_collar_share x perm_job_loser_share` is negative and statistically significant. This implies that state-years with larger shares of unemployed workers who are permanent job losers have higher continuing-claims take-up, consistent with stronger prior attachment and a greater propensity to claim UI. But that positive relationship is weaker in state-years where the unemployed pool is more white collar, which fits the idea that even among high-attachment unemployed workers, whiter-collar labor markets generate less continuing-claims take-up.

### IC/CE

```{r claim-propensity-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_claim, "Quarterly" = icu_qr_claim, "Annual" = icu_yr_claim),
  title = "Claim-Propensity Interaction: Initial Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `perm_job_loser_share`, and `perm_job_loser_share × white_collar_share` to the weighted baseline specification. A negative interaction indicates that the relationship between permanent job losers and take-up is weaker in whiter-collar unemployment pools."
)
```

For **IC/CE**, the annual regression points in the same direction, though a bit less cleanly. Higher `perm_job_loser_share` predicts more initial claiming, but the interaction with `white_collar_share` is negative, indicating that this positive relationship is muted in whiter-collar unemployment environments. That pattern is consistent with a story in which even relatively attached white-collar workers file less aggressively, perhaps because of severance, lower filing urgency, or easier expected reemployment.

:::

## Interactions: New Entrants x Permanent Job Losers

This interaction asks whether the new-entrant margin matters differently depending on how much of the unemployed pool consists of **permanent job losers**. A **negative** interaction would mean that the low-take-up association of new entrants is even stronger when the rest of the unemployment pool is more concentrated among plausibly attached job losers.

```{r new-entrant-perm-job-loser-interaction-models}
fit_newentrant_permjobloser_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + perm_job_loser_share + new_entrant_share:perm_job_loser_share | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_newent_perm <- fit_newentrant_permjobloser_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_newent_perm <- fit_newentrant_permjobloser_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_newent_perm <- fit_newentrant_permjobloser_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_newent_perm <- fit_newentrant_permjobloser_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_newent_perm <- fit_newentrant_permjobloser_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_newent_perm <- fit_newentrant_permjobloser_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

:::: {.panel-tabset}

### CC/CE

```{r new-entrant-perm-job-loser-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_newent_perm, "Quarterly" = ccu_qr_newent_perm, "Annual" = ccu_yr_newent_perm),
  title = "New Entrants x Permanent Job Losers: Continued Claims / Covered Employment",
  notes = "Each model adds `perm_job_loser_share` and `new_entrant_share x perm_job_loser_share` to the weighted baseline specification. A negative interaction indicates that the new-entrant penalty is stronger when unemployment is more concentrated among permanent job losers."
)
```

### IC/CE

```{r new-entrant-perm-job-loser-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_newent_perm, "Quarterly" = icu_qr_newent_perm, "Annual" = icu_yr_newent_perm),
  title = "New Entrants x Permanent Job Losers: Initial Claims / Covered Employment",
  notes = "Each model adds `perm_job_loser_share` and `new_entrant_share x perm_job_loser_share` to the weighted baseline specification. A negative interaction indicates that the new-entrant penalty is stronger when unemployment is more concentrated among permanent job losers."
)
```

::::

- From the `new_entrant_share` perspective, a positive interaction means the depressing effect of new entrants on aggregate take-up is smaller when the unemployed pool contains more permanent job losers, because the claimant margin is then driven more by attached workers who are plausibly eligible to file.
- From the `perm_job_loser_share` perspective, the same interaction means the positive take-up effect of permanent job losers is strongest when they are not being offset by a large weak-attachment new-entrant margin; in that sense, the coefficient captures a composition-offset relationship between highly claim-relevant and weakly claim-relevant unemployed workers.
## Interactions: New Entrants x Post-2020

This interaction asks whether the relationship between `new_entrant_share` and UI take-up changed after **2020**. Here `post_2020` equals one for **2021 onward**. A **negative** interaction would mean that the low-take-up association of new entrants became even stronger in the post-2020 period.

```{r new-entrant-post2020-interaction-models}
fit_newentrant_post2020_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + post_2020 + new_entrant_share:post_2020 | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_newent_post <- fit_newentrant_post2020_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_newent_post <- fit_newentrant_post2020_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_newent_post <- fit_newentrant_post2020_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_newent_post <- fit_newentrant_post2020_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_newent_post <- fit_newentrant_post2020_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_newent_post <- fit_newentrant_post2020_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

:::: {.panel-tabset}

### CC/CE

```{r new-entrant-post2020-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_newent_post, "Quarterly" = ccu_qr_newent_post, "Annual" = ccu_yr_newent_post),
  title = "New Entrants x Post-2020: Continued Claims / Covered Employment",
  notes = "Each model adds `post_2020` and `new_entrant_share x post_2020` to the weighted baseline specification. Here `post_2020` is an indicator for 2021 onward. A negative interaction indicates that the new-entrant penalty became stronger after 2020."
)
```

### IC/CE

```{r new-entrant-post2020-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_newent_post, "Quarterly" = icu_qr_newent_post, "Annual" = icu_yr_newent_post),
  title = "New Entrants x Post-2020: Initial Claims / Covered Employment",
  notes = "Each model adds `post_2020` and `new_entrant_share x post_2020` to the weighted baseline specification. Here `post_2020` is an indicator for 2021 onward. A negative interaction indicates that the new-entrant penalty became stronger after 2020."
)
```

::::

## Interactions: New Entrants x Post-2022

This interaction asks whether the relationship between `new_entrant_share` and UI take-up changed after **2022**. Here `post_2022` equals one for **2023 onward**. The motivating question is whether the post-2022 environment, including possible changes in job search and white-collar labor-market dynamics, altered the relationship between new entrants and claiming behavior. A **negative** interaction would mean that the low-take-up association of new entrants became even stronger from 2023 onward.

```{r new-entrant-post2022-interaction-models}
fit_newentrant_post2022_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + post_2022 + new_entrant_share:post_2022 | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_newent_post2022 <- fit_newentrant_post2022_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_newent_post2022 <- fit_newentrant_post2022_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_newent_post2022 <- fit_newentrant_post2022_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_newent_post2022 <- fit_newentrant_post2022_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_newent_post2022 <- fit_newentrant_post2022_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_newent_post2022 <- fit_newentrant_post2022_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

:::: {.panel-tabset}

### CC/CE

```{r new-entrant-post2022-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_newent_post2022, "Quarterly" = ccu_qr_newent_post2022, "Annual" = ccu_yr_newent_post2022),
  title = "New Entrants x Post-2022: Continued Claims / Covered Employment",
  notes = "Each model adds `post_2022` and `new_entrant_share x post_2022` to the weighted baseline specification. Here `post_2022` is an indicator for 2023 onward. A negative interaction indicates that the new-entrant penalty became stronger after 2022."
)
```

### IC/CE

```{r new-entrant-post2022-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_newent_post2022, "Quarterly" = icu_qr_newent_post2022, "Annual" = icu_yr_newent_post2022),
  title = "New Entrants x Post-2022: Initial Claims / Covered Employment",
  notes = "Each model adds `post_2022` and `new_entrant_share x post_2022` to the weighted baseline specification. Here `post_2022` is an indicator for 2023 onward. A negative interaction indicates that the new-entrant penalty became stronger after 2022."
)
```

::::

## Interactions: White-Collar Share x Post-2020

This interaction asks whether the relationship between `white_collar_share` and UI take-up changed after **2020**. Here `post_2020` equals one for **2021 onward**. A **negative** interaction would mean that the low-take-up association of white-collar unemployment became even stronger in the post-2020 period, consistent with pandemic-era shifts in white-collar labor markets (remote work, tech layoffs, reduced filing urgency).

```{r white-collar-post2020-interaction-models}
fit_whitecollar_post2020_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + white_collar_share + post_2020 + white_collar_share:post_2020 | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_wc_post2020 <- fit_whitecollar_post2020_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_wc_post2020 <- fit_whitecollar_post2020_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_wc_post2020 <- fit_whitecollar_post2020_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_wc_post2020 <- fit_whitecollar_post2020_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_wc_post2020 <- fit_whitecollar_post2020_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_wc_post2020 <- fit_whitecollar_post2020_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

:::: {.panel-tabset}

### CC/CE

```{r white-collar-post2020-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_wc_post2020, "Quarterly" = ccu_qr_wc_post2020, "Annual" = ccu_yr_wc_post2020),
  title = "White-Collar Share x Post-2020: Continued Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `post_2020`, and `white_collar_share x post_2020` to the weighted baseline specification. Here `post_2020` is an indicator for 2021 onward. A negative interaction indicates that the white-collar take-up penalty became stronger after 2020."
)
```

### IC/CE

```{r white-collar-post2020-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_wc_post2020, "Quarterly" = icu_qr_wc_post2020, "Annual" = icu_yr_wc_post2020),
  title = "White-Collar Share x Post-2020: Initial Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `post_2020`, and `white_collar_share x post_2020` to the weighted baseline specification. Here `post_2020` is an indicator for 2021 onward. A negative interaction indicates that the white-collar take-up penalty became stronger after 2020."
)
```

::::

## Interactions: White-Collar Share x Post-2022

This interaction asks whether the relationship between `white_collar_share` and UI take-up changed after **2022**. Here `post_2022` equals one for **2023 onward**. The motivating question is whether white-collar labor-market dynamics in the post-2022 period — including tech-sector layoffs and shifting remote-work norms — altered the relationship between white-collar unemployment composition and claiming behavior. A **negative** interaction would mean that the low-take-up association of white-collar unemployment became even stronger from 2023 onward.

```{r white-collar-post2022-interaction-models}
fit_whitecollar_post2022_interaction <- function(outcome, data, time_fe) {
  feols(
    as.formula(
      paste0(
        outcome,
        " ~ ", baseline_rhs, " + white_collar_share + post_2022 + white_collar_share:post_2022 | state + ",
        time_fe
      )
    ),
    data = data,
    weights = ~annual_population,
    cluster = ~state
  )
}

ccu_mo_wc_post2022 <- fit_whitecollar_post2022_interaction("continued_claims_per_covered_employment", panel_month, "year_month")
ccu_qr_wc_post2022 <- fit_whitecollar_post2022_interaction("continued_claims_per_covered_employment", panel_quarter, "year_quarter")
ccu_yr_wc_post2022 <- fit_whitecollar_post2022_interaction("continued_claims_per_covered_employment", panel_year, "year")

icu_mo_wc_post2022 <- fit_whitecollar_post2022_interaction("initial_claims_per_covered_employment", panel_month, "year_month")
icu_qr_wc_post2022 <- fit_whitecollar_post2022_interaction("initial_claims_per_covered_employment", panel_quarter, "year_quarter")
icu_yr_wc_post2022 <- fit_whitecollar_post2022_interaction("initial_claims_per_covered_employment", panel_year, "year")
```

:::: {.panel-tabset}

### CC/CE

```{r white-collar-post2022-interaction-ccce}
msummary_styled(
  list("Monthly" = ccu_mo_wc_post2022, "Quarterly" = ccu_qr_wc_post2022, "Annual" = ccu_yr_wc_post2022),
  title = "White-Collar Share x Post-2022: Continued Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `post_2022`, and `white_collar_share x post_2022` to the weighted baseline specification. Here `post_2022` is an indicator for 2023 onward. A negative interaction indicates that the white-collar take-up penalty became stronger after 2022."
)
```

### IC/CE

```{r white-collar-post2022-interaction-icce}
msummary_styled(
  list("Monthly" = icu_mo_wc_post2022, "Quarterly" = icu_qr_wc_post2022, "Annual" = icu_yr_wc_post2022),
  title = "White-Collar Share x Post-2022: Initial Claims / Covered Employment",
  notes = "Each model adds `white_collar_share`, `post_2022`, and `white_collar_share x post_2022` to the weighted baseline specification. Here `post_2022` is an indicator for 2023 onward. A negative interaction indicates that the white-collar take-up penalty became stronger after 2022."
)
```

::::

# Source Notes

This memo is based on a review of:

- FRED's **State Unemployment Insurance Weekly Claims Report**
- IPUMS CPS variable availability for `WHYUNEMP`, `STATEFIP`, `NATIVITY`, `CITIZEN`, `OCC1990`, and `CLASSWKR`
- IPUMS CPS migration-variable availability, including the limits of using `MIGRATE1` in a monthly design
- All CPS-derived report inputs are produced by `build_panels.R` from the unified CPS extract and saved as parquet files under `data/final`.

Useful source pages:

- [FRED state UI claims release](https://fred.stlouisfed.org/release?rid=469)
- [IPUMS CPS: WHYUNEMP](https://cps.ipums.org/cps-action/variables/WHYUNEMP)
- [IPUMS CPS: STATEFIP](https://cps.ipums.org/cps-action/variables/STATEFIP/)
- [IPUMS CPS: NATIVITY](https://cps.ipums.org/cps-action/variables/NATIVITY)
- [IPUMS CPS: CITIZEN](https://cps.ipums.org/cps-action/variables/CITIZEN)
- [IPUMS CPS: MIGRATE1](https://cps.ipums.org/cps-action/variables/MIGRATE1)
