---
title: "Composition of Unemployment by Reason"
subtitle: "CPS Table A-11 · Post-1994 Redesign · Six-Way Decomposition"
date: today
format:
  html:
    theme: flatly
    toc: true
    toc-depth: 3
    code-fold: true
    code-summary: "Show code"
    fig-width: 10
    fig-height: 5.5
    embed-resources: true
    self-contained-math: true
execute:
  warning: false
  message: false
---

## Overview

The headline unemployment rate tells you *how many* people are unemployed, but not *why*. A rising rate driven by permanent layoffs signals a very different economy than one driven by temporary furloughs, voluntary quits, or new graduates entering the job market for the first time.

The Bureau of Labor Statistics classifies every unemployed person into one of six mutually exclusive categories based on their reason for unemployment: **temporary layoffs** (expecting recall), **permanent job losers** (no recall expected), **completed temporary jobs**, **job leavers** (quit voluntarily), **reentrants** (returning to the labor force after a period out), and **new entrants** (searching for their first job ever). These six groups sum to total unemployment. We also aggregate them into two broader groups — **job separations** (the first four) and **labor force entry** (reentrants + new entrants) — to isolate the demand-side from the supply-side channel.

This report decomposes the unemployment rate along these dimensions to answer a central question: **what is driving changes in unemployment, and how has the composition shifted over time?** Each component's level divided by the civilian labor force gives its contribution to the total unemployment rate; stacking these contributions reconstructs the headline number. The distinction matters for policy: a surge in permanent job losers calls for countercyclical intervention, a spike in temporary layoffs may signal a shock firms expect to reverse, rising job leavers can indicate labor market confidence, and a rise in new entrants may reflect demographic shifts or an expanding economy drawing in first-time workers. By tracking both the rate contributions and the period-over-period changes of each component, we can distinguish between cyclical deterioration, structural shifts, and signs of labor market strength.

```{r setup}
#| label: setup

library(tidyverse)
library(fredr)
library(ggsci)      # for scale_*_jco()
library(scales)
library(gt)
library(lubridate)

# ── FRED API key ─────────────────────────────────────────────────────────
# Set your FRED API key. Get one free at https://fred.stlouisfed.org/docs/api/api_key.html
# Uncomment and set below, or use Sys.setenv() in your .Renviron
fredr_set_key("ccd350785a210a7653a20d02ce92d83a")

# ── Theme ────────────────────────────────────────────────────────────────
theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )

theme_set(theme_report)

# ── Palettes ─────────────────────────────────────────────────────────────
# 6-color palette for the six-way cause decomposition
cause_cols <- pal_jco()(6)
names(cause_cols) <- c("Temporary Layoff", "Permanent Job Losers",
                       "Completed Temp Job", "Job Leavers",
                       "Reentrants", "New Entrants")

# 2-color palette for two-group decomposition
group_cols <- c("Job Separations" = "#0073C2", "Labor Force Entry" = "#EFC000")

# 4-color palette (kept for legacy charts)
jco_cols <- pal_jco()(4)
names(jco_cols) <- c("Job Losers", "Job Leavers", "Reentrants", "New Entrants")

# ── NBER recession dates (post-1994) ────────────────────────────────────
recessions <- tibble(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-01", "2009-06-01", "2020-04-01"))
)

add_recessions <- function(p, alpha = 0.08) {
  p + annotate("rect",
    xmin = recessions$start, xmax = recessions$end,
    ymin = -Inf, ymax = Inf,
    fill = "grey40", alpha = alpha
  )
}
```

```{r data-pull}
#| label: data-pull
#| cache: true

# ── Pull all series from FRED ────────────────────────────────────────────
# Four-way series (legacy)
series_ids_4 <- c(
  "LNS13023621",   # Job Losers (aggregate)
  "LNS13023705",   # Job Leavers
  "LNS13023557",   # Reentrants
  "LNS13023569",   # New Entrants
  "UNEMPLOY"        # Total Unemployed
)
series_names_4 <- c("Job Losers", "Job Leavers", "Reentrants", "New Entrants", "Total")

# Six-way series (granular cause decomposition)
series_ids_6 <- c(
  "LNS13023653",   # On Temporary Layoff
  "LNS13026638",   # Permanent Job Losers
  "LNS13026637",   # Completed Temporary Job
  "LNS13023705",   # Job Leavers
  "LNS13023557",   # Reentrants
  "LNS13023569",   # New Entrants
  "UNEMPLOY",       # Total Unemployed
  "CLF16OV"         # Civilian Labor Force (for rate conversion)
)
series_names_6 <- c("Temporary Layoff", "Permanent Job Losers", "Completed Temp Job",
                     "Job Leavers", "Reentrants", "New Entrants", "Total", "CLF")

# Pull four-way
raw <- map2_dfr(series_ids_4, series_names_4, \(id, nm) {
  fredr(
    series_id = id,
    observation_start = as.Date("1994-01-01"),
    observation_end   = Sys.Date()
  ) |>
    transmute(date, series = nm, value)
})

# Pull six-way
raw6 <- map2_dfr(series_ids_6, series_names_6, \(id, nm) {
  fredr(
    series_id = id,
    observation_start = as.Date("1994-01-01"),
    observation_end   = Sys.Date()
  ) |>
    transmute(date, series = nm, value)
})

# ── Four-way: pivot wide and compute shares ──────────────────────────────
df_wide <- raw |>
  pivot_wider(names_from = series, values_from = value) |>
  drop_na() |>
  mutate(
    component_sum = `Job Losers` + `Job Leavers` + Reentrants + `New Entrants`,
    across(c(`Job Losers`, `Job Leavers`, Reentrants, `New Entrants`),
           \(x) x / component_sum * 100,
           .names = "{.col}_share")
  )

component_order <- c("Job Losers", "Job Leavers", "Reentrants", "New Entrants")

df_shares <- df_wide |>
  select(date, ends_with("_share")) |>
  pivot_longer(-date, names_to = "component", values_to = "share") |>
  mutate(
    component = str_remove(component, "_share"),
    component = factor(component, levels = component_order)
  )

df_levels <- raw |>
  filter(series != "Total") |>
  mutate(series = factor(series, levels = component_order))

# ── Six-way: pivot wide and compute rate contributions ───────────────────
cause_order <- c("Temporary Layoff", "Permanent Job Losers", "Completed Temp Job",
                 "Job Leavers", "Reentrants", "New Entrants")

df6 <- raw6 |>
  pivot_wider(names_from = series, values_from = value) |>
  drop_na() |>
  mutate(
    across(all_of(cause_order),
           \(x) x / CLF * 100,
           .names = "{.col}_rate"),
    urate = Total / CLF * 100
  )

# Long format for stacked bar
df6_long <- df6 |>
  select(date, ends_with("_rate")) |>
  pivot_longer(-date, names_to = "cause", values_to = "rate") |>
  mutate(
    cause = str_remove(cause, "_rate"),
    cause = factor(cause, levels = cause_order)
  )

# ── Two-group aggregation ────────────────────────────────────────────────
group_order <- c("Job Separations", "Labor Force Entry")

df2 <- df6 |>
  mutate(
    `Job Separations` = `Temporary Layoff` + `Permanent Job Losers` +
                         `Completed Temp Job` + `Job Leavers`,
    `Labor Force Entry` = Reentrants + `New Entrants`,
    `Job Separations_rate` = `Job Separations` / CLF * 100,
    `Labor Force Entry_rate` = `Labor Force Entry` / CLF * 100
  )

df2_long <- df2 |>
  select(date, `Job Separations_rate`, `Labor Force Entry_rate`) |>
  pivot_longer(-date, names_to = "group", values_to = "rate") |>
  mutate(
    group = str_remove(group, "_rate"),
    group = factor(group, levels = group_order)
  )
```

## 1. Unemployment Rate by Cause (1994–Present)

::: {.panel-tabset}

### Six-Way

```{r fig-urate-cause-6}
#| label: fig-urate-cause-6
#| fig-cap: "Unemployment rate decomposed by cause (six-way). Bars stack to total unemployment rate."

p_cause <- df6_long |>
  ggplot(aes(x = date, y = rate, fill = cause)) +
  geom_col(alpha = 0.85, width = 28, position = "stack") +
  geom_line(
    data = df6, aes(x = date, y = urate, fill = NULL),
    color = "black", linewidth = 0.4, alpha = 0.5
  ) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Unemployment Rate by Cause",
    subtitle = "Each component = cause level / civilian labor force; bars stack to total unemployment rate",
    y = "Unemployment Rate (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )

add_recessions(p_cause)
```

### Two-Group

```{r fig-urate-cause-2}
#| label: fig-urate-cause-2
#| fig-cap: "Unemployment rate decomposed into job separations vs. labor force entry. Bars stack to total unemployment rate."

p_cause2 <- df2_long |>
  ggplot(aes(x = date, y = rate, fill = group)) +
  geom_col(alpha = 0.85, width = 28, position = "stack") +
  geom_line(
    data = df2, aes(x = date, y = urate, fill = NULL),
    color = "black", linewidth = 0.4, alpha = 0.5
  ) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Unemployment Rate: Job Separations vs. Labor Force Entry",
    subtitle = "Separations = layoffs + completed temp jobs + quits; Entry = reentrants + new entrants",
    y = "Unemployment Rate (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )

add_recessions(p_cause2)
```

:::

The stacked bars show how each cause contributes to the total unemployment rate. Permanent job losers and temporary layoffs dominate during recessions — in 2020, temporary layoffs alone drove the rate above 8% as firms furloughed workers expecting a recall. The two-group view makes clear that job separations drive virtually all cyclical variation, while labor force entry contributes a relatively stable base.

## 1b. Unemployment Rate by Cause — Recent Detail (2023–Present)

::: {.panel-tabset}

### Six-Way

```{r fig-urate-cause-recent-6}
#| label: fig-urate-cause-recent-6
#| fig-cap: "Unemployment rate by cause, zoomed to 2023–present (six-way)."

df6_long |>
  filter(date >= as.Date("2023-01-01")) |>
  ggplot(aes(x = date, y = rate, fill = cause)) +
  geom_col(alpha = 0.85, width = 22, position = "stack") +
  geom_line(
    data = df6 |> filter(date >= as.Date("2023-01-01")),
    aes(x = date, y = urate, fill = NULL),
    color = "black", linewidth = 0.5, alpha = 0.5
  ) +
  geom_point(
    data = df6 |> filter(date >= as.Date("2023-01-01")),
    aes(x = date, y = urate, fill = NULL),
    color = "black", size = 1
  ) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "Unemployment Rate by Cause — Recent Detail",
    subtitle = "Six-way decomposition, 2023–present",
    y = "Unemployment Rate (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )
```

### Two-Group

```{r fig-urate-cause-recent-2}
#| label: fig-urate-cause-recent-2
#| fig-cap: "Unemployment rate: separations vs. entry, zoomed to 2023–present."

df2_long |>
  filter(date >= as.Date("2023-01-01")) |>
  ggplot(aes(x = date, y = rate, fill = group)) +
  geom_col(alpha = 0.85, width = 22, position = "stack") +
  geom_line(
    data = df2 |> filter(date >= as.Date("2023-01-01")),
    aes(x = date, y = urate, fill = NULL),
    color = "black", linewidth = 0.5, alpha = 0.5
  ) +
  geom_point(
    data = df2 |> filter(date >= as.Date("2023-01-01")),
    aes(x = date, y = urate, fill = NULL),
    color = "black", size = 1
  ) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"),
                     expand = expansion(mult = c(0, 0.02))) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "Unemployment Rate: Separations vs. Entry — Recent Detail",
    subtitle = "Two-group decomposition, 2023–present",
    y = "Unemployment Rate (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )
```

:::

## 2. Change in Unemployment Rate by Cause

```{r urate-chg-compute}
#| label: urate-chg-compute

# Compute YoY change in each cause's rate contribution (six-way)
df6_chg <- df6 |>
  arrange(date) |>
  mutate(
    across(all_of(paste0(cause_order, "_rate")),
           \(x) x - lag(x, 12),
           .names = "{.col}_chg"),
    urate_chg = urate - lag(urate, 12)
  ) |>
  filter(!is.na(urate_chg))

df6_chg_long <- df6_chg |>
  select(date, ends_with("_rate_chg")) |>
  pivot_longer(-date, names_to = "cause", values_to = "chg") |>
  mutate(
    cause = str_remove(cause, "_rate_chg"),
    cause = factor(cause, levels = cause_order)
  )

# Compute YoY change (two-group)
df2_chg <- df2 |>
  arrange(date) |>
  mutate(
    `Job Separations_rate_chg` = `Job Separations_rate` - lag(`Job Separations_rate`, 12),
    `Labor Force Entry_rate_chg` = `Labor Force Entry_rate` - lag(`Labor Force Entry_rate`, 12),
    urate_chg = urate - lag(urate, 12)
  ) |>
  filter(!is.na(urate_chg))

df2_chg_long <- df2_chg |>
  select(date, ends_with("_rate_chg")) |>
  pivot_longer(-date, names_to = "group", values_to = "chg") |>
  mutate(
    group = str_remove(group, "_rate_chg"),
    group = factor(group, levels = group_order)
  )
```

::: {.panel-tabset}

### Six-Way

```{r fig-urate-cause-chg-6}
#| label: fig-urate-cause-chg-6
#| fig-cap: "Year-over-year change in unemployment rate decomposed by cause (six-way)."

p_cause_chg <- df6_chg_long |>
  ggplot(aes(x = date, y = chg, fill = cause)) +
  geom_col(alpha = 0.75, width = 25) +
  geom_line(
    data = df6_chg, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", linewidth = 0.6
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "YoY Change in Unemployment Rate by Cause",
    subtitle = "Additive decomposition: cause contributions sum to total YoY change in unemployment rate (black line)",
    y = "Change (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )

add_recessions(p_cause_chg, alpha = 0.05)
```

### Two-Group

```{r fig-urate-cause-chg-2}
#| label: fig-urate-cause-chg-2
#| fig-cap: "Year-over-year change in unemployment rate: job separations vs. labor force entry."

df2_chg_long |>
  ggplot(aes(x = date, y = chg, fill = group)) +
  geom_col(alpha = 0.75, width = 25) +
  geom_line(
    data = df2_chg, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", linewidth = 0.6
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "YoY Change in Unemployment Rate: Separations vs. Entry",
    subtitle = "Additive decomposition: two-group contributions sum to total YoY change (black line)",
    y = "Change (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

:::

The change decomposition reveals which causes are driving movements in the unemployment rate. Job separations drive virtually all recessionary surges and recoveries. Labor force entry contributes a smaller, more stable component to rate changes.

## 3. Recent Detail: Change in Unemployment Rate by Cause (2023–Present)

```{r urate-chg-recent-compute}
#| label: urate-chg-recent-compute

df6_chg_recent <- df6_chg |> filter(date >= as.Date("2023-01-01"))
df6_chg_recent_long <- df6_chg_long |> filter(date >= as.Date("2023-01-01"))
df2_chg_recent <- df2_chg |> filter(date >= as.Date("2023-01-01"))
df2_chg_recent_long <- df2_chg_long |> filter(date >= as.Date("2023-01-01"))
```

::: {.panel-tabset}

### Six-Way

```{r fig-urate-cause-chg-recent-6}
#| label: fig-urate-cause-chg-recent-6
#| fig-cap: "Year-over-year change in unemployment rate by cause, zoomed to the post-pandemic period (six-way)."

df6_chg_recent_long |>
  ggplot(aes(x = date, y = chg, fill = cause)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df6_chg_recent, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df6_chg_recent, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", size = 1
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "YoY Change in Unemployment Rate by Cause — Recent Detail",
    subtitle = "Post-pandemic period: which causes are driving rate changes?",
    y = "Change (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

### Two-Group

```{r fig-urate-cause-chg-recent-2}
#| label: fig-urate-cause-chg-recent-2
#| fig-cap: "Year-over-year change in unemployment rate: job separations vs. labor force entry, post-pandemic."

df2_chg_recent_long |>
  ggplot(aes(x = date, y = chg, fill = group)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df2_chg_recent, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df2_chg_recent, aes(x = date, y = urate_chg, fill = NULL),
    color = "black", size = 1
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "YoY Change in Unemployment Rate — Recent Detail",
    subtitle = "Post-pandemic: separations vs. entry contributions to rate change",
    y = "Change (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

:::

## 4. Unemployment Composition: Stacked Shares (1994–Present)

```{r stacked-shares-compute}
#| label: stacked-shares-compute

# Compute six-way shares relative to component sum so they stack to exactly 100%
df6_shares <- df6 |>
  mutate(
    component_sum = `Temporary Layoff` + `Permanent Job Losers` + `Completed Temp Job` +
                    `Job Leavers` + Reentrants + `New Entrants`,
    across(all_of(cause_order),
           \(x) x / component_sum * 100,
           .names = "{.col}_share")
  ) |>
  select(date, ends_with("_share")) |>
  pivot_longer(-date, names_to = "cause", values_to = "share") |>
  mutate(
    cause = str_remove(cause, "_share"),
    cause = factor(cause, levels = cause_order)
  )

# Compute two-group shares
df2_shares <- df2 |>
  mutate(
    component_sum = `Job Separations` + `Labor Force Entry`,
    `Job Separations_share` = `Job Separations` / component_sum * 100,
    `Labor Force Entry_share` = `Labor Force Entry` / component_sum * 100
  ) |>
  select(date, `Job Separations_share`, `Labor Force Entry_share`) |>
  pivot_longer(-date, names_to = "group", values_to = "share") |>
  mutate(
    group = str_remove(group, "_share"),
    group = factor(group, levels = group_order)
  )
```

::: {.panel-tabset}

### Six-Way

```{r fig-stacked-6}
#| label: fig-stacked-6
#| fig-cap: "Share of total unemployment by reason (six-way). Shaded areas indicate NBER recessions."

p_stacked <- df6_shares |>
  ggplot(aes(x = date, y = share, fill = cause)) +
  geom_col(alpha = 0.85, width = 28, position = "stack") +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 100),
                     expand = expansion(mult = c(0, 0))) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Composition of Unemployment by Reason",
    subtitle = "Share of total unemployed, seasonally adjusted (six-way decomposition)",
    y = "Share of Total Unemployed (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )

add_recessions(p_stacked)
```

### Two-Group

```{r fig-stacked-2}
#| label: fig-stacked-2
#| fig-cap: "Share of total unemployment: job separations vs. labor force entry."

p_stacked2 <- df2_shares |>
  ggplot(aes(x = date, y = share, fill = group)) +
  geom_col(alpha = 0.85, width = 28, position = "stack") +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(x, "%"), limits = c(0, 100),
                     expand = expansion(mult = c(0, 0))) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "Composition of Unemployment: Separations vs. Entry",
    subtitle = "Share of total unemployed, seasonally adjusted",
    y = "Share of Total Unemployed (%)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11 · Seasonally Adjusted"
  )

add_recessions(p_stacked2)
```

:::

Temporary and permanent layoffs together dominate during and after recessions, with temporary layoffs spiking sharply in 2020 as firms furloughed workers. The two-group view shows job separations consistently account for 55–70% of unemployment, rising sharply during recessions.

## 5. New Entrants Share

::: {.panel-tabset}

### Share of Total Unemployed

```{r fig-ne-share-unemp}
#| label: fig-ne-share-unemp
#| fig-cap: "New entrant share of total unemployment with 12-month moving average."
#| fig-height: 4

ne_share <- df6 |>
  transmute(date, share = `New Entrants` / Total * 100) |>
  mutate(ma12 = zoo::rollmean(share, k = 12, fill = NA, align = "right"))

p_ne <- ne_share |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = share), color = cause_cols["New Entrants"], alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = ma12), color = cause_cols["New Entrants"], linewidth = 1, alpha = 0.75) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "New Entrants as a Share of Total Unemployed",
    subtitle = "Monthly (light) and 12-month moving average (bold)",
    y = "New Entrants Share (%)", x = NULL,
    caption = "Source: BLS via FRED · LNS13023569 / UNEMPLOY"
  )

add_recessions(p_ne)
```

### Share of Labor Force

```{r fig-ne-share-clf}
#| label: fig-ne-share-clf
#| fig-cap: "New entrants as a share of the civilian labor force with 12-month moving average."
#| fig-height: 4

ne_clf <- df6 |>
  transmute(date, share = `New Entrants` / CLF * 100) |>
  mutate(ma12 = zoo::rollmean(share, k = 12, fill = NA, align = "right"))

p_ne_clf <- ne_clf |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = share), color = cause_cols["New Entrants"], alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = ma12), color = cause_cols["New Entrants"], linewidth = 1, alpha = 0.75) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "New Entrants as a Share of the Civilian Labor Force",
    subtitle = "Monthly (light) and 12-month moving average (bold)",
    y = "New Entrants / CLF (%)", x = NULL,
    caption = "Source: BLS via FRED · LNS13023569 / CLF16OV"
  )

add_recessions(p_ne_clf)
```

:::

The share-of-unemployed view is counter-cyclical: it rises in expansions (as first-timers enter a strong market) and dips in recessions (as job losers surge and dominate the denominator). The share-of-labor-force view removes this denominator effect and shows the true incidence of new-entrant unemployment, which is more stable over time.

## 6. YoY Contributions to Unemployment Growth (1995–Present)

```{r yoy-contrib}
#| label: yoy-compute

# ── Contribution decomposition functions ─────────────────────────────────
calc_contributions6 <- function(data, lag = 12, start_date = NULL) {
  data |>
    arrange(date) |>
    mutate(
      across(all_of(cause_order),
             \(x) (x - lag(x, lag)) / lag(Total, lag) * 100,
             .names = "{.col}_contrib"),
      total_chg = (Total - lag(Total, lag)) / lag(Total, lag) * 100
    ) |>
    filter(!is.na(total_chg)) |>
    (\(d) if (!is.null(start_date)) filter(d, date >= start_date) else d)()
}

calc_contributions2 <- function(data, lag = 12, start_date = NULL) {
  data |>
    arrange(date) |>
    mutate(
      across(all_of(group_order),
             \(x) (x - lag(x, lag)) / lag(Total, lag) * 100,
             .names = "{.col}_contrib"),
      total_chg = (Total - lag(Total, lag)) / lag(Total, lag) * 100
    ) |>
    filter(!is.na(total_chg)) |>
    (\(d) if (!is.null(start_date)) filter(d, date >= start_date) else d)()
}

# Six-way YoY
df_yoy <- calc_contributions6(df6, lag = 12)

df_yoy_long <- df_yoy |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(
    cols = ends_with("_contrib"),
    names_to = "cause",
    values_to = "contrib"
  ) |>
  mutate(
    cause = str_remove(cause, "_contrib"),
    cause = factor(cause, levels = cause_order)
  )

# Two-group YoY
df_yoy2 <- calc_contributions2(df2, lag = 12)

df_yoy2_long <- df_yoy2 |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(
    cols = ends_with("_contrib"),
    names_to = "group",
    values_to = "contrib"
  ) |>
  mutate(
    group = str_remove(group, "_contrib"),
    group = factor(group, levels = group_order)
  )
```

::: {.panel-tabset}

### Six-Way

```{r fig-yoy-full-6}
#| label: fig-yoy-full-6
#| fig-cap: "Year-over-year contributions to total unemployment change by reason (six-way)."

p_yoy <- df_yoy_long |>
  ggplot(aes(x = date, y = contrib, fill = cause)) +
  geom_col(alpha = 0.75, width = 25) +
  geom_line(
    data = df_yoy, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.6
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.0f", x), "pp")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "YoY Contributions to Unemployment Growth",
    subtitle = "Additive decomposition: six cause contributions sum to total YoY % change (black line)",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )

add_recessions(p_yoy, alpha = 0.05)
```

### Two-Group

```{r fig-yoy-full-2}
#| label: fig-yoy-full-2
#| fig-cap: "Year-over-year contributions to unemployment change: separations vs. entry."

df_yoy2_long |>
  ggplot(aes(x = date, y = contrib, fill = group)) +
  geom_col(alpha = 0.75, width = 25) +
  geom_line(
    data = df_yoy2, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.6
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.0f", x), "pp")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(
    title = "YoY Contributions to Unemployment Growth: Separations vs. Entry",
    subtitle = "Two-group contributions sum to total YoY % change (black line)",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )

```

:::

Each bar decomposes the YoY % change in total unemployment into additive contributions. Job separations drive virtually all cyclical swings. Labor force entry contributes modestly and more steadily.

## 7. YoY Contributions — Recent Detail (2023–Present)

```{r yoy-recent-compute}
#| label: yoy-recent-compute

df_yoy_recent <- calc_contributions6(df6, lag = 12, start_date = as.Date("2023-01-01"))
df_yoy_recent_long <- df_yoy_recent |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(ends_with("_contrib"), names_to = "cause", values_to = "contrib") |>
  mutate(cause = str_remove(cause, "_contrib"), cause = factor(cause, levels = cause_order))

df_yoy_recent2 <- calc_contributions2(df2, lag = 12, start_date = as.Date("2023-01-01"))
df_yoy_recent2_long <- df_yoy_recent2 |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(ends_with("_contrib"), names_to = "group", values_to = "contrib") |>
  mutate(group = str_remove(group, "_contrib"), group = factor(group, levels = group_order))
```

::: {.panel-tabset}

### Six-Way

```{r fig-yoy-recent-6}
#| label: fig-yoy-recent-6
#| fig-cap: "Year-over-year contributions zoomed to the post-pandemic period (six-way)."

df_yoy_recent_long |>
  ggplot(aes(x = date, y = contrib, fill = cause)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df_yoy_recent, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df_yoy_recent, aes(x = date, y = total_chg, fill = NULL),
    color = "black", size = 1
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.0f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "YoY Contributions — Recent Detail",
    subtitle = "Post-pandemic period: six-way decomposition of unemployment growth",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

### Two-Group

```{r fig-yoy-recent-2}
#| label: fig-yoy-recent-2
#| fig-cap: "Year-over-year contributions zoomed to post-pandemic: separations vs. entry."

df_yoy_recent2_long |>
  ggplot(aes(x = date, y = contrib, fill = group)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df_yoy_recent2, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df_yoy_recent2, aes(x = date, y = total_chg, fill = NULL),
    color = "black", size = 1
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.0f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "YoY Contributions — Recent Detail: Separations vs. Entry",
    subtitle = "Post-pandemic period: two-group decomposition",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

:::

Through 2022, total unemployment was still falling sharply year-over-year as pandemic-era job losses unwound. The six-way breakdown reveals which sub-categories of job loss are driving the recent inflection. The two-group view makes clear whether the driver is separations or new entry into the labor force.

## 8. MoM Contributions to Unemployment Growth (2023–Present)

```{r mom-compute}
#| label: mom-compute

df_mom <- calc_contributions6(df6, lag = 1, start_date = as.Date("2023-01-01"))
df_mom_long <- df_mom |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(ends_with("_contrib"), names_to = "cause", values_to = "contrib") |>
  mutate(cause = str_remove(cause, "_contrib"), cause = factor(cause, levels = cause_order))

df_mom2 <- calc_contributions2(df2, lag = 1, start_date = as.Date("2023-01-01"))
df_mom2_long <- df_mom2 |>
  select(date, ends_with("_contrib"), total_chg) |>
  pivot_longer(ends_with("_contrib"), names_to = "group", values_to = "contrib") |>
  mutate(group = str_remove(group, "_contrib"), group = factor(group, levels = group_order))
```

::: {.panel-tabset}

### Six-Way

```{r fig-mom-recent-6}
#| label: fig-mom-recent-6
#| fig-cap: "Month-over-month contributions to unemployment change, 2023 onward (six-way)."
#| fig-height: 4.5

df_mom_long |>
  ggplot(aes(x = date, y = contrib, fill = cause)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df_mom, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df_mom, aes(x = date, y = total_chg, fill = NULL),
    color = "black", size = 1.5
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = cause_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "MoM Contributions to Unemployment Growth",
    subtitle = "Monthly flow dynamics since January 2023 (six-way decomposition)",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

### Two-Group

```{r fig-mom-recent-2}
#| label: fig-mom-recent-2
#| fig-cap: "Month-over-month contributions: separations vs. entry, 2023 onward."
#| fig-height: 4.5

df_mom2_long |>
  ggplot(aes(x = date, y = contrib, fill = group)) +
  geom_col(alpha = 0.75, width = 22) +
  geom_line(
    data = df_mom2, aes(x = date, y = total_chg, fill = NULL),
    color = "black", linewidth = 0.7
  ) +
  geom_point(
    data = df_mom2, aes(x = date, y = total_chg, fill = NULL),
    color = "black", size = 1.5
  ) +
  geom_hline(yintercept = 0, color = "grey60", linewidth = 0.3) +
  scale_fill_manual(values = group_cols) +
  scale_y_continuous(labels = \(x) paste0(sprintf("%+.1f", x), "pp")) +
  scale_x_date(date_breaks = "3 months", date_labels = "%b\n'%y") +
  labs(
    title = "MoM Contributions: Separations vs. Entry",
    subtitle = "Monthly flow dynamics since January 2023 (two-group)",
    y = "Contribution (pp)", x = NULL,
    caption = "Source: BLS via FRED · CPS Table A-11"
  )
```

:::

Month-to-month contributions are noisy but the six-way breakdown reveals whether monthly swings are driven by temporary layoffs (often reversible) or permanent job losers (more persistent). The two-group view cleanly separates the separation and entry channels.

## 9. Unemployment by Reason: Definitions & Examples

The [BLS](https://www.bls.gov/cps/definitions.htm) classifies every unemployed person into exactly one of six mutually exclusive categories. These six groups sum to total unemployment.

```{r definitions-table}
#| label: tbl-definitions
#| tbl-cap: "BLS unemployment categories: definitions, distinctions, and examples."

tibble(
  Category = c("Temporary Layoff", "Permanent Job Losers", "Completed Temp Job",
                "Job Leavers", "Reentrants", "New Entrants"),
  `BLS Definition` = c(
    "People who have been given a date to return to work or who expect to return within 6 months.",
    "People who have been permanently separated from their previous job — the position was eliminated, they were fired, or the employer closed.",
    "People whose temporary job or assignment has ended and who are now actively seeking new work.",
    "People who voluntarily quit their previous job and are now actively searching for new employment.",
    "People who previously held a job, left the labor force for a period (stopped working and stopped looking), and have now returned to actively searching for work.",
    "People who have never held a job and are actively looking for their first one. No prior work experience of any kind."
  ),
  `Key Distinction` = c(
    "Expects to be recalled. Highly volatile — surges in sudden shocks (e.g., COVID furloughs) and reverses quickly.",
    "No expectation of recall. The most 'recessionary' cause — rises with structural deterioration.",
    "Job ended by design, not involuntary separation. Smaller and less cyclical than permanent losers.",
    "Had a job but chose to leave. A confidence indicator — rises when the market is strong.",
    "Have prior work experience but were out of the labor force entirely (not employed, not searching).",
    "No work history at all — this distinguishes them from reentrants, who have worked before."
  ),
  `FRED Series` = c(
    "[LNS13023653](https://fred.stlouisfed.org/series/LNS13023653)",
    "[LNS13026638](https://fred.stlouisfed.org/series/LNS13026638)",
    "[LNS13026637](https://fred.stlouisfed.org/series/LNS13026637)",
    "[LNS13023705](https://fred.stlouisfed.org/series/LNS13023705)",
    "[LNS13023557](https://fred.stlouisfed.org/series/LNS13023557)",
    "[LNS13023569](https://fred.stlouisfed.org/series/LNS13023569)"
  )
) |>
  gt() |>
  fmt_markdown(columns = `FRED Series`) |>
  tab_options(
    table.font.size = px(12),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  cols_width(
    Category ~ px(120),
    `BLS Definition` ~ px(220),
    `Key Distinction` ~ px(220),
    `FRED Series` ~ px(120)
  ) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(columns = Category)
  ) |>
  tab_style(
    style = cell_fill(color = "#eef2f7"),
    locations = cells_body(rows = c(1, 3, 5))
  ) |>
  tab_style(
    style = cell_fill(color = "#fef9f0"),
    locations = cells_body(rows = c(2, 4, 6))
  )
```

A common source of confusion is the distinction between **reentrants** and **new entrants**. Both are entering the labor force, but the dividing line is simple: has the person ever held a job? If yes → reentrant. If no → new entrant. Among job losers, the key distinction is between **temporary layoffs** (expects recall) and **permanent job losers** (no recall expected) — the latter is the most concerning for the business cycle. **Completed temporary jobs** are a smaller, less cyclical component where the job ended by design. The **job leaver** share is sometimes called a "confidence ratio" because it rises when workers feel secure enough to quit without a job in hand.

## 10. Data Series Reference

```{r series-table}
#| label: tbl-series
#| tbl-cap: "FRED series used in this report."

# Pull latest values for the table
latest4 <- df_wide |> slice_max(date, n = 1)
latest6 <- df6 |> slice_max(date, n = 1)

tibble(
  `Series ID` = c(
    "[LNS13023653](https://fred.stlouisfed.org/series/LNS13023653)",
    "[LNS13026638](https://fred.stlouisfed.org/series/LNS13026638)",
    "[LNS13026637](https://fred.stlouisfed.org/series/LNS13026637)",
    "[LNS13023705](https://fred.stlouisfed.org/series/LNS13023705)",
    "[LNS13023557](https://fred.stlouisfed.org/series/LNS13023557)",
    "[LNS13023569](https://fred.stlouisfed.org/series/LNS13023569)",
    "[LNS13023621](https://fred.stlouisfed.org/series/LNS13023621)",
    "[UNEMPLOY](https://fred.stlouisfed.org/series/UNEMPLOY)",
    "[CLF16OV](https://fred.stlouisfed.org/series/CLF16OV)"
  ),
  Name = c(
    "Unemployment Level — On Temporary Layoff",
    "Unemployment Level — Permanent Job Losers",
    "Unemployment Level — Completed Temporary Job",
    "Unemployment Level — Job Leavers",
    "Unemployment Level — Reentrants",
    "Unemployment Level — New Entrants",
    "Unemployment Level — Job Losers (Aggregate)",
    "Unemployment Level — Total",
    "Civilian Labor Force Level"
  ),
  Units = c(rep("Thousands", 8), "Thousands"),
  SA = rep("Yes", 9),
  Frequency = rep("Monthly", 9),
  `BLS Table` = c(
    rep("[Table A-11](https://www.bls.gov/news.release/empsit.t11.htm)", 8),
    "[Table A-1](https://www.bls.gov/news.release/empsit.t01.htm)"
  )
) |>
  gt() |>
  fmt_markdown(columns = c(`Series ID`, `BLS Table`)) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  ) |>
  tab_source_note(
    source_note = md("All series from the [Current Population Survey (CPS)](https://www.bls.gov/cps/). Release schedule at [bls.gov/schedule](https://www.bls.gov/schedule/news_release/empsit.htm).")
  )
```

```{r related-series-table}
#| label: tbl-related
#| tbl-cap: "Related FRED series (not directly used but referenced)."

tibble(
  `Series ID` = c(
    "[LNS13023570](https://fred.stlouisfed.org/series/LNS13023570)",
    "[LNS13023622](https://fred.stlouisfed.org/series/LNS13023622)",
    "[LNS13023706](https://fred.stlouisfed.org/series/LNS13023706)",
    "[LNS13023558](https://fred.stlouisfed.org/series/LNS13023558)",
    "[LNS14023569](https://fred.stlouisfed.org/series/LNS14023569)"
  ),
  Name = c(
    "New Entrants as % of Total Unemployed",
    "Job Losers as % of Total Unemployed",
    "Job Leavers as % of Total Unemployed",
    "Reentrants as % of Total Unemployed",
    "Unemployment Rate — New Entrants"
  ),
  Description = c(
    "Pre-computed share: NE / Total × 100",
    "Pre-computed share: JL / Total × 100",
    "Pre-computed share: JV / Total × 100",
    "Pre-computed share: RE / Total × 100",
    "NE unemployed as share of NE labor force"
  )
) |>
  gt() |>
  fmt_markdown(columns = `Series ID`) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#5d6d7e"
  ) |>
  tab_style(
    style = cell_text(color = "white"),
    locations = cells_column_labels()
  )
```

All series available at [fred.stlouisfed.org](https://fred.stlouisfed.org). BLS source: [Table A-11](https://www.bls.gov/news.release/empsit.t11.htm), [Employment Situation](https://www.bls.gov/news.release/empsit.nr0.htm) report. BLS concept definitions: [CPS Definitions](https://www.bls.gov/cps/definitions.htm).

## 11. Methodology

The six BLS reason-for-unemployment categories (temporary layoff, permanent job losers, completed temporary job, job leavers, reentrants, and new entrants) are exhaustive and mutually exclusive; they sum to total unemployment.

**Unemployment rate by cause**: Each component's contribution to the unemployment rate is simply:

$$
r_{X,t} = \frac{X_t}{L_t} \times 100
$$

where $X_t$ is the unemployment level for cause $X$ and $L_t$ is the civilian labor force. These sum to the total unemployment rate $u_t = U_t / L_t \times 100$.

**Change decomposition**: The YoY change in each cause's rate contribution is:

$$
\Delta r_{X,t}^{YoY} = r_{X,t} - r_{X,t-12} = \frac{X_t}{L_t} \times 100 - \frac{X_{t-12}}{L_{t-12}} \times 100
$$

These changes sum to the total change in the unemployment rate: $\sum_X \Delta r_{X,t}^{YoY} = u_t - u_{t-12}$.

**YoY contribution (four-way)**: For each component $X$, the contribution to the YoY percentage change in total unemployment is:

$$
\text{Contrib}_{X,t}^{YoY} = \frac{X_t - X_{t-12}}{U_{t-12}} \times 100
$$

These contributions sum exactly to the total YoY percentage change:

$$
\sum_X \text{Contrib}_{X,t}^{YoY} = \frac{U_t - U_{t-12}}{U_{t-12}} \times 100
$$

**MoM contribution**: Same formula with a 1-month lag:

$$
\text{Contrib}_{X,t}^{MoM} = \frac{X_t - X_{t-1}}{U_{t-1}} \times 100
$$

**Seasonal adjustment and the component-sum discrepancy**: Although the six reason-for-unemployment categories are conceptually exhaustive and mutually exclusive, the BLS seasonally adjusts each component series *independently* of the headline UNEMPLOY total. Because seasonal adjustment is a nonlinear filtering process, the sum of the six independently adjusted components does not exactly equal the independently adjusted total.

```{r sa-discrepancy}
#| label: sa-discrepancy

sa_disc <- df6 |>
  mutate(
    component_sum = `Temporary Layoff` + `Permanent Job Losers` + `Completed Temp Job` +
                    `Job Leavers` + Reentrants + `New Entrants`,
    discrepancy = component_sum - Total
  )

mean_disc <- round(mean(abs(sa_disc$discrepancy), na.rm = TRUE))
max_disc  <- round(max(abs(sa_disc$discrepancy), na.rm = TRUE))
mean_total <- round(mean(sa_disc$Total, na.rm = TRUE))
```

In our data the absolute discrepancy averages roughly `r comma(mean_disc)` thousand and can reach as high as `r comma(max_disc)` thousand — small relative to a mean total unemployment level of `r comma(mean_total)` thousand. To ensure the stacked-share chart (Section 4) sums to exactly 100%, we compute each component's share as a fraction of the *component sum* rather than UNEMPLOY. The unemployment-rate-by-cause charts (Sections 1–1b) and the contribution decompositions (Sections 2–3, 6–8) use the civilian labor force (CLF) or UNEMPLOY as the denominator to match standard BLS definitions; these may exhibit minor rounding discrepancies relative to the published headline rate.

All series begin in January 1994 to avoid comparability issues with the pre-redesign CPS.

---

*Source: [U.S. Bureau of Labor Statistics](https://www.bls.gov/) via [FRED](https://fred.stlouisfed.org/). [NBER recession dates](https://www.nber.org/research/data/us-business-cycle-expansions-and-contractions). For informational purposes only.*
