---
title: "CFI Blog"
author: "Kristin Butcher, Scott Brave, Aryan Safi"
format:
  html:
    code-fold: true
  pdf:
    toc: false
    number-sections: false
    keep-tex: false
    echo: false
editor: visual
---

```{r setup, include=F}

library(tidyverse)
library(arrow)
library(fredr)
library(here)
library(ggsci)      # for scale_*_jco()
library(scales)
library(gt)
library(lubridate)
library(patchwork)
library(sandwich)
library(lmtest)
rm(list=ls())

fredr_set_key("ccd350785a210a7653a20d02ce92d83a")

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )

theme_set(theme_report)

```

## Introduction

The Department of Labor releases initial and continuing unemployment insurance (UI) claims data every Thursday at 8:30 a.m. Eastern, covering the week that ended the previous Saturday for initial claims and the week before that for continuing claims. Three features make claims data attractive as a real-time signal relative to the monthly Bureau of Labor Statistics (BLS) employment situation release:

1.  **Frequency**. Released weekly, which provides more detailed snapshots of the labor market in the periods between the monthly BLS report.

2.  **Timeliness**. Published with a five-day lag, versus the three-week lag of the BLS payroll report, enabling closer-to-real-time updates on the health of the labor market.

3.  **Administrative grounding**. Claims are counted when people file, not estimated from a survey.

UI programs are designed to provide unemployment benefits to eligible workers who become unemployed through no fault of their own and meet other eligibility requirements. For this reason, the majority of initial claims for unemployment insurance benefits are filed because of a separation from a job the filer did not initiate. Causes of such separations are layoffs, temporary furloughs, completed temporary contracts, among others (see [How Do I File for Unemployment Insurance](https://www.dol.gov/general/topic/unemployment-insurance)). Because of UI program design, initial claims are a **noisy, lower-bound read on separations**. Specifically, initial claims will miss **separations that do not generate claims** such as short-tenure workers, part-timers, gig/1099 workers and people who do not satisfy work-history or monetary-based eligibility requirements.

Continuing claims are the number of people currently drawing unemployment insurance benefits and can change for two reasons. The first is that the number of people who *initiate* claims changes (i.e., changes to the *inflows* to the insured-unemployment pool). The second is that people stop claiming benefits (i.e., changes to *outflows* from the insured-unemployment pool). Broadly, there are three reasons why people stop claiming: they find a job; they exhaust the UI benefits available to them; they fail to maintain eligibility.

Because of these desirable features, UI claims became the default "nowcast" input for labor-market watchers long before high-frequency alternatives existed. But the labor market signal UI claims provide depends on *who the unemployed are*. When the unemployed pool is composed of groups UI programs were not designed to cover, the same claims release can imply something very different about the state of the labor market. This matters today in a specific way: the share of the unemployed who are New Entrants, people who have never held a job, has sat above its long-run average since 2022. That can mean the labor market is healthy in the familiar cyclical way, with separations low and the unemployed pool tilted toward people UI was never designed to cover. Or it can mean hiring has weakened in ways that disproportionately affect those same groups, so claims stay quiet not because the labor market is strong but because the weakening is concentrated where claims cannot see it. In this post, we use the Current Population Survey's (CPS) unemployment-by-cause data and the flow-consistent unemployment framework behind the Chicago Fed Labor Market Indicators to tell these two stories apart.

## Why People are Unemployed

The CPS provides six mutually exclusive and exhaustive reason-for-unemployment categories: Temporary Layoff, Permanent Job Losers, Completed Temp Job, Job Leavers, Reentrants, and New Entrants. The definitions of those categories are provided in Table 1.

```{r table-1, fig.cap = 'Table 1: CPS WHYUNEMP Categories' }
cause_order = c("Temporary Layoff", "Permanent Job Losers", "Completed Temp Job",
                 "Job Leavers", "Reentrants", "New Entrants")
tibble(
  Category = cause_order,
  WHYUNEMP = 1:6,
  `BLS Definition` = c(
    "Given a return date or expects recall within 6 months",
    "Permanently separated; position eliminated, fired, or employer closed",
    "Temporary job or assignment has ended",
    "Voluntarily quit and now actively searching",
    "Previously held a job, left the labor force, now returned to searching",
    "Never held a job; actively looking for first one"
  )
) |>
  gt() |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) 

```

These categories map onto UI eligibility in very different ways. Temporary Layoffs and Permanent Job Losers are the program's target group: people who have separated from their job through no fault of their own, and who typically have the work history and earnings needed to qualify. The remaining four groups are variously outside the program's design. Completed Temp Jobs can go either way depending on state rules. Job Leavers usually don't qualify except in narrow circumstances. Reentrants often fail work-history tests because their last job was too long ago. New Entrants have no prior job at all, so they cannot meet the eligibility criteria by construction.

To see which reason-for-unemployment categories are most plausibly captured by UI versus most plausibly missed by it, we construct a profile of each cause. For every cause category in Table 1, we identify the dominant demographic group and compute a *representation ratio* on a set of standard dimensions: age, education, sex, race, nativity, citizenship, and prior class of worker. The representation ratio is defined as:

$$
\text{representation ratio} = \frac{\text{group's share of this cause's unemployed pool}}{\text{group's share of the overall labor force}}
$$

A value of 1 means the group's presence among that cause's unemployed is exactly proportional to its presence in the labor force. A value above 1 means the group is overrepresented; a value below 1 means underrepresented. Values are computed on the pooled CPS sample (1994–present) so that each ratio reflects a long-run average rather than a single-year snapshot. We present each cause's ratios on a radar ("spiderweb") plot. Each spoke is one demographic dimension, labeled at the end with the dominant group on that dimension (e.g., "Age 16–24" on the age spoke, "HS Diploma" on the education spoke). The radius at that spoke is the group's representation ratio. A polygon hugging the 1 ring says the cause's unemployed look roughly like the country's workers along each margin. A polygon stretching far outside the 1 ring says that a cause's unemployed are overrepresented relative to the country's workers. A polygon deep within the 1 ring says that a cause's unemployed are underrepresented relative to the country's workers. 

```{r radar-representation-ratio, fig.width=15, fig.height=16}
#| fig-cap: "Dominant-group representation ratios by reason-for-unemployment category, pooled CPS 1994–present. Each spoke shows the group with the largest share on that dimension; the radius is that group's share of the cause-specific unemployed pool divided by its share of the labor force. Solid ring = 1 (proportional representation)."

dominant_profile_ratios <- arrow::read_parquet(
  here("data", "final", "unemployment_dominant_profile_ratios.parquet")
) |>
  as_tibble() |>
  mutate(
    cause = factor(cause, levels = cause_order),
    dimension = factor(dimension, levels = c("Age", "Education", "Sex", "Race/Ethnicity",
                                             "Occupation", "Class", "Foreign Born", "Noncitizen"))
  )

cause_cols <- pal_jco()(6)
names(cause_cols) <- cause_order

pooled_ratios <- dominant_profile_ratios |>
  filter(variant == "pooled_ratio", cause %in% cause_order)

plot_dominant_radar_ratio <- function(ratio_df, cause_name, subtitle, max_ratio = 3.0) {
  plot_df <- ratio_df |>
    filter(cause == cause_name) |>
    arrange(dimension) |>
    mutate(axis_id = row_number())

  n_axes <- nrow(plot_df)
  angles <- pi / 2 - 2 * pi * (seq_len(n_axes) - 1) / n_axes
  radial_breaks <- c(0, 0.5, 1.0, 1.5, 2.0, max_ratio)

  point_df <- plot_df |>
    mutate(
      angle = angles,
      radius = pmin(ratio, max_ratio) / max_ratio,
      x = radius * cos(angle),
      y = radius * sin(angle)
    )

  closed_df <- bind_rows(point_df, slice(point_df, 1))

  spoke_df <- tibble(
    axis_id = seq_len(n_axes),
    angle = angles,
    x = 0, y = 0,
    xend = cos(angle), yend = sin(angle)
  )

  ring_df <- map_dfr(radial_breaks, \(ring) {
    tibble(
      ring = ring,
      ring_radius = ring / max_ratio,
      axis_id = c(seq_len(n_axes), 1L),
      angle = angles[c(seq_len(n_axes), 1L)],
      x = (ring / max_ratio) * cos(angle),
      y = (ring / max_ratio) * sin(angle)
    )
  })

  axis_label_df <- point_df |>
    transmute(
      dimension, angle,
      label_x = 1.17 * cos(angle),
      label_y = 1.17 * sin(angle),
      hjust = case_when(label_x > 0.15 ~ 0, label_x < -0.15 ~ 1, TRUE ~ 0.5),
      vjust = case_when(label_y > 0.15 ~ 0, label_y < -0.15 ~ 1, TRUE ~ 0.5)
    )

  point_label_df <- point_df |>
    transmute(
      axis_id, angle,
      point_label = label,
      label_radius = case_when(
        radius >= 0.72 ~ pmax(radius - 0.14, 0.16),
        TRUE ~ pmin(radius + 0.11, 0.88)
      ),
      tangent_shift = if_else(axis_id %% 2 == 0, 0.045, -0.045),
      label_x = label_radius * cos(angle) + tangent_shift * (-sin(angle)),
      label_y = label_radius * sin(angle) + tangent_shift * cos(angle),
      hjust = case_when(label_x > 0.18 ~ 0, label_x < -0.18 ~ 1, TRUE ~ 0.5),
      vjust = case_when(label_y > 0.18 ~ 0, label_y < -0.18 ~ 1, TRUE ~ 0.5)
    )

  radial_label_df <- tibble(
    ring = radial_breaks,
    x = radial_breaks / max_ratio,
    y = 0.02,
    label = paste0(scales::number(radial_breaks, accuracy = 0.1), "x")
  )

  ggplot() +
    geom_path(
      data = filter(ring_df, abs(ring - 1.0) > 0.01),
      aes(x = x, y = y, group = ring),
      color = "grey72", linewidth = 0.15
    ) +
    geom_path(
      data = filter(ring_df, abs(ring - 1.0) <= 0.01),
      aes(x = x, y = y, group = ring),
      color = "grey40", linewidth = 0.6, linetype = "solid"
    ) +
    geom_segment(
      data = spoke_df,
      aes(x = x, y = y, xend = xend, yend = yend),
      color = "grey76", linewidth = 0.15
    ) +
    geom_polygon(
      data = closed_df,
      aes(x = x, y = y),
      fill = scales::alpha(cause_cols[[cause_name]], 0.45),
      color = cause_cols[[cause_name]],
      linewidth = 0.15, linejoin = "round"
    ) +
    geom_point(
      data = point_df,
      aes(x = x, y = y),
      color = cause_cols[[cause_name]],
      fill = cause_cols[[cause_name]],
      shape = 21, stroke = 0.2, size = 1.4
    ) +
    geom_text(
      data = point_label_df,
      aes(x = label_x, y = label_y, label = point_label, hjust = hjust, vjust = vjust),
      size = 2.7, lineheight = 0.92,
      color = cause_cols[[cause_name]]
    ) +
    geom_text(
      data = axis_label_df,
      aes(x = label_x, y = label_y, label = dimension, hjust = hjust, vjust = vjust),
      size = 3.6, lineheight = 0.9
    ) +
    geom_text(
      data = radial_label_df,
      aes(x = x, y = y, label = label),
      color = "grey65", size = 3.3, hjust = 0.5, vjust = 0
    ) +
    coord_equal(clip = "off") +
    xlim(-1.28, 1.28) + ylim(-1.26, 1.28) +
    labs(title = cause_name, subtitle = subtitle, x = NULL, y = NULL) +
    theme_report +
    theme(
      axis.text = element_blank(),
      axis.ticks = element_blank(),
      axis.line = element_blank(),
      panel.grid = element_blank(),
      plot.margin = margin(20, 35, 20, 35)
    )
}

radar_plots <- map(cause_order, \(cn) {
  plot_dominant_radar_ratio(pooled_ratios, cn, "Pooled — ratio to labor force share")
})

patchwork::wrap_plots(radar_plots, ncol = 2)
```

For five of the six categories, the polygons hug the 1-ring, deviating on at most one spoke. New Entrants are the clear outlier, sitting far outside the 1-ring on age, education, and, by construction, occupation and class, since they have no prior job. Because UI eligibility depends on recent work history and earnings thresholds, these demographic differences map directly onto how much of a cause's unemployed actually shows up in the claims data.

```{r ne-composition-narrative, include=FALSE}
# Provenance: this parquet is produced by build_panels.R section 8c
# (unemployment_new_entrant_composition.parquet). The paragraph below cites it
# inline via nec() so the numbers always match the build.
ne_comp <- arrow::read_parquet(
  here("data", "final", "unemployment_new_entrant_composition.parquet")
)
nec <- function(w, m) {
  sprintf("%.0f%%", 100 * ne_comp$value[ne_comp$window == w & ne_comp$metric == m])
}
```

One might object that New Entrants are just teenagers — and pooled over 1994–present they largely are, with `r nec("pooled", "age_16_19")` aged 16–19 and `r nec("pooled", "educ_below_hs")` below a high-school diploma. But that pooled picture averages over thirty years, and 2025 looks different: `r nec("2025", "age_18plus")` are 18 or older (up from `r nec("pooled", "age_18plus")`), `r nec("2025", "educ_hs_plus")` have at least a high-school diploma (up from `r nec("pooled", "educ_hs_plus")`), and the college-educated share has roughly doubled, to `r nec("2025", "educ_college_plus")`. The base still skews young, but the recent tilt toward older, degree-holding first-job seekers is exactly the group the entry-level hiring concern is about.

## Composition and UI Takeup

The CPS cause categories appear to be useful proxies for UI eligibility in the unemployed pool. To test whether the  distinction shows up in the UI claims data, we estimate a simple national monthly regression of initial and continuing claims intensity on the composition of the unemployed:

$$
\text{Claims}_t = \alpha + \sum_{j \neq \text{Completed Temp Job}} \beta_j C_j + \gamma_m + \varepsilon_t
$$

The outcomes are continuing claims and initial claims, each scaled by covered employment; the regressors are the five cause shares, with Completed Temp Job omitted as the reference category. Month-of-year fixed effects $\gamma_m$ absorb seasonality. The sample is monthly national data from 1994 to the present, excluding the pandemic window (March 2020–December 2021) during which both compositions and claiming broke from their historical relationships.

We use cause *shares* rather than counts so the regression answers a purely compositional question: when the unemployed pool tilts toward one cause and away from another, holding the total number of unemployed fixed, how does claims intensity respond? A regression on counts would also pick up the mechanical fact that both claims and cause counts rise when total unemployment rises, plus three decades of secular labor-force growth. Shares strip both of those out.

We scale by covered employment, not total employment or the labor force, because covered employment is the population from which these claims can originate: workers whose wages are subject to UI taxes. A broader denominator would dilute the numerator with workers who are not UI-eligible by virtue of their employment status (independent contractors, the self-employed, some agricultural and federal workers), confounding claiming behavior with shifts in the mix of covered and non-covered jobs. CC/CE is also the "insured unemployment rate" published weekly by the Department of Labor and the form in which claims enter the CFLMI's partial-least-squares step, so it keeps our units aligned with the downstream analysis.

```{r composition-regression-data, warning=F, message=F}
cause_series <- c(
  "Temporary Layoff"     = "LNS13023653",
  "Permanent Job Losers" = "LNS13026638",
  "Completed Temp Job"   = "LNS13026637",
  "Job Leavers"          = "LNS13023705",
  "Reentrants"           = "LNS13023557",
  "New Entrants"         = "LNS13023569"
)

cause_monthly <- map_dfr(names(cause_series), \(nm) {
  fredr(series_id = cause_series[[nm]], observation_start = as.Date("1994-01-01")) |>
    transmute(date, cause = nm, value)
}) |>
  pivot_wider(names_from = cause, values_from = value)

total_unemp <- fredr(series_id = "UNEMPLOY", observation_start = as.Date("1994-01-01")) |>
  transmute(date, total = value)

claims_weekly <- bind_rows(
  fredr(series_id = "CCSA",   observation_start = as.Date("1994-01-01")) |>
    transmute(date, series = "CCSA",   value),
  fredr(series_id = "ICSA",   observation_start = as.Date("1994-01-01")) |>
    transmute(date, series = "ICSA",   value),
  fredr(series_id = "COVEMP", observation_start = as.Date("1994-01-01")) |>
    transmute(date, series = "COVEMP", value)
) |>
  mutate(month = floor_date(date, "month")) |>
  group_by(month, series) |>
  summarise(value = mean(value, na.rm = TRUE), .groups = "drop") |>
  pivot_wider(names_from = series, values_from = value) |>
  transmute(
    date      = month,
    cc_ce_pct = 100 * CCSA / COVEMP,
    ic_ce_pct = 100 * ICSA / COVEMP
  )

reg_df <- cause_monthly |>
  inner_join(total_unemp, by = "date") |>
  mutate(
    temp_layoff_share    = `Temporary Layoff`     / total,
    perm_job_loser_share = `Permanent Job Losers` / total,
    completed_temp_share = `Completed Temp Job`   / total,
    job_leaver_share     = `Job Leavers`          / total,
    reentrant_share      = `Reentrants`           / total,
    new_entrant_share    = `New Entrants`         / total,
    entrant_share        = (`Reentrants` + `New Entrants`) / total,
    month_of_year        = factor(month.abb[month(date)], levels = month.abb),
    pandemic             = date >= as.Date("2020-03-01") & date <= as.Date("2021-12-01")
  ) |>
  inner_join(claims_weekly, by = "date") |>
  filter(!pandemic) |>
  drop_na(cc_ce_pct, ic_ce_pct, temp_layoff_share, perm_job_loser_share,
          job_leaver_share, reentrant_share, new_entrant_share)

comp_rhs <- "temp_layoff_share + perm_job_loser_share + job_leaver_share + reentrant_share + new_entrant_share + month_of_year"

cc_mod <- lm(as.formula(paste("cc_ce_pct ~", comp_rhs)), data = reg_df)
ic_mod <- lm(as.formula(paste("ic_ce_pct ~", comp_rhs)), data = reg_df)

# Companion specification: collapse Reentrants + New Entrants into a single
# labor-force-entry share. Completed Temp Job remains the omitted reference.
comp_rhs_ent <- "temp_layoff_share + perm_job_loser_share + job_leaver_share + entrant_share + month_of_year"

cc_mod_ent <- lm(as.formula(paste("cc_ce_pct ~", comp_rhs_ent)), data = reg_df)
ic_mod_ent <- lm(as.formula(paste("ic_ce_pct ~", comp_rhs_ent)), data = reg_df)
```

```{r composition-regression-table}
#| tbl-cap: "National monthly regressions of claims intensity on by-cause unemployment shares, 1994–present excluding March 2020–December 2021. Completed Temp Job is the omitted reference category; month-of-year fixed effects absorb seasonality. CC/CE and IC/CE are expressed as percent; coefficients are the percentage-point change in claims intensity per 1 percentage-point shift in share from Completed Temp Job. Standard errors are HAC (Newey-West) with automatic bandwidth selection."

cause_labels <- c(
  temp_layoff_share    = "Temporary layoff share",
  perm_job_loser_share = "Permanent job-loser share",
  job_leaver_share     = "Job leaver share",
  reentrant_share      = "Reentrant share",
  new_entrant_share    = "New entrant share"
)

extract_share_terms <- function(model, outcome_label, labels = cause_labels) {
  ct <- lmtest::coeftest(model, vcov = sandwich::NeweyWest(model, prewhite = FALSE))
  as.data.frame(unclass(ct)) |>
    rownames_to_column("term") |>
    as_tibble() |>
    filter(term %in% names(labels)) |>
    mutate(
      Outcome = outcome_label,
      Term = factor(labels[term], levels = labels),
      `Coef (pp per 1 pp shift)` = Estimate * 0.01,
      `Std. Error` = `Std. Error` * 0.01
    ) |>
    arrange(Term) |>
    select(Outcome, Term, `Coef (pp per 1 pp shift)`, `Std. Error`, `p-value` = `Pr(>|t|)`)
}

bind_rows(
  extract_share_terms(cc_mod, "CC / CE"),
  extract_share_terms(ic_mod, "IC / CE")
) |>
  gt(groupname_col = "Outcome") |>
  fmt_number(columns = c(`Coef (pp per 1 pp shift)`, `Std. Error`, `p-value`), decimals = 3) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) |>
  tab_source_note(sprintf(
    "R-squared: CC/CE = %.3f, IC/CE = %.3f. N = %d months.",
    summary(cc_mod)$r.squared, summary(ic_mod)$r.squared, nrow(reg_df)
  ))
```

```{r composition-narrative, include=FALSE}
# NE coefficient in pp per 1 pp shift (same scaling as the table above)
ne_coef_cc <- as.numeric(coef(cc_mod)[["new_entrant_share"]]) * 0.01

# NE share of the unemployed over the full monthly history (percent)
ne_share_ts <- cause_monthly |>
  inner_join(total_unemp, by = "date") |>
  transmute(date, s = 100 * `New Entrants` / total) |>
  arrange(date)
ne_share_full_mean <- mean(ne_share_ts$s, na.rm = TRUE)
ne_share_last12    <- mean(tail(ne_share_ts$s, 12), na.rm = TRUE)
ne_gap_pp          <- ne_share_last12 - ne_share_full_mean
implied_pp         <- ne_gap_pp * ne_coef_cc

# Latest insured unemployment rate (CC/CE), last finite month
iur_latest <- tail(na.omit(claims_weekly$cc_ce_pct), 1)
```

The CC/CE column tells a clean compositional story. Temporary Layoffs and Permanent Job Losers, the two causes closest to UI's target population, enter positively: months when they make up a larger share of the unemployed see higher continuing claims intensity. New Entrants and Job Leavers enter negatively, consistent with those groups being largely UI-ineligible. The New Entrants coefficient on CC/CE is `r sprintf("%.3f", ne_coef_cc)`, meaning that a 1 pp rise in the New Entrants share is associated with an insured unemployment rate about `r sprintf("%.2f", abs(ne_coef_cc))` pp lower, holding other causes and month fixed. With the NE share running about `r sprintf("%.1f", ne_gap_pp)` pp above its long-run mean over the past twelve months, composition alone would imply today's insured unemployment rate sits about `r sprintf("%.2f", abs(implied_pp))` pp below what the rest of the mix would predict. That is a meaningful fraction of today's \~`r sprintf("%.1f", iur_latest)`% reading, and points in the direction of the UI-ineligible story. The IC/CE coefficients are about seven times smaller than their CC/CE counterparts. Initial claims are a weekly flow of new filers; continuing claims are a stock that accumulates those filers over typical spell duration, so any composition effect on filing rates shows up amplified in continuing claims. A tilt toward eligible groups also tends to pull in filers with longer spells, which compounds the stock effect.

We focus on New Entrants in what follows for three reasons. They are UI-ineligible by construction, which makes the "claims cannot see them" story cleanest. Their share has risen the most in the recent run-up. And the widespread concern that new graduates are having a hard time finding work is squarely about this group. Job Leavers tell a consistent story since voluntary quitters are also largely UI-ineligible, but they are not currently drawing the bulk of the attention. Reentrants load weakly positive, contrary to what their partial UI-ineligibility would predict, so we do not lean on their coefficient.

As a companion to the by-cause specification, we re-estimate with Reentrants and New Entrants collapsed into a single labor-force-entry share, holding the omitted reference (Completed Temp Job), the remaining cause shares, and the month fixed effects unchanged. This is the coarser entry-margin cut: it summarizes both never-worked and returning searchers in one coefficient rather than splitting them.

```{r composition-entrant-narrative, include=FALSE}
# Combined-entrants coefficient on CC/CE, same pp-per-1-pp-shift scaling.
ent_coef_cc <- as.numeric(coef(cc_mod_ent)[["entrant_share"]]) * 0.01
```

```{r composition-regression-table-entrants}
#| tbl-cap: "Companion specification with Reentrants and New Entrants collapsed into a single labor-force-entry share. National monthly regressions of claims intensity on by-cause unemployment shares, 1994–present excluding March 2020–December 2021. Completed Temp Job is the omitted reference category; month-of-year fixed effects absorb seasonality. CC/CE and IC/CE are expressed as percent; coefficients are the percentage-point change in claims intensity per 1 percentage-point shift in share from Completed Temp Job. Standard errors are HAC (Newey-West) with automatic bandwidth selection."

cause_labels_ent <- c(
  temp_layoff_share    = "Temporary layoff share",
  perm_job_loser_share = "Permanent job-loser share",
  job_leaver_share     = "Job leaver share",
  entrant_share        = "Entrants share (New + Reentrant)"
)

bind_rows(
  extract_share_terms(cc_mod_ent, "CC / CE", labels = cause_labels_ent),
  extract_share_terms(ic_mod_ent, "IC / CE", labels = cause_labels_ent)
) |>
  gt(groupname_col = "Outcome") |>
  fmt_number(columns = c(`Coef (pp per 1 pp shift)`, `Std. Error`, `p-value`), decimals = 3) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) |>
  tab_source_note(sprintf(
    "R-squared: CC/CE = %.3f, IC/CE = %.3f. N = %d months.",
    summary(cc_mod_ent)$r.squared, summary(ic_mod_ent)$r.squared, nrow(reg_df)
  ))
```

The combined entry margin enters CC/CE with a coefficient of `r sprintf("%.3f", ent_coef_cc)`. Because Reentrants load weakly positive and New Entrants negatively, the pooled entry coefficient sits between the two and is a blunter read on the eligibility blind spot than the New-Entrants-only coefficient above; the by-cause split is what isolates the group claims cannot see.

The regression table is a useful descriptive benchmark. However, it is important to emphasize that these estimates are not causal. Composition shares are jointly determined with claims by the same underlying labor-market trends. The figure below plots New Entrants as a share of the unemployed (top) and as a share of the civilian labor force (bottom) from 1994 to present, each with a 12-month moving average and NBER recessions shaded.

```{r new-entrants-share, fig.width=10, fig.height=8, warning=F, message=F}
#| fig-cap: "New Entrants as a share of total unemployed (top) and as a share of the civilian labor force (bottom), monthly (light) and 12-month moving average (bold). Shaded bands are NBER recessions. Source: BLS via FRED."

cause_series <- c(
  "Temporary Layoff"     = "LNS13023653",
  "Permanent Job Losers" = "LNS13026638",
  "Completed Temp Job"   = "LNS13026637",
  "Job Leavers"          = "LNS13023705",
  "Reentrants"           = "LNS13023557",
  "New Entrants"         = "LNS13023569"
)

tot_raw <- fredr(series_id = "UNEMPLOY", observation_start = as.Date("1994-01-01")) |>
  transmute(date, total = value)

cause_raw <- bind_rows(lapply(names(cause_series), function(cn) {
  fredr(series_id = cause_series[[cn]], observation_start = as.Date("1994-01-01")) |>
    transmute(date, value, cause = cn)
}))

recessions <- tibble(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-01", "2009-06-01", "2020-04-01"))
)

add_recessions <- function(p, alpha = 0.08) {
  p + annotate("rect",
               xmin = recessions$start, xmax = recessions$end,
               ymin = -Inf, ymax = Inf,
               fill = "grey40", alpha = alpha)
}

shares <- cause_raw |>
  inner_join(tot_raw, by = "date") |>
  transmute(date, cause, share = value / total * 100) |>
  arrange(cause, date) |>
  group_by(cause) |>
  mutate(ma12 = zoo::rollapplyr(share, width = 12, FUN = mean,
                                na.rm = TRUE, fill = NA)) |>
  ungroup() |>
  mutate(cause = factor(cause, levels = cause_order))

make_share_panel <- function(cause_name) {
  d <- shares |> filter(cause == cause_name)
  p <- d |>
    ggplot(aes(x = date)) +
    geom_line(aes(y = share), color = cause_cols[cause_name], alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = ma12),  color = cause_cols[cause_name], linewidth = 1, alpha = 0.85) +
    scale_y_continuous(labels = \(x) paste0(x, "%")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = "Share of Unemployed", y = "Share (%)", x = NULL)
  add_recessions(p)
}

lf_raw <- fredr(series_id = "CLF16OV", observation_start = as.Date("1994-01-01")) |>
  transmute(date, lf = value)

ne_lf <- cause_raw |>
  filter(cause == "New Entrants") |>
  inner_join(lf_raw, by = "date") |>
  arrange(date) |>
  mutate(
    lf_share = 100 * value / lf,
    ma12 = zoo::rollapplyr(lf_share, width = 12, FUN = mean,
                           na.rm = TRUE, fill = NA)
  )

make_lf_panel <- function() {
  p <- ne_lf |>
    ggplot(aes(x = date)) +
    geom_line(aes(y = lf_share), color = cause_cols["New Entrants"], alpha = 0.3, linewidth = 0.4) +
    geom_line(aes(y = ma12),     color = cause_cols["New Entrants"], linewidth = 1, alpha = 0.85) +
    scale_y_continuous(labels = \(x) paste0(x, "%")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = "Share of Labor Force", y = "Share (%)", x = NULL)
  add_recessions(p)
}

make_share_panel("New Entrants") / make_lf_panel() +
  plot_annotation(
    title = "New Entrants: Share of Unemployed vs Share of Labor Force",
    subtitle = "Monthly (light) and 12-month moving average (bold)"
  )
```

```{r ne-share-narrative, include=FALSE}
ne_u <- shares |> filter(cause == "New Entrants") |> arrange(date)
ne_share_u_mean   <- mean(ne_u$share, na.rm = TRUE)
ne_share_u_last12 <- mean(tail(ne_u$share, 12), na.rm = TRUE)
ne_share_u_pre    <- mean(ne_u$share[ne_u$date < as.Date("2020-03-01")], na.rm = TRUE)
ne_lf_last12  <- mean(tail(ne_lf$lf_share, 12), na.rm = TRUE)
ne_lf_median  <- median(ne_lf$lf_share, na.rm = TRUE)
```

Three things stand out. First, the New Entrants share (top panel) moves *opposite to the unemployment rate*: it falls in recessions, because Permanent Job Losers and Temporary Layoffs spike and dilute the entrant share, and drifts back up in expansions. Second, the 12-month average of the share has sat above its long-run mean (\~`r sprintf("%.1f", ne_share_u_mean)`%) since 2022, and averaged \~`r sprintf("%.1f", ne_share_u_last12)`% over the last twelve months, which is notably higher than the pre-pandemic norm (\~`r sprintf("%.1f", ne_share_u_pre)`%). Third, the labor-force share (bottom panel) tells a quieter story: New Entrants have averaged about `r sprintf("%.2f", ne_lf_last12)`% of the civilian labor force over the last twelve months, close to the long-run median (\~`r sprintf("%.2f", ne_lf_median)`%). The share of the unemployed is elevated; the share of the labor force is only modestly so. That the New Entrants share (i.e., UI-ineligible persons) is highest precisely when the unemployment rate is lowest is what makes today's read of claims data complicated. Historically, the fraction of the unemployed that UI claims data miss is widest exactly when the labor market is strong.

The same picture for the combined entry margin — Reentrants plus New Entrants, the "Labor Force Entry" group of the six-way decomposition — is shown below. Reentrants are the larger and more cyclical of the two, so the combined series sits well above the New-Entrants-only lines and swings more over the cycle, but the recent tilt is the same: the share of the unemployed is elevated while the share of the labor force is much steadier.

```{r entrants-combined-share, fig.width=10, fig.height=8, warning=F, message=F}
#| fig-cap: "Entrants (New Entrants + Reentrants) as a share of total unemployed (top) and as a share of the civilian labor force (bottom), monthly (light) and 12-month moving average (bold). Shaded bands are NBER recessions. Source: BLS via FRED."

ent_col <- "#EFC000"  # Labor Force Entry (gold), per the two-group decomposition

ent_series <- cause_raw |>
  filter(cause %in% c("Reentrants", "New Entrants")) |>
  group_by(date) |>
  summarise(value = sum(value), .groups = "drop") |>
  arrange(date)

ent_u <- ent_series |>
  inner_join(tot_raw, by = "date") |>
  mutate(share = 100 * value / total,
         ma12  = zoo::rollapplyr(share, 12, mean, na.rm = TRUE, fill = NA))

ent_lf <- ent_series |>
  inner_join(lf_raw, by = "date") |>
  mutate(lf_share = 100 * value / lf,
         ma12     = zoo::rollapplyr(lf_share, 12, mean, na.rm = TRUE, fill = NA))

p_ent_u <- ent_u |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = share), color = ent_col, alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = ma12),  color = ent_col, linewidth = 1, alpha = 0.85) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "Share of Unemployed", y = "Share (%)", x = NULL)
p_ent_u <- add_recessions(p_ent_u)

p_ent_lf <- ent_lf |>
  ggplot(aes(x = date)) +
  geom_line(aes(y = lf_share), color = ent_col, alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = ma12),     color = ent_col, linewidth = 1, alpha = 0.85) +
  scale_y_continuous(labels = \(x) paste0(x, "%")) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "Share of Labor Force", y = "Share (%)", x = NULL)
p_ent_lf <- add_recessions(p_ent_lf)

p_ent_u / p_ent_lf +
  plot_annotation(
    title = "Entrants (New + Reentrant): Share of Unemployed vs Share of Labor Force",
    subtitle = "Monthly (light) and 12-month moving average (bold)"
  )
```

```{r ent-share-narrative, include=FALSE}
ent_share_u_mean   <- mean(ent_u$share, na.rm = TRUE)
ent_share_u_last12 <- mean(tail(ent_u$share, 12), na.rm = TRUE)
ent_lf_last12      <- mean(tail(ent_lf$lf_share, 12), na.rm = TRUE)
ent_lf_median      <- median(ent_lf$lf_share, na.rm = TRUE)
```

Over the last twelve months the combined entry margin averaged \~`r sprintf("%.1f", ent_share_u_last12)`% of the unemployed (long-run mean \~`r sprintf("%.1f", ent_share_u_mean)`%) but only \~`r sprintf("%.2f", ent_lf_last12)`% of the labor force (long-run median \~`r sprintf("%.2f", ent_lf_median)`%) — the same elevated-in-the-numerator, steady-in-the-denominator pattern seen for New Entrants alone.

But a high New Entrants share of the unemployed can arise for two very different reasons, and they imply different things about the value of the claims signal. The first is the familiar cyclical pattern. In a healthy labor market, separations fall, the unemployed pool shrinks, and New Entrants make up a bigger share of what is left. In this world, claims are low because separations are low, and claims are blind to a bigger fraction of the unemployed because more of them are ineligible. Onlookers of this labor market could reasonably conclude that it is doing well. 

A second reason is the share could rise because the pool of New Entrants itself is elevated relative to the labor force. In this scenario, finding a first job has gotten harder, for reasons such as weaker hiring, business uncertainty, or automation eating entry-level roles. New Entrants would accumulate in the unemployed pool not because separations have fallen, but because employers aren't hiring them. The share of New Entrants would read high, which under the historical patterns would read as a sign of a strong labor market. But the real story is that the labor market is weakening on the hiring side, concentrated in the groups claims cannot see. These two worlds imply very different things about the claims signal. Telling them apart requires a measure that does not share this eligibility blind spot.

So which world are we in today? The labor-force share panel above leans toward the first: today's elevated share-of-unemployed looks more like a denominator effect from low separations than a pile-up of unemployed entrants relative to the size of the labor force. But the labor-force share alone does not speak to the hiring environment directly. To check whether hiring for entrants is actually keeping pace, we turn to the CFLMI.

## The Flow-Consistent Rate of Unemployment

This is where the Chicago Fed Labor Market Indicators (CFLMI) are helpful. CFLMI combines Shimer (2005, 2012) CPS-based estimates of the separation rate $s$ and finding rate $f$ with a set of real-time indicators via partial least squares. Two of those indicators are continuing-claims intensity (CC/CE) and initial-claims intensity (IC/CE), which enter alongside signals from job postings and household surveys. Higher claims intensities load as more labor-market slack: they pull $s$ up and $f$ down.

That CFLMI does not rely on claims alone is useful for this article. Its inputs include online job postings (from Indeed, Lightcast, and ADP) and household perceptions of how hard it is to find a job (from the Conference Board's "jobs plentiful vs. hard to get" diffusion index and the New York Fed's Survey of Consumer Expectations). These data are useful companion indicators to claims because they compensate for its blind spots by measuring hiring for UI-ineligible and UI-eligible persons alike.  

The figure below plots the CFLMI's published finding rate, separation rate, and flow-consistent unemployment rate from 2008 to the latest release.

```{r cflmi-rates, fig.width=10, fig.height=9, warning=F, message=F}
#| fig-cap: "Chicago Fed Labor Market Indicators: the job finding rate (Hiring Rate for Unemployed Workers), the job separation rate (Layoffs and Other Separations Rate), and the flow-consistent unemployment rate, monthly. Shaded bands are NBER recessions. Source: Federal Reserve Bank of Chicago."

cflmi <- arrow::read_parquet(here("data", "final", "cflmi_rates.parquet")) |>
  mutate(date = as.Date(date))

mk_cflmi_panel <- function(y, title, color) {
  d <- cflmi |> filter(!is.na(.data[[y]]))
  dr <- range(d$date)
  p <- ggplot(d, aes(x = date, y = .data[[y]])) +
    geom_line(color = color, linewidth = 0.8) +
    scale_y_continuous(labels = \(x) paste0(x, "%")) +
    scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
    labs(title = title, y = "Rate (%)", x = NULL)
  add_recessions(p) + coord_cartesian(xlim = dr)
}

p_f   <- mk_cflmi_panel("hiring_rate_uw",     "Job Finding Rate (Hiring Rate for Unemployed Workers)",  ggpubr::get_palette('jco', 3)[1])
p_s   <- mk_cflmi_panel("layoffs_other_seps", "Job Separation Rate (Layoffs and Other Separations Rate)",  ggpubr::get_palette('jco', 3)[2])
p_fcr <- mk_cflmi_panel("fcr",                "Flow-Consistent Unemployment Rate",  ggpubr::get_palette('jco', 3)[3])

p_f / p_s / p_fcr
```

```{r cflmi-narrative, include=FALSE}
# Provenance: all figures below come from cflmi_rates.parquet (build_panels.R
# section 8d), cited inline so the prose tracks the monthly CFLMI release.
cflmi_stat <- function(col) {
  d <- cflmi[!is.na(cflmi[[col]]), ]
  d <- d[order(d$date), ]
  v <- d[[col]]
  list(latest = v[length(v)], median = median(v), mean = mean(v))
}
fcr_s  <- cflmi_stat("fcr")
find_s <- cflmi_stat("hiring_rate_uw")
seps_s <- cflmi_stat("layoffs_other_seps")
# Spring-2025 (Mar–May) finding rate for the 12-month drift sentence.
find_spring25 <- mean(
  cflmi$hiring_rate_uw[format(cflmi$date, "%Y") == "2025" &
                       as.integer(format(cflmi$date, "%m")) %in% 3:5],
  na.rm = TRUE
)
pct1 <- function(x) sprintf("%.1f%%", x)   # FCR / separations (one decimal)
pct0 <- function(x) sprintf("%.0f%%", x)   # finding rate (whole percent)
```

A few things from the CFLMI plot are worth flagging explicitly. The flow-consistent rate of unemployment sits at about `r pct1(fcr_s$latest)` in the latest reading, below its full-sample median (\~`r pct1(fcr_s$median)`) and well below its full-sample mean (\~`r pct1(fcr_s$mean)`). The finding rate sits at roughly `r pct0(find_s$latest)`, above its long-run median (\~`r pct0(find_s$median)`) and its long-run mean (\~`r pct0(find_s$mean)`). Separations are near `r pct1(seps_s$latest)`, below their long-run median of \~`r pct1(seps_s$median)`. Together these readings describe a cyclically strong labor market: low separations, high finding, low flow-consistent unemployment. Because the CFLMI's hiring-side inputs do not share the claims blind spot, this reading is not being manufactured by the same quietness we were worried about. On balance, the first story fits better than the second.

That said, the finding rate has drifted down over the past twelve months, from roughly `r pct0(find_spring25)` in spring 2025 to around `r pct0(find_s$latest)` in the latest release. The drift is gentle and the absolute level is still healthy, but the finding rate is the margin most likely to move first if the hiring-side weakening story starts to take hold. This is a number to keep an eye on moving forward.

## Conclusion

UI claims are an invaluable signal for monitoring the labor market. When unemployment rises because firms are laying people off, claims catch it early, cleanly, and nearly in full. Because claims don't measure the hiring side directly, there is a real theoretical concern that a weakening concentrated in groups UI does not cover could go unseen. Today's elevated New Entrants share is exactly the configuration in which that concern would matter. Reading the CFLMI directly, however, the hiring-side indicators tell a consistent story with claims: the finding rate is still historically high, even if it has drifted down over the past twelve months. On this test, today's high New Entrants share is behaving like the familiar cyclical pattern, not a hidden hiring slowdown. The broader caution stands: when composition and claims diverge, the claims release can go quiet while the underlying picture deteriorates. For now, it isn't flashing that warning, but the finding rate is the margin to keep an eye on.
