---
title: "Pseudo-Out-of-Sample Test for Claims Blindness in High-NE Regimes"
author: "Kristin Butcher, Scott Brave, Aryan Safi"
format:
  html:
    code-fold: true
editor: visual
---

```{r setup, include=F}

library(tidyverse)
library(fredr)
library(lubridate)
library(sandwich)
library(lmtest)
library(ggsci)
library(scales)
library(gt)
library(patchwork)
rm(list=ls())

fredr_set_key("ccd350785a210a7653a20d02ce92d83a")

fredr_retry <- function(..., tries = 5, wait = 3) {
  for (i in seq_len(tries)) {
    res <- tryCatch(fredr(...), error = function(e) e)
    if (!inherits(res, "error")) return(res)
    Sys.sleep(wait)
  }
  stop(res)
}

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )

theme_set(theme_report)

set.seed(4272026)
```

## Research Question

If claims are less informative about aggregate unemployment when the unemployed pool skews toward ineligible groups, then a claims-only forecasting model should produce systematically larger forecast errors in months when the New Entrants share is elevated, relative to months when it is not. This notebook implements that test as a rolling-origin pseudo-out-of-sample exercise, following the principles in Hyndman & Athanasopoulos, *Forecasting Principles and Practice*.

## Target Variable

The unemployment rate is highly persistent. Forecasting UR in levels lets a claims-only model win by predicting "last month's UR," which has nothing to do with claims. Following Hyndman (ch. 9 on stationarity, ch. 5 on benchmarks), we forecast the first difference:

$$
\Delta \text{UR}_t = \text{UR}_t - \text{UR}_{t-1}
$$

in percentage points. Forecasting changes is what makes claims-as-leading-indicator interesting: the question is whether a move in claims predicts the next month's turn in UR, not whether claims help explain the level.

## Data

```{r pull-data, warning=F, message=F}
start_date <- as.Date("1994-01-01")

ur_m <- fredr_retry(series_id = "UNRATE", observation_start = start_date) |>
  transmute(date, UR = value)

claims_monthly <- function(sid) {
  fredr_retry(series_id = sid, observation_start = start_date) |>
    mutate(month = floor_date(date, "month")) |>
    group_by(month) |>
    summarise(value = mean(value, na.rm = TRUE), .groups = "drop") |>
    rename(date = month)
}

ic_m <- claims_monthly("ICSA")   |> rename(IC = value)
cc_m <- claims_monthly("CCSA")   |> rename(CC = value)
ce_m <- claims_monthly("COVEMP") |> rename(CE = value)

total_unemp <- fredr_retry(series_id = "UNEMPLOY", observation_start = start_date) |>
  transmute(date, tot = value)

cause_series <- c(
  tl  = "LNS13023653",
  pjl = "LNS13026638",
  ctj = "LNS13026637",
  jl  = "LNS13023705",
  re  = "LNS13023557",
  ne  = "LNS13023569"
)

cause_m <- map_dfr(names(cause_series), \(nm) {
  fredr_retry(series_id = cause_series[[nm]], observation_start = start_date) |>
    transmute(date, cause = nm, value)
}) |>
  pivot_wider(names_from = cause, values_from = value)

df <- ur_m |>
  inner_join(ic_m,       by = "date") |>
  inner_join(cc_m,       by = "date") |>
  inner_join(ce_m,       by = "date") |>
  inner_join(total_unemp, by = "date") |>
  inner_join(cause_m,    by = "date") |>
  arrange(date) |>
  mutate(
    ne_share  = ne  / tot,
    jl_share  = jl  / tot,
    tl_share  = tl  / tot,
    pjl_share = pjl / tot,
    re_share  = re  / tot,
    ctj_share = ctj / tot,
    IC_CE = IC / CE,
    CC_CE = CC / CE,
    dUR     = UR - lag(UR),
    dlIC_CE = log(IC_CE) - log(lag(IC_CE)),
    dlCC_CE = log(CC_CE) - log(lag(CC_CE)),
    pandemic = date >= as.Date("2020-03-01") & date <= as.Date("2021-12-01")
  ) |>
  mutate(
    dUR_l1         = lag(dUR),
    dUR_l12        = lag(dUR, 12),
    dlIC_CE_l1     = lag(dlIC_CE),
    dlIC_CE_l2     = lag(dlIC_CE, 2),
    dlCC_CE_l1     = lag(dlCC_CE),
    dlCC_CE_l2     = lag(dlCC_CE, 2),
    ne_share_l1    = lag(ne_share)
  )
```

## Variable construction

Raw FRED series: $\text{UR}$ (UNRATE), $\text{IC}$ (ICSA, weekly initial claims averaged to monthly), $\text{CC}$ (CCSA, weekly continuing claims averaged to monthly), $\text{CE}$ (COVEMP, covered employment averaged to monthly), $\text{NE}_t$ (LNS13023569), $\text{Unemp}_t$ (UNEMPLOY).

Constructed variables:

$$
\begin{aligned}
\Delta \text{UR}_t &\;=\; \text{UR}_t - \text{UR}_{t-1} \\[2pt]
\Delta \ln(\text{IC}/\text{CE})_t &\;=\; \ln\!\left(\frac{\text{IC}_t}{\text{CE}_t}\right) - \ln\!\left(\frac{\text{IC}_{t-1}}{\text{CE}_{t-1}}\right) \\[2pt]
\Delta \ln(\text{CC}/\text{CE})_t &\;=\; \ln\!\left(\frac{\text{CC}_t}{\text{CE}_t}\right) - \ln\!\left(\frac{\text{CC}_{t-1}}{\text{CE}_{t-1}}\right) \\[2pt]
\text{NE share}_t &\;=\; \frac{\text{NE}_t}{\text{Unemp}_t}
\end{aligned}
$$

Claims are scaled by covered employment to match the insured-unemployment-rate units used in the main paper; log-differencing removes secular trend and gives a growth-rate interpretation comparable across decades. $\Delta\text{UR}$ is in percentage points. A pandemic indicator $\text{P}_t = \mathbb{1}\{t \in [\text{Mar 2020}, \text{Dec 2021}]\}$ is used to mask training windows and forecast targets.

## Models

At each forecast origin $t$, three parametric models are re-estimated by OLS on the rolling 30-month window $\mathcal{W}_t = \{t-29, \dots, t\}$ and used to forecast $\Delta \text{UR}_{t+1}$:

$$
\text{Model A (claims only):}\quad \Delta \text{UR}_s \;=\; \beta_0 \;+\; \sum_{k=1}^{2} \beta^{\text{IC}}_k \,\Delta\ln(\text{IC}/\text{CE})_{s-k} \;+\; \sum_{k=1}^{2} \beta^{\text{CC}}_k \,\Delta\ln(\text{CC}/\text{CE})_{s-k} \;+\; \varepsilon_s
$$

$$
\text{Model B (claims + NE):}\quad \Delta \text{UR}_s \;=\; \beta_0 \;+\; \sum_{k=1}^{2} \beta^{\text{IC}}_k \,\Delta\ln(\text{IC}/\text{CE})_{s-k} \;+\; \sum_{k=1}^{2} \beta^{\text{CC}}_k \,\Delta\ln(\text{CC}/\text{CE})_{s-k} \;+\; \gamma \,\text{NE share}_{s-1} \;+\; \varepsilon_s
$$

$$
\text{Benchmark AR(1):}\quad \Delta \text{UR}_s \;=\; \alpha_0 \;+\; \alpha_1 \,\Delta \text{UR}_{s-1} \;+\; \varepsilon_s
$$

$$
\text{Random walk:}\quad \widehat{\Delta \text{UR}}_{t+1} \;=\; 0, \qquad \text{Seasonal naive:}\quad \widehat{\Delta \text{UR}}_{t+1} \;=\; \Delta \text{UR}_{t-11}
$$

Model A is deliberately naive: no own-lags of $\Delta \text{UR}$. The question is whether a forecaster looking only at claims does worse when the unemployed pool skews toward ineligible groups; adding UR's own history would let Model A piggyback on UR persistence that has nothing to do with claims, and the "claims-only" label would no longer mean claims-only. The AR(1) benchmark isolates that own-history content separately.

## Rolling-origin protocol

For each candidate origin $t$ with $30 \le t \le T - 1$:

1. Skip if $\text{P}_s = 1$ for any $s \in \mathcal{W}_t \cup \{t+1\}$.
2. Fit Model A, Model B, AR(1) by OLS on $\mathcal{W}_t$.
3. Form $\widehat{\Delta \text{UR}}_{t+1}$ for each model; record error $e^{(m)}_{t+1} = \Delta \text{UR}_{t+1} - \widehat{\Delta \text{UR}}^{(m)}_{t+1}$.
4. Tag target month $t+1$ as **High NE** if $\text{NE share}_t \ge q_{0.75}$, else **Rest**, where $q_{0.75}$ is the 75th percentile of $\text{NE share}$ on the non-pandemic sample. Regime assignment uses information available at the origin, not the target.

## Metrics

For each (model $m$, regime $r$) pair:

$$
\text{MAE}^{(m)}_r \;=\; \frac{1}{n_r} \sum_{t+1 \in r} |e^{(m)}_{t+1}|, \qquad
\text{RMSE}^{(m)}_r \;=\; \sqrt{\frac{1}{n_r} \sum_{t+1 \in r} (e^{(m)}_{t+1})^2},
$$

$$
\text{MASE}^{(m)}_r \;=\; \frac{\text{MAE}^{(m)}_r}{\tfrac{1}{T^\ast - 1} \sum_{s: \text{P}_s = 0,\, \text{P}_{s-1}=0} |\Delta\text{UR}_s - \Delta\text{UR}_{s-1}|}
$$

MASE's denominator is the mean absolute one-month change in $\Delta\text{UR}$ on the non-pandemic full sample, so errors across regimes are put on a common yardstick.

## Inference

**Block bootstrap on MAE differences.** For each model $m$, let $\mathbf{e}^{(m)}_H$ and $\mathbf{e}^{(m)}_R$ be the High-NE and Rest error series. Block bootstrap with block length 6 and $B=2000$ replications gives a sampling distribution of $\text{MAE}(\mathbf{e}^{(m),\star}_H) - \text{MAE}(\mathbf{e}^{(m),\star}_R)$; the 2.5 / 97.5 percentiles form a 95 % CI and the two-sided p-value is $2 \min\{\Pr(d^\star \le 0), \Pr(d^\star \ge 0)\}$.

**Diebold-Mariano test, Model A vs Model B.** With squared-error loss $L^{(m)}_s = (e^{(m)}_s)^2$ and differential $d_s = L^{(A)}_s - L^{(B)}_s$, regress $d_s$ on a constant and test $H_0{:}\ \mathbb{E}[d_s] = 0$ with Newey-West HAC (lag $= \max(h-1, 0) = 0$ for $h=1$).

```{r walk-forward, warning=F, message=F}
walk_forward <- function(df, features, target = "dUR", window = 30, label = "model") {
  n <- nrow(df)
  out <- vector("list", n)
  for (end_train in window:(n - 1)) {
    train_rows <- (end_train - window + 1):end_train
    test_row   <- end_train + 1
    if (any(df$pandemic[train_rows], na.rm = TRUE)) next
    if (isTRUE(df$pandemic[test_row])) next

    cols       <- c(target, features)
    train_data <- df[train_rows, cols, drop = FALSE]
    test_data  <- df[test_row,   cols, drop = FALSE]
    if (anyNA(train_data) || anyNA(test_data)) next

    form <- as.formula(paste(target, "~", paste(features, collapse = " + ")))
    mod  <- lm(form, data = train_data)
    pred <- as.numeric(predict(mod, newdata = test_data))

    out[[test_row]] <- tibble(
      date       = df$date[test_row],
      target_val = df[[target]][test_row],
      forecast   = pred,
      err        = df[[target]][test_row] - pred,
      model      = label
    )
  }
  bind_rows(out)
}

rw_forecast <- function(df, target = "dUR", window = 30, label = "RW") {
  n <- nrow(df)
  out <- vector("list", n)
  for (end_train in window:(n - 1)) {
    train_rows <- (end_train - window + 1):end_train
    test_row   <- end_train + 1
    if (any(df$pandemic[train_rows], na.rm = TRUE)) next
    if (isTRUE(df$pandemic[test_row])) next
    if (is.na(df[[target]][test_row])) next
    out[[test_row]] <- tibble(
      date       = df$date[test_row],
      target_val = df[[target]][test_row],
      forecast   = 0,
      err        = df[[target]][test_row],
      model      = label
    )
  }
  bind_rows(out)
}

snaive_forecast <- function(df, target = "dUR", window = 30, label = "SNaive") {
  n <- nrow(df)
  out <- vector("list", n)
  for (end_train in window:(n - 1)) {
    train_rows <- (end_train - window + 1):end_train
    test_row   <- end_train + 1
    if (any(df$pandemic[train_rows], na.rm = TRUE)) next
    if (isTRUE(df$pandemic[test_row])) next
    if (is.na(df[[target]][test_row]) || is.na(df$dUR_l12[test_row])) next
    if (test_row - 12 >= 1 && isTRUE(df$pandemic[test_row - 12])) next
    pred <- df$dUR_l12[test_row]
    out[[test_row]] <- tibble(
      date       = df$date[test_row],
      target_val = df[[target]][test_row],
      forecast   = pred,
      err        = df[[target]][test_row] - pred,
      model      = label
    )
  }
  bind_rows(out)
}

features_A  <- c("dlIC_CE_l1", "dlIC_CE_l2", "dlCC_CE_l1", "dlCC_CE_l2")
features_B  <- c(features_A, "ne_share_l1")
features_AR <- c("dUR_l1")

fc_A      <- walk_forward(df, features_A,  label = "A (claims)")
fc_B      <- walk_forward(df, features_B,  label = "B (claims+NE)")
fc_AR1    <- walk_forward(df, features_AR, label = "AR(1)")
fc_RW     <- rw_forecast(df,                label = "RW")
fc_SN     <- snaive_forecast(df,            label = "SNaive")

all_fc <- bind_rows(fc_A, fc_B, fc_AR1, fc_RW, fc_SN)
```

## Regime Definition

For each forecast target, we tag by the NE share *at time $t$* (the forecast origin), keeping the regime assignment strictly out-of-sample. The threshold is the 75th percentile of the non-pandemic NE share distribution.

```{r regime-tag, warning=F, message=F}
ne_cutoff <- df |>
  filter(!pandemic) |>
  pull(ne_share) |>
  quantile(0.75, na.rm = TRUE)

regime_lookup <- df |>
  transmute(
    target_date = date,
    ne_share_at_origin = lag(ne_share),
    regime = if_else(lag(ne_share) >= ne_cutoff, "High NE", "Rest")
  )

all_fc <- all_fc |>
  left_join(regime_lookup, by = c("date" = "target_date")) |>
  mutate(regime = factor(regime, levels = c("High NE", "Rest")))

regime_counts <- all_fc |>
  filter(model == "A (claims)") |>
  count(regime)

cat(sprintf("NE share 75th-percentile cutoff (non-pandemic): %.4f\n", ne_cutoff))
print(regime_counts)
```

## Metrics

For each regime × model we compute MAE, RMSE, and MASE. The MASE denominator is the mean absolute one-month change in $\Delta \text{UR}$ on the full non-pandemic sample, so errors are scaled by a common yardstick across regimes.

```{r metrics, warning=F, message=F}
mase_scale <- df |>
  filter(!pandemic) |>
  mutate(d2UR = dUR - lag(dUR)) |>
  pull(d2UR) |>
  abs() |>
  mean(na.rm = TRUE)

metrics_by_regime <- all_fc |>
  group_by(model, regime) |>
  summarise(
    n    = n(),
    mae  = mean(abs(err), na.rm = TRUE),
    rmse = sqrt(mean(err^2, na.rm = TRUE)),
    mase = mean(abs(err), na.rm = TRUE) / mase_scale,
    .groups = "drop"
  ) |>
  arrange(regime, model)

metrics_by_regime |>
  gt(groupname_col = "regime") |>
  fmt_number(columns = c(mae, rmse, mase), decimals = 3) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) |>
  tab_source_note(sprintf(
    "MASE denominator (mean |Δ²UR|): %.4f pp. NE 75th-pct cutoff: %.4f.",
    mase_scale, ne_cutoff
  ))
```

## Inference

### Block bootstrap on MAE difference (same model across regimes)

For each model, block-bootstrap (block length = 6 months) the difference $\text{MAE}_{\text{high}} - \text{MAE}_{\text{rest}}$.

```{r boot-mae-diff, warning=F, message=F}
mbb_sample <- function(e, block_len) {
  n <- length(e)
  nblocks <- ceiling(n / block_len)
  starts  <- sample.int(n - block_len + 1, nblocks, replace = TRUE)
  idx     <- unlist(lapply(starts, \(s) s:(s + block_len - 1)))
  e[idx[seq_len(n)]]
}

boot_mae_diff <- function(fc_one_model, B = 2000, block_len = 6) {
  e_h <- fc_one_model |> filter(regime == "High NE") |> pull(err) |> na.omit()
  e_r <- fc_one_model |> filter(regime == "Rest")    |> pull(err) |> na.omit()
  if (length(e_h) < block_len || length(e_r) < block_len) {
    return(tibble(observed_diff = NA, ci_lo = NA, ci_hi = NA, p_two_sided = NA))
  }
  diffs <- replicate(B, {
    mean(abs(mbb_sample(e_h, block_len))) - mean(abs(mbb_sample(e_r, block_len)))
  })
  observed <- mean(abs(e_h)) - mean(abs(e_r))
  tibble(
    observed_diff = observed,
    ci_lo         = as.numeric(quantile(diffs, 0.025)),
    ci_hi         = as.numeric(quantile(diffs, 0.975)),
    p_two_sided   = 2 * min(mean(diffs <= 0), mean(diffs >= 0))
  )
}

mae_diff_table <- all_fc |>
  group_by(model) |>
  group_modify(~ boot_mae_diff(.x)) |>
  ungroup()

mae_diff_table |>
  gt() |>
  fmt_number(columns = c(observed_diff, ci_lo, ci_hi, p_two_sided), decimals = 3) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) |>
  tab_header(title = "Block bootstrap: MAE(High NE) − MAE(Rest)",
             subtitle = "Block length = 6 months, B = 2000")
```

### Diebold-Mariano: Model A vs Model B within each regime

Does adding the NE share to a claims-only model reduce forecast error, and does the improvement differ across regimes?

```{r dm-test, warning=F, message=F}
dm_test <- function(err_a, err_b, h = 1) {
  d <- err_a^2 - err_b^2
  ok <- !is.na(d)
  d  <- d[ok]
  if (length(d) < 10) {
    return(tibble(dm_stat = NA_real_, dm_p = NA_real_, n = length(d), mean_loss_diff = NA_real_))
  }
  mod <- lm(d ~ 1)
  nw  <- sandwich::NeweyWest(mod, lag = max(h - 1, 0), prewhite = FALSE)
  ct  <- lmtest::coeftest(mod, vcov = nw)
  tibble(
    dm_stat        = ct[1, "t value"],
    dm_p           = ct[1, "Pr(>|t|)"],
    n              = length(d),
    mean_loss_diff = mean(d)
  )
}

ab_joined <- all_fc |>
  filter(model == "A (claims)") |>
  select(date, regime, err_A = err) |>
  inner_join(
    all_fc |> filter(model == "B (claims+NE)") |> select(date, err_B = err),
    by = "date"
  )

dm_by_regime <- ab_joined |>
  group_by(regime) |>
  summarise(dm_test(err_A, err_B), .groups = "drop")

dm_by_regime |>
  gt() |>
  fmt_number(columns = c(dm_stat, dm_p, mean_loss_diff), decimals = 4) |>
  tab_options(
    table.font.size = px(11),
    column_labels.font.weight = "bold",
    column_labels.background.color = "#2c3e50"
  ) |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels()) |>
  tab_header(title = "Diebold-Mariano: Model A vs Model B",
             subtitle = "Positive mean loss diff = Model B (claims+NE) has smaller squared errors")
```

## Presentation

### Absolute forecast error over time

```{r error-plot, fig.width=10, fig.height=5, warning=F, message=F}
ne_ribbon <- df |>
  filter(!pandemic) |>
  mutate(high = ne_share >= ne_cutoff) |>
  transmute(date, high, rle_grp = cumsum(high != lag(high, default = FALSE))) |>
  filter(high) |>
  group_by(rle_grp) |>
  summarise(xmin = min(date), xmax = max(date) + months(1), .groups = "drop")

p_err <- fc_A |>
  ggplot(aes(x = date, y = abs(err))) +
  geom_rect(data = ne_ribbon, aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
            inherit.aes = FALSE, fill = "#EFC000", alpha = 0.25) +
  geom_line(color = "#0073C2", linewidth = 0.4, alpha = 0.75) +
  scale_y_continuous(labels = \(x) paste0(x, " pp")) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  labs(
    title = "Absolute forecast error: Model A (claims only)",
    subtitle = "Shaded bands = months when NE share at origin was in the top quartile",
    x = NULL, y = "|error| (pp of UR)"
  )

p_err
```

### NE share through time, with the cutoff

```{r ne-share-plot, fig.width=10, fig.height=4, warning=F, message=F}
df |>
  filter(!pandemic) |>
  ggplot(aes(x = date, y = ne_share)) +
  geom_line(color = "#0073C2", alpha = 0.4, linewidth = 0.4) +
  geom_hline(yintercept = ne_cutoff, color = "#CD534C", linewidth = 0.6, linetype = "dashed") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 0.1)) +
  scale_x_date(date_breaks = "3 years", date_labels = "%Y") +
  labs(
    title = "NE share with 75th-percentile cutoff",
    x = NULL, y = "NE share of unemployed"
  )
```

## Discussion

The pattern is consistent with the blind-spot story, though the within-model regime gap is not statistically distinguishable from zero on its own.

**The claims-only model is a bit worse in high-NE periods.** Model A's MAE is 0.124 pp when the NE share is in its top quartile and 0.113 pp in the rest. The gap is +0.011 pp — about 10 percent of the rest-period MAE. The block-bootstrap 95 percent CI is roughly $[-0.015, +0.039]$ with a two-sided p-value of 0.43, so on its own this is suggestive, not conclusive.

**The sharper signal is how claims stack up against a naive AR(1).** In "Rest" months, Model A and AR(1) are essentially tied (0.113 vs 0.114). In "High NE" months, AR(1) actually improves (to 0.107) while Model A deteriorates (to 0.124). So the relative accuracy of the claims-only read versus a simple "project last month's UR change forward" rule flips sign precisely when the unemployed pool skews toward groups claims cannot see. A forecaster with only claims in hand does about as well as a forecaster with only UR's own history — except when the ineligible share is high, at which point the UR-history forecaster starts to win.

**Adding NE share to Model A doesn't close the gap.** Model B augments the claims features with lagged NE share. The Diebold-Mariano test on squared errors gives a pooled mean loss difference of $-0.0009$ pp² (p = 0.16), and within-regime DM tests are similarly null. If the blind spot were easy to repair with a single aggregate composition indicator, Model B would do visibly better than Model A in high-NE months. It doesn't — which fits the character of the problem: the ineligible pool is identified by who these people are (work history, earnings, attachment), not by the level of one share variable.

What this exercise does *not* do is settle the level question. Representation ratios show, mechanically from the eligibility rules, that new entrants are outside the IC/CC window. The forecasting test says something narrower: at the monthly horizon and under a simple linear specification, a claims-only nowcast of $\Delta\text{UR}$ is detectably less accurate than a UR-history benchmark in months when the unemployed pool is most weighted toward the group claims cannot see. It's a pattern, not a confidence-interval-certified fact. The block bootstrap says the within-model High-vs-Rest gap could plausibly be zero; what it can't easily explain is why the AR(1) comparison flips sign across regimes.

**One timing caveat.** High-NE months in our sample are concentrated in 2012–2016 (about 57 of 83), with smaller clusters in 2017–2019 and 2024–2025. This is mostly the post-GFC recovery: a low-volatility expansion with a subdued claims signal on both sides of the ratio. A forecast-gap test has less power in that environment than it would in a period with more movement, which is part of why the within-model bootstrap is indeterminate despite the suggestive point estimates.

## Risks & Caveats

1. **Regime imbalance**: high-NE months are concentrated in the post-GFC recovery (2012–2016, 57 of 83 months), with a secondary cluster in 2017–2019 and a new cluster in 2024–2025. This is effectively a test of "subdued expansion" vs everything else, not "high-NE cycle phase" vs "low-NE cycle phase."
2. **Business-cycle confounding**: NE share is correlated with the cycle. High-NE periods are typically expansions with lower UR volatility. MASE partially addresses this via a common denominator, but any gap-based test has less power when the base UR is quiet.
3. **Horizon and frequency**: $\Delta \text{UR}$ at $h=1$ monthly is a demanding outcome — AR(1) already captures most of the predictable variation. A longer horizon or a smoother target (e.g., three-month changes) could reveal differences this test cannot see.
4. **Data vintage**: current-vintage (revised) FRED data, not real-time ALFRED. A real-time exercise would also face claims revisions around benchmarks.
5. **Small post-pandemic window**: any forecast whose 30-month training window is entirely post-pandemic only starts around July 2024, so post-pandemic coverage is ~1.5 years.
