---
title: "Recent College Graduates: Quick Facts"
author: "Kristin Butcher and Aryan Safi"
format:
  html:
    code-fold: true
    embed-resources: true
    toc: true
    toc-depth: 3
editor: visual
---

```{r setup, include=F}
library(data.table)
library(here)
library(ggplot2)
library(ggsci)
library(scales)
library(gt)
library(lubridate)
library(zoo)
library(fredr)

here::i_am("exploring_new_entrants/markdown/rcg_faq.qmd")

fredr_set_key("ccd350785a210a7653a20d02ce92d83a")

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )
theme_set(theme_report)

jco          <- pal_jco()(6)
rcg_col      <- jco[1]
overall_col  <- jco[2]
ylb_col      <- jco[3]

recessions <- data.table(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-01", "2009-06-01", "2020-04-01"))
)

add_recessions <- function(p, alpha = 0.08) {
  p + annotate("rect",
               xmin = recessions$start, xmax = recessions$end,
               ymin = -Inf, ymax = Inf,
               fill = "grey40", alpha = alpha)
}

weighted_median <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (!any(ok)) return(NA_real_)
  x <- x[ok]; w <- w[ok]
  ord <- order(x)
  x <- x[ord]; w <- w[ord]
  cw <- cumsum(w) / sum(w)
  x[which(cw >= 0.5)[1]]
}

weighted_mean <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (!any(ok)) return(NA_real_)
  sum(x[ok] * w[ok]) / sum(w[ok])
}

# 12-month rolling helpers (right-aligned, fill leading NAs)
roll_mean12   <- function(x) frollmean(x, 12, align = "right", na.rm = TRUE)
roll_median12 <- function(x) zoo::rollapplyr(x, 12, median, na.rm = TRUE, fill = NA)

fmt_pct <- function(x, d = 1) paste0(formatC(x, digits = d, format = "f"), "%")

# Shared styling for gt tables
style_gt <- function(g) {
  g |>
    tab_options(
      table.font.size = px(11),
      column_labels.font.weight = "bold",
      column_labels.background.color = "#2c3e50"
    ) |>
    tab_style(style = cell_text(color = "white"),
              locations = cells_column_labels())
}
```

## Definition

A **Recent College Graduate (RCG)** is a person who is **age 21–26** and holds a **bachelor's degree or higher** (`AGE %in% 21:26` and `EDUC >= 111` in IPUMS CPS). This captures people near the usual age of bachelor's-degree completion and the first several years afterward.

A **Young Less-Bachelor's (YLB)** person is **age 18–25** with **less than a bachelor's degree** (`AGE %in% 18:25` and `EDUC < 111`). This is the age-peer comparison group for RCGs who did not complete a four-year degree; several questions below add a YLB series alongside RCG and overall figures.

All shares, rates, means, and medians use the IPUMS basic monthly person weight (`WTFINL`). The sample is the regular monthly basic CPS from January 1994 through the latest extract month, with the ASEC supplement excluded.

```{r data-prep, message=F, warning=F}
csv_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.csv.gz")

cps <- fread(csv_path)

cps[, `:=`(
  date     = make_date(YEAR, MONTH, 1L),
  is_rcg   = AGE >= 21 & AGE <= 26 & EDUC >= 111,
  is_ylb   = AGE >= 18 & AGE <= 25 & EDUC < 111,
  is_lf    = LABFORCE == 2,
  is_emp   = EMPSTAT %in% c(10L, 12L),
  is_unemp = EMPSTAT %in% c(20L, 21L, 22L)
)]
cps[, is_ne_re := is_unemp & WHYUNEMP %in% c(5L, 6L)]
cps[, is_ne    := is_unemp & WHYUNEMP == 6L]
cps[, is_re    := is_unemp & WHYUNEMP == 5L]
cps[, dur_wks  := fifelse(is_unemp & DURUNEMP < 999,
                          as.numeric(DURUNEMP), NA_real_)]

latest_month    <- cps[, max(date)]
latest_12_start <- latest_month - months(11)
format_window   <- paste0(format(latest_12_start, "%b %Y"),
                          " – ", format(latest_month, "%b %Y"))

# ── Monthly aggregates for shares / rates ───────────────────────────────────
monthly <- cps[, .(
  rcg_unemp_w   = sum(WTFINL * (is_rcg & is_unemp), na.rm = TRUE),
  rcg_lf_w      = sum(WTFINL * (is_rcg & is_lf),    na.rm = TRUE),
  ylb_unemp_w   = sum(WTFINL * (is_ylb & is_unemp), na.rm = TRUE),
  ylb_lf_w      = sum(WTFINL * (is_ylb & is_lf),    na.rm = TRUE),
  total_unemp_w = sum(WTFINL * is_unemp,            na.rm = TRUE),
  total_lf_w    = sum(WTFINL * is_lf,               na.rm = TRUE),
  ne_re_w       = sum(WTFINL * is_ne_re,            na.rm = TRUE),
  ne_w          = sum(WTFINL * is_ne,               na.rm = TRUE),
  re_w          = sum(WTFINL * is_re,               na.rm = TRUE)
), by = date]
setorder(monthly, date)
monthly[, `:=`(
  rcg_unemp_rate     = 100 * rcg_unemp_w / rcg_lf_w,
  ylb_unemp_rate     = 100 * ylb_unemp_w / ylb_lf_w,
  overall_unemp_rate = 100 * total_unemp_w / total_lf_w,
  rcg_share_unemp    = 100 * rcg_unemp_w / total_unemp_w,
  rcg_share_lf       = 100 * rcg_lf_w    / total_lf_w,
  ylb_share_unemp    = 100 * ylb_unemp_w / total_unemp_w,
  ylb_share_lf       = 100 * ylb_lf_w    / total_lf_w,
  ne_re_share        = 100 * ne_re_w     / total_unemp_w,
  ne_share           = 100 * ne_w        / total_unemp_w,
  re_share           = 100 * re_w        / total_unemp_w,
  ne_re_lf_share     = 100 * ne_re_w     / total_lf_w,
  rcg_lf_millions    = rcg_lf_w / 1e6
)]

# ── Monthly aggregates for duration ──────────────────────────────────────────
unemp_dt    <- cps[is_unemp & !is.na(dur_wks)]
monthly_dur <- unemp_dt[, .(
  rcg_mean       = weighted_mean(dur_wks[is_rcg], WTFINL[is_rcg]),
  rcg_median     = weighted_median(dur_wks[is_rcg], WTFINL[is_rcg]),
  ylb_mean       = weighted_mean(dur_wks[is_ylb], WTFINL[is_ylb]),
  ylb_median     = weighted_median(dur_wks[is_ylb], WTFINL[is_ylb]),
  overall_mean   = weighted_mean(dur_wks, WTFINL),
  overall_median = weighted_median(dur_wks, WTFINL)
), by = date]
setorder(monthly_dur, date)

# ── Window-summary helpers (for the headline tables) ────────────────────────
window_label_12  <- paste0("Latest 12 months (", format_window, ")")
window_label_all <- paste0("Pooled Jan 1994–", format(latest_month, "%b %Y"))

in_window_12 <- function(d) d >= latest_12_start & d <= latest_month

# ── Per-question time-series panel helper (for share/rate plots) ────────────
rate_panel <- function(dt, col, title, color,
                       y_label = "Percent", y_suffix = "%") {
  d <- copy(dt)
  d[, .v   := get(col)]
  d[, .ma  := roll_mean12(.v)]
  d[, .med := roll_median12(.v)]
  p <- ggplot(d, aes(x = date)) +
    geom_line(aes(y = .v),   color = color, alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = .ma),  color = color, linewidth = 1.0, alpha = 0.9) +
    geom_line(aes(y = .med), color = color, linewidth = 0.9, alpha = 0.9,
              linetype = "dashed") +
    scale_y_continuous(labels = \(x) paste0(x, y_suffix)) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly (light); 12-mo rolling mean (solid); 12-mo rolling median (dashed). NBER recessions shaded.",
         y = y_label, x = NULL)
  add_recessions(p)
}

# ── Two-series comparison panel (e.g., RCG vs Overall) ──────────────────────
compare_panel <- function(dt, cols, labels, colors, title, y_label = "Percent",
                          y_suffix = "%") {
  d <- copy(dt)[, c("date", cols), with = FALSE]
  d_long <- melt(d, id.vars = "date", measure.vars = cols,
                 variable.name = "series", value.name = "value")
  d_long[, series := factor(series, levels = cols, labels = labels)]
  d_long[, ma := roll_mean12(value), by = series]

  p <- ggplot(d_long, aes(x = date, color = series)) +
    geom_line(aes(y = value), alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = ma),    linewidth = 1.0, alpha = 0.9) +
    scale_color_manual(values = setNames(colors, labels)) +
    scale_y_continuous(labels = \(x) paste0(x, y_suffix)) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly (light); 12-month rolling mean (bold). NBER recessions shaded.",
         y = y_label, x = NULL)
  add_recessions(p)
}
```

The most recent month in the extract is **`r format(latest_month, "%B %Y")`**. Below, *latest 12 months* means **`r format_window`** and *pooled* means **January 1994–`r format(latest_month, "%B %Y")`**.

## Q1. What percentage of recent college graduates are unemployed?

Unemployed RCGs divided by RCGs in the labor force, weighted by `WTFINL`.

```{r q1-table}
q1_value <- function(dt) {
  d <- dt[is_rcg & is_lf]
  d[, sum(WTFINL * is_unemp, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q1_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG unemployment rate` = c(
    fmt_pct(100 * q1_value(cps[in_window_12(date)])),
    fmt_pct(100 * q1_value(cps))
  )
)

q1_tbl |> gt() |> style_gt()
```

::: panel-tabset
### RCG only

```{r q1-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_unemp_rate",
           "RCG unemployment rate (% of RCG in labor force)", jco[1])
```

### RCG vs Overall

```{r q1-ts-compare, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("rcg_unemp_rate", "overall_unemp_rate"),
  labels = c("RCG", "Overall"),
  colors = c(rcg_col, overall_col),
  title  = "Unemployment rate: RCG vs Overall"
)
```

### RCG vs YLB vs Overall

```{r q1-ts-compare-ylb, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("rcg_unemp_rate", "ylb_unemp_rate", "overall_unemp_rate"),
  labels = c("RCG", "YLB", "Overall"),
  colors = c(rcg_col, ylb_col, overall_col),
  title  = "Unemployment rate: RCG vs YLB vs Overall"
)
```
:::

## Q2. What percentage of the unemployed are recent college graduates?

```{r q2-table}
q2_value <- function(dt) {
  d <- dt[is_unemp == TRUE]
  d[, sum(WTFINL * is_rcg, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}
q2_ylb_value <- function(dt) {
  d <- dt[is_unemp == TRUE]
  d[, sum(WTFINL * is_ylb, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q2_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG share of unemployed` = c(
    fmt_pct(100 * q2_value(cps[in_window_12(date)])),
    fmt_pct(100 * q2_value(cps))
  ),
  `YLB share of unemployed` = c(
    fmt_pct(100 * q2_ylb_value(cps[in_window_12(date)])),
    fmt_pct(100 * q2_ylb_value(cps))
  )
)

q2_tbl |> gt() |> style_gt()
```

::: panel-tabset
### RCG share of unemployed

```{r q2-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_share_unemp",
           "RCG share of the unemployed (% of all unemployed)", jco[2])
```

### Share of unemployed vs Share of LF

```{r q2-ts-compare, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("rcg_share_unemp", "rcg_share_lf"),
  labels = c("Share of unemployed", "Share of LF"),
  colors = c(jco[2], jco[3]),
  title  = "RCG: share of unemployed vs share of labor force"
)
```

When the "share of unemployed" line sits above the "share of LF" line, RCGs are *over*-represented in the unemployed pool relative to their presence in the labor force, and vice versa.

### YLB: share of unemployed vs share of LF

```{r q2-ts-compare-ylb, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("ylb_share_unemp", "ylb_share_lf"),
  labels = c("Share of unemployed", "Share of LF"),
  colors = c(jco[2], jco[3]),
  title  = "YLB: share of unemployed vs share of labor force"
)
```

The same reading applies to YLB: when the "share of unemployed" line sits above the "share of LF" line, young workers without a bachelor's are over-represented among the unemployed relative to their share of the labor force.
:::

## Q3. What percentage of the labor force are recent college graduates?

```{r q3-table}
q3_value <- function(dt) {
  d <- dt[is_lf == TRUE]
  d[, sum(WTFINL * is_rcg, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q3_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG share of labor force` = c(
    fmt_pct(100 * q3_value(cps[in_window_12(date)])),
    fmt_pct(100 * q3_value(cps))
  )
)

q3_tbl |> gt() |> style_gt()
```

::: panel-tabset
### Share of LF

```{r q3-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_share_lf",
           "RCG share of the labor force (% of LF)", jco[3])
```

### Counts (millions)

```{r q3-ts-counts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_lf_millions",
           "RCG in the labor force (weighted, millions of persons)", jco[3],
           y_label = "Millions of persons", y_suffix = "M")
```
:::

## Q4. What share of the unemployed — and of the labor force — are New Entrants or Reentrants?

Computed from CPS `WHYUNEMP`: codes 5 (Reentrant) and 6 (New Entrant). Two denominators answer two different questions. Dividing by **all unemployed people** (`EMPSTAT %in% 20:22`) shows how much of the unemployed *pool* these entry categories make up — a composition measure that is roughly independent of how high unemployment is. Dividing by **everyone in the labor force** (`LABFORCE == 2`) gives the slice of the headline unemployment rate that entrants and re-entrants contribute, so it rises and falls with the business cycle.

```{r q4-table}
q4_value <- function(dt) {
  d <- dt[is_unemp == TRUE]
  d[, sum(WTFINL * is_ne_re, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}
ne_re_lf_value <- function(dt) {
  d <- dt[is_lf == TRUE]
  d[, sum(WTFINL * is_ne_re, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q4_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `Share of unemployed` = c(
    fmt_pct(100 * q4_value(cps[in_window_12(date)])),
    fmt_pct(100 * q4_value(cps))
  ),
  `Share of labor force` = c(
    fmt_pct(100 * ne_re_lf_value(cps[in_window_12(date)])),
    fmt_pct(100 * ne_re_lf_value(cps))
  )
)

q4_tbl |> gt() |> style_gt()
```

::: panel-tabset
### Overlay: unemployed vs labor force

```{r q4-ts-overlay, fig.width=10, fig.height=4, message=F, warning=F}
# Share-of-unemployed (~15-25%) and share-of-labor-force (~1-2%) differ by
# roughly the unemployment rate, so the labor-force series is rescaled onto a
# secondary right-hand axis (sec_axis divides the scaling factor back out).
sf <- mean(monthly$ne_re_share, na.rm = TRUE) /
      mean(monthly$ne_re_lf_share, na.rm = TRUE)

ov_long <- rbindlist(list(
  monthly[, .(date, series = "Share of unemployed",  value = ne_re_share)],
  monthly[, .(date, series = "Share of labor force", value = ne_re_lf_share * sf)]
))
ov_long[, series := factor(series,
                           levels = c("Share of unemployed", "Share of labor force"))]
ov_long[, ma := roll_mean12(value), by = series]

ov_cols <- c("Share of unemployed"  = jco[4],
             "Share of labor force" = jco[2])

p_ov <- ggplot(ov_long, aes(x = date, color = series)) +
  geom_line(aes(y = value), alpha = 0.25, linewidth = 0.4) +
  geom_line(aes(y = ma),    linewidth = 1.0, alpha = 0.9) +
  scale_color_manual(values = ov_cols) +
  scale_y_continuous(
    name     = "Share of unemployed",
    labels   = \(x) paste0(x, "%"),
    sec.axis = sec_axis(~ . / sf, name = "Share of labor force",
                        labels = \(x) paste0(x, "%"))
  ) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "NE + Reentrant: share of unemployed vs share of labor force",
       subtitle = "Monthly (light); 12-month rolling mean (bold). Left axis = share of unemployed; right axis = share of labor force. NBER recessions shaded.",
       x = NULL)
add_recessions(p_ov)
```

### Share of the unemployed

```{r q4-ts-unemp, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "ne_re_share",
           "NE + Reentrant share of the unemployed (%)", jco[4])
```

### Share of the labor force

```{r q4-ts-lf, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "ne_re_lf_share",
           "NE + Reentrant share of the labor force (%)", jco[2])
```

### Decomposed (NE vs Reentrant)

```{r q4-ts-decomposed, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("ne_share", "re_share"),
  labels = c("New Entrants", "Reentrants"),
  colors = c(jco[4], jco[5]),
  title  = "NE-only and Reentrant-only shares of the unemployed"
)
```
:::

## Q5. How long are recent college graduates unemployed compared to everyone else?

Weighted **mean** and **median** weeks unemployed (`DURUNEMP`) for unemployed RCGs and for all unemployed people.

```{r q5-table, message=F, warning=F}
q5_value <- function(dt) {
  rcg <- dt[is_unemp & is_rcg]
  ylb <- dt[is_unemp & is_ylb]
  all <- dt[is_unemp == TRUE]
  data.table(
    `RCG mean (weeks)`       = weighted_mean(rcg$dur_wks,  rcg$WTFINL),
    `RCG median (weeks)`     = weighted_median(rcg$dur_wks, rcg$WTFINL),
    `YLB mean (weeks)`       = weighted_mean(ylb$dur_wks,  ylb$WTFINL),
    `YLB median (weeks)`     = weighted_median(ylb$dur_wks, ylb$WTFINL),
    `Overall mean (weeks)`   = weighted_mean(all$dur_wks,   all$WTFINL),
    `Overall median (weeks)` = weighted_median(all$dur_wks, all$WTFINL)
  )
}

q5_tbl <- rbindlist(list(
  cbind(data.table(Window = window_label_12),  q5_value(cps[in_window_12(date)])),
  cbind(data.table(Window = window_label_all), q5_value(cps))
))

q5_tbl |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 1) |>
  style_gt()
```

```{r q5-ts-setup, include=F}
# Reshape to long for clean ggplot legend (used by both mean and median tabs)
dur_long_mean <- melt(monthly_dur,
                      id.vars = "date",
                      measure.vars = c("rcg_mean", "ylb_mean", "overall_mean"),
                      variable.name = "group", value.name = "value")
dur_long_mean[, group := factor(group,
                                levels = c("rcg_mean", "ylb_mean", "overall_mean"),
                                labels = c("RCG", "YLB", "Overall"))]
dur_long_mean[, ma := roll_mean12(value), by = group]

dur_long_med <- melt(monthly_dur,
                     id.vars = "date",
                     measure.vars = c("rcg_median", "ylb_median", "overall_median"),
                     variable.name = "group", value.name = "value")
dur_long_med[, group := factor(group,
                               levels = c("rcg_median", "ylb_median", "overall_median"),
                               labels = c("RCG", "YLB", "Overall"))]
dur_long_med[, ma := roll_mean12(value), by = group]

dur_cols <- c(RCG = rcg_col, YLB = ylb_col, Overall = overall_col)

make_dur_panel <- function(d, title) {
  p <- ggplot(d, aes(x = date, color = group)) +
    geom_line(aes(y = value), alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = ma),    linewidth = 1.0, alpha = 0.9) +
    scale_color_manual(values = dur_cols) +
    scale_y_continuous(labels = \(x) paste0(x, " wks")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly weighted value (light); 12-month rolling mean (bold). NBER recessions shaded.",
         y = "Weeks", x = NULL)
  add_recessions(p)
}
```

::: panel-tabset
### Mean weeks

```{r q5-ts-mean, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_panel(dur_long_mean, "Mean weeks unemployed: RCG vs YLB vs overall")
```

### Median weeks

```{r q5-ts-median, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_panel(dur_long_med, "Median weeks unemployed: RCG vs YLB vs overall")
```
:::

## Q6. When unemployed RCGs are asked *why*, what do they say?

Q4 asked what share of *all* unemployed people are New Entrants or Reentrants. This question conditions on being unemployed and crosstabs **RCG status** against the CPS reason-for-unemployment (`WHYUNEMP`). Each column is a distribution over reasons that **sums to 100% within that group**, so the table reads as: *of unemployed RCGs, what percent give each reason?* — and the same for non-RCGs, for comparison.

A new entrant has **never held a full-time job**; a reentrant has worked before but was out of the labor force just prior to looking again. For a 21–26-year-old with a fresh bachelor's degree, "new entrant" is the answer that most directly describes someone entering the job market for the first time after school.

```{r q6-setup}
why_levels <- c(1L, 2L, 3L, 4L, 5L, 6L)
why_labels <- c("Temporary layoff",        # WHYUNEMP 1: job loser – on layoff
                "Permanent job loser",     # WHYUNEMP 2: other job loser
                "Completed temp job",      # WHYUNEMP 3: temporary job ended
                "Job leaver",              # WHYUNEMP 4
                "Reentrant",               # WHYUNEMP 5
                "New entrant")             # WHYUNEMP 6

# Weighted distribution over WHYUNEMP reasons within a group of the unemployed.
# Denominator = unemployed in the group with a valid reason code, so the
# returned vector of percentages sums to 100.
reason_share <- function(dt, rcg_flag) {
  d   <- dt[is_unemp & is_rcg == rcg_flag & WHYUNEMP %in% why_levels]
  tot <- d[, sum(WTFINL, na.rm = TRUE)]
  vapply(why_levels, function(k)
    100 * d[WHYUNEMP == k, sum(WTFINL, na.rm = TRUE)] / tot,
    numeric(1))
}

# Same distribution, but for the unemployed in an arbitrary group flag column
# (e.g. "is_ylb"). Denominator = that group's unemployed with a valid reason.
reason_share_col <- function(dt, col) {
  d   <- dt[is_unemp & get(col) == TRUE & WHYUNEMP %in% why_levels]
  tot <- d[, sum(WTFINL, na.rm = TRUE)]
  vapply(why_levels, function(k)
    100 * d[WHYUNEMP == k, sum(WTFINL, na.rm = TRUE)] / tot,
    numeric(1))
}
```

```{r q6-table}
q6_12  <- list(rcg = reason_share(cps[in_window_12(date)], TRUE),
               non = reason_share(cps[in_window_12(date)], FALSE),
               ylb = reason_share_col(cps[in_window_12(date)], "is_ylb"))
q6_all <- list(rcg = reason_share(cps, TRUE),
               non = reason_share(cps, FALSE),
               ylb = reason_share_col(cps, "is_ylb"))

q6_tbl <- data.table(
  Reason             = why_labels,
  `RCG (12mo)`       = q6_12$rcg,
  `Non-RCG (12mo)`   = q6_12$non,
  `YLB (12mo)`       = q6_12$ylb,
  `RCG (pooled)`     = q6_all$rcg,
  `Non-RCG (pooled)` = q6_all$non,
  `YLB (pooled)`     = q6_all$ylb
)

q6_tbl |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 1, pattern = "{x}%") |>
  tab_spanner(label = window_label_12,
              columns = c(`RCG (12mo)`, `Non-RCG (12mo)`, `YLB (12mo)`)) |>
  tab_spanner(label = window_label_all,
              columns = c(`RCG (pooled)`, `Non-RCG (pooled)`, `YLB (pooled)`)) |>
  cols_label(
    `RCG (12mo)` = "RCG", `Non-RCG (12mo)` = "Non-RCG", `YLB (12mo)` = "YLB",
    `RCG (pooled)` = "RCG", `Non-RCG (pooled)` = "Non-RCG", `YLB (pooled)` = "YLB"
  ) |>
  style_gt()
```

Reading the table: each column answers "of this group's unemployed, what percent give each reason?" Comparing the **New entrant** and **Reentrant** rows across the RCG and Non-RCG columns shows whether unemployed RCGs are unusually likely to be entering or re-entering the labor force rather than having lost or left a job.

```{r q6-ts-prep, message=F, warning=F}
# Monthly reason mix *within RCG unemployed* (denominator = RCG unemployed).
monthly_rcg_reason <- cps[, .(
  rcg_unemp_w = sum(WTFINL * (is_rcg & is_unemp), na.rm = TRUE),
  rcg_ne_w    = sum(WTFINL * (is_rcg & is_ne),    na.rm = TRUE),
  rcg_re_w    = sum(WTFINL * (is_rcg & is_re),    na.rm = TRUE)
), by = date]
setorder(monthly_rcg_reason, date)
monthly_rcg_reason[, `:=`(
  rcg_ne_share    = 100 * rcg_ne_w / rcg_unemp_w,
  rcg_re_share    = 100 * rcg_re_w / rcg_unemp_w,
  rcg_ne_re_share = 100 * (rcg_ne_w + rcg_re_w) / rcg_unemp_w
)]

# Same NE + Reentrant share, but within YLB unemployed (denominator = YLB unemp).
monthly_ylb_reason <- cps[, .(
  ylb_unemp_w = sum(WTFINL * (is_ylb & is_unemp), na.rm = TRUE),
  ylb_ne_w    = sum(WTFINL * (is_ylb & is_ne),    na.rm = TRUE),
  ylb_re_w    = sum(WTFINL * (is_ylb & is_re),    na.rm = TRUE)
), by = date]
setorder(monthly_ylb_reason, date)
monthly_ylb_reason[, ylb_ne_re_share := 100 * (ylb_ne_w + ylb_re_w) / ylb_unemp_w]
```

::: panel-tabset
### NE + Reentrant share of RCG unemployed

```{r q6-ts-combined, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_rcg_reason, "rcg_ne_re_share",
           "New Entrant + Reentrant share of RCG unemployed (%)", jco[4])
```

### Decomposed (NE vs Reentrant)

```{r q6-ts-decomposed, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly_rcg_reason,
  cols   = c("rcg_ne_share", "rcg_re_share"),
  labels = c("New Entrants", "Reentrants"),
  colors = c(jco[4], jco[5]),
  title  = "Among RCG unemployed: New Entrant vs Reentrant share"
)
```

### RCG vs all unemployed

```{r q6-ts-vs-all, fig.width=10, fig.height=4, message=F, warning=F}
q6_cmp <- merge(
  monthly_rcg_reason[, .(date, rcg_ne_re_share)],
  monthly[, .(date, ne_re_share)],
  by = "date"
)
compare_panel(
  q6_cmp,
  cols   = c("rcg_ne_re_share", "ne_re_share"),
  labels = c("Among RCG unemployed", "Among all unemployed"),
  colors = c(rcg_col, overall_col),
  title  = "NE + Reentrant share: RCG unemployed vs all unemployed"
)
```

### RCG vs all unemployed vs YLB

```{r q6-ts-vs-all-ylb, fig.width=10, fig.height=4, message=F, warning=F}
q6_cmp3 <- Reduce(function(x, y) merge(x, y, by = "date"), list(
  monthly_rcg_reason[, .(date, rcg_ne_re_share)],
  monthly[, .(date, ne_re_share)],
  monthly_ylb_reason[, .(date, ylb_ne_re_share)]
))
compare_panel(
  q6_cmp3,
  cols   = c("rcg_ne_re_share", "ne_re_share", "ylb_ne_re_share"),
  labels = c("Among RCG unemployed", "Among all unemployed", "Among YLB unemployed"),
  colors = c(rcg_col, overall_col, ylb_col),
  title  = "NE + Reentrant share: RCG vs all unemployed vs YLB"
)
```
:::

::: callout-warning
The RCG-unemployed pool is small in any single month, so the New Entrant and Reentrant counts behind the RCG monthly lines can be thin (the unweighted sample sizes are reported in the appendix's *Recoded indicators* table). Lean on the 12-month rolling mean (bold) rather than the raw monthly line (light), and on the pooled column of the table above.
:::

::: callout-note
Source: IPUMS CPS basic monthly microdata, January 1994–`r format(latest_month, "%b %Y")`. ASEC samples excluded. All estimates weighted by `WTFINL`.
:::

## Q7. What percentage of New Entrants and Reentrants are recent college graduates?

Q6 conditioned on being an unemployed RCG and asked what reason they give. This question turns the crosstab around: it conditions on the **reason** and asks what share of that group is RCG. The denominators are the unemployed New Entrants (`WHYUNEMP == 6`) and the unemployed Reentrants (`WHYUNEMP == 5`); the numerators are the RCGs within each.

```{r q7-table}
# Share of a reason-group's unemployed that are RCG.
# Denominator = unemployed with that reason; numerator = RCGs among them.
rcg_share_of <- function(dt, flag_col) {
  d <- dt[is_unemp == TRUE & get(flag_col) == TRUE]
  d[, sum(WTFINL * is_rcg, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q7_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG share of New Entrants` = c(
    fmt_pct(100 * rcg_share_of(cps[in_window_12(date)], "is_ne")),
    fmt_pct(100 * rcg_share_of(cps, "is_ne"))
  ),
  `RCG share of Reentrants` = c(
    fmt_pct(100 * rcg_share_of(cps[in_window_12(date)], "is_re")),
    fmt_pct(100 * rcg_share_of(cps, "is_re"))
  )
)

q7_tbl |> gt() |> style_gt()
```

```{r q7-ts-prep, message=F, warning=F}
# Monthly RCG share *within* each reason group (denominator = that group).
monthly_rcg_in_entrants <- cps[, .(
  ne_w        = sum(WTFINL * is_ne,             na.rm = TRUE),
  re_w        = sum(WTFINL * is_re,             na.rm = TRUE),
  rcg_in_ne_w = sum(WTFINL * (is_ne & is_rcg),  na.rm = TRUE),
  rcg_in_re_w = sum(WTFINL * (is_re & is_rcg),  na.rm = TRUE)
), by = date]
setorder(monthly_rcg_in_entrants, date)
monthly_rcg_in_entrants[, `:=`(
  rcg_in_ne_share = 100 * rcg_in_ne_w / ne_w,
  rcg_in_re_share = 100 * rcg_in_re_w / re_w
)]
```

::: panel-tabset
### RCG share of New Entrants

```{r q7-ts-ne, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_rcg_in_entrants, "rcg_in_ne_share",
           "RCG share of unemployed New Entrants (%)", jco[4])
```

### RCG share of Reentrants

```{r q7-ts-re, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_rcg_in_entrants, "rcg_in_re_share",
           "RCG share of unemployed Reentrants (%)", jco[5])
```

### Both

```{r q7-ts-both, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly_rcg_in_entrants,
  cols   = c("rcg_in_ne_share", "rcg_in_re_share"),
  labels = c("Share of New Entrants", "Share of Reentrants"),
  colors = c(jco[4], jco[5]),
  title  = "RCG share within each entry group"
)
```
:::

::: callout-warning
These denominators (unemployed New Entrants, unemployed Reentrants) are themselves modest in any single month, so the monthly RCG-share lines are noisy. Lean on the 12-month rolling mean (bold) and the pooled column of the table.
:::

## Q8. How long are people unemployed, by reason for unemployment?

Q5 split duration of unemployment by *who* the person is (RCG, YLB, overall); this question splits it by *why* they are unemployed. Each row is a `WHYUNEMP` category, and the cells are the weighted **mean** and **median** weeks unemployed (`DURUNEMP`) among that reason's unemployed. New Entrants and Reentrants have no prior job to lose, so their "duration" is time spent searching since they began looking — useful context for the entry-focused questions above.

```{r q8-table, message=F, warning=F}
# Weighted mean/median weeks unemployed within each WHYUNEMP reason.
# Rows are reordered to why_levels; reasons absent from the window come back NA.
dur_reason_summary <- function(dt) {
  d <- dt[is_unemp & !is.na(dur_wks) & WHYUNEMP %in% why_levels]
  out <- d[, .(
    mean_wks   = weighted_mean(dur_wks, WTFINL),
    median_wks = weighted_median(dur_wks, WTFINL)
  ), by = WHYUNEMP]
  out[match(why_levels, WHYUNEMP)]
}

s12  <- dur_reason_summary(cps[in_window_12(date)])
sall <- dur_reason_summary(cps)

q8_tbl <- data.table(
  Reason            = why_labels,
  `Mean (12mo)`     = s12$mean_wks,
  `Median (12mo)`   = s12$median_wks,
  `Mean (pooled)`   = sall$mean_wks,
  `Median (pooled)` = sall$median_wks
)

q8_tbl |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 1) |>
  tab_spanner(label = window_label_12,
              columns = c(`Mean (12mo)`, `Median (12mo)`)) |>
  tab_spanner(label = window_label_all,
              columns = c(`Mean (pooled)`, `Median (pooled)`)) |>
  cols_label(
    `Mean (12mo)` = "Mean (wks)", `Median (12mo)` = "Median (wks)",
    `Mean (pooled)` = "Mean (wks)", `Median (pooled)` = "Median (wks)"
  ) |>
  style_gt()
```

```{r q8-ts-prep, message=F, warning=F}
# Monthly weighted mean/median weeks unemployed, by WHYUNEMP reason.
monthly_dur_reason <- cps[is_unemp & !is.na(dur_wks) & WHYUNEMP %in% why_levels, .(
  mean_wks   = weighted_mean(dur_wks, WTFINL),
  median_wks = weighted_median(dur_wks, WTFINL)
), by = .(date, WHYUNEMP)]
setorder(monthly_dur_reason, WHYUNEMP, date)
monthly_dur_reason[, reason := factor(WHYUNEMP, levels = why_levels, labels = why_labels)]
monthly_dur_reason[, `:=`(
  mean_ma   = roll_mean12(mean_wks),
  median_ma = roll_mean12(median_wks)
), by = WHYUNEMP]

reason_cols <- setNames(jco, why_labels)

# Six reasons overlaid would be unreadable as raw monthly lines, so only the
# 12-month rolling mean is drawn for each reason.
make_dur_reason_panel <- function(ma_col, title) {
  d <- copy(monthly_dur_reason)
  d[, .y := get(ma_col)]
  p <- ggplot(d, aes(x = date, y = .y, color = reason)) +
    geom_line(linewidth = 0.9, alpha = 0.9) +
    scale_color_manual(values = reason_cols) +
    scale_y_continuous(labels = \(x) paste0(x, " wks")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "12-month rolling mean of the monthly weighted value. NBER recessions shaded.",
         y = "Weeks", x = NULL)
  add_recessions(p)
}
```

::: panel-tabset
### Mean weeks

```{r q8-ts-mean, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_reason_panel("mean_ma", "Mean weeks unemployed, by reason for unemployment")
```

### Median weeks

```{r q8-ts-median, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_reason_panel("median_ma", "Median weeks unemployed, by reason for unemployment")
```
:::

::: callout-note
Duration here is total weeks unemployed as of the survey, not completed spells, so it is a point-in-time snapshot rather than how long a typical spell ultimately lasts. Permanent job losers and reentrants are the larger reason groups; New Entrants and Job Leavers can be thin in a single month, so lean on the rolling means and the pooled columns.
:::

```{r q8b-prep, message=F, warning=F}
# Weeks unemployed for the pooled entry group (New Entrants + Reentrants).
# is_ne_re already requires is_unemp, so we only drop missing duration.
monthly_dur_nere <- cps[is_ne_re & !is.na(dur_wks), .(
  mean_wks   = weighted_mean(dur_wks, WTFINL),
  median_wks = weighted_median(dur_wks, WTFINL)
), by = date]
setorder(monthly_dur_nere, date)

# Median-duration levels and gaps for the comparison plot below.
# Levels: overall, RCG, and the pooled entry group. Gaps are differences in
# the monthly weighted medians, so they sit near zero on the same week axis.
dur_levels <- merge(
  monthly_dur[, .(date, overall_median, rcg_median)],
  monthly_dur_nere[, .(date, nere_median = median_wks)],
  by = "date", all = TRUE
)
setorder(dur_levels, date)
dur_levels[, `:=`(
  gap_nere_rcg     = nere_median - rcg_median,
  gap_overall_nere = overall_median - nere_median
)]
```

### Entry group only: mean vs median weeks

This pools the two entry reasons from the by-reason plots above — **New Entrants** and **Reentrants** — into a single line and asks how long that combined group sits unemployed. The two lines compare the **mean** (the average spell) against the **median** (the typical, middle spell). When the mean runs above the median, a handful of unusually long searches are pulling the average up while most people's spells are shorter.

```{r q8-ts-nere, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly_dur_nere,
  cols   = c("mean_wks", "median_wks"),
  labels = c("Mean", "Median"),
  colors = c(jco[4], jco[5]),
  title  = "Weeks unemployed: New Entrants + Reentrants (combined)",
  y_label = "Weeks", y_suffix = " wks"
)
```

### Median weeks: levels and gaps

This puts five **median**-duration series on one chart, all measured in weeks. Three are levels you can read directly: (i) the **overall** median across all unemployed, (ii) the **RCG** median, and (iii) the **New Entrants + Reentrants** median. The other two are gaps that hover near zero on the same axis: (iv) **entry group minus RCG**, which asks whether first-job and returning searchers take longer than recent graduates, and (v) **overall minus entry group**, which shows how much shorter (positive) or longer (negative) entry spells run compared with the typical unemployed worker. A gap above zero means the first group in its label waits longer.

```{r q8-ts-median-gaps, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  dur_levels,
  cols   = c("overall_median", "rcg_median", "nere_median",
             "gap_nere_rcg", "gap_overall_nere"),
  labels = c("Overall median", "RCG median", "NE + Reentrant median",
             "Gap: (NE + Re) − RCG", "Gap: overall − (NE + Re)"),
  colors = jco[1:5],
  title  = "Median weeks unemployed: levels and gaps",
  y_label = "Weeks", y_suffix = " wks"
)
```

::: callout-note
The gap lines (iv, v) are differences of monthly weighted medians, so they sit near zero against the same week axis as the level lines and can turn negative. As with the other duration plots, the entry-group medians can be thin in a single month — lean on the 12-month rolling mean (bold) rather than the raw monthly line (light).
:::

## Appendix: Extract check

This section checks the IPUMS pull before reporting estimates: variables, month coverage, sample size, and recoded indicators.

### Weighted median

The weighted median asks: at what value have we reached 50% of the total survey weight? The `weighted_median()` helper drops missing values and non-positive weights, sorts the data by the value, adds weights cumulatively, and returns the first observed value where cumulative weight reaches at least 50%.

Suppose three unemployed people report durations of 4, 10, and 20 weeks, with survey weights 1, 8, and 1. Total weight is 10. After sorting by duration, the cumulative weight shares are 10%, 90%, and 100%. The first duration where cumulative weight reaches at least 50% is 10 weeks, so the weighted median is 10 weeks.

```{r weighted-median-example}
example_duration <- c(4, 10, 20)
example_weight   <- c(1, 8, 1)

example_median_steps <- data.table(
  `Duration (weeks)` = example_duration,
  Weight = example_weight
)
example_median_steps[, `Cumulative weight share` := cumsum(Weight) / sum(Weight)]

example_median_steps |>
  gt() |>
  fmt_percent(columns = `Cumulative weight share`, decimals = 0) |>
  style_gt()
```

This approach always returns an observed value, rather than interpolating between adjacent values.

### Variables present

```{r extract-vars}
expected_vars <- c("YEAR", "MONTH", "WTFINL",
                   "AGE", "EDUC",
                   "EMPSTAT", "LABFORCE",
                   "DURUNEMP", "WHYUNEMP")

var_summary <- data.table(
  Variable = expected_vars,
  Present  = expected_vars %in% names(cps),
  Type     = vapply(expected_vars, \(v) {
    if (v %in% names(cps)) class(cps[[v]])[1] else NA_character_
  }, character(1)),
  `Non-NA rows` = vapply(expected_vars, \(v) {
    if (!v %in% names(cps)) return(NA_character_)
    format(sum(!is.na(cps[[v]])), big.mark = ",")
  }, character(1)),
  Range = vapply(expected_vars, \(v) {
    if (!v %in% names(cps)) return(NA_character_)
    x <- cps[[v]]
    if (is.numeric(x)) {
      sprintf("%s – %s",
              format(min(x, na.rm = TRUE), big.mark = ","),
              format(max(x, na.rm = TRUE), big.mark = ","))
    } else {
      paste(head(sort(unique(x)), 4), collapse = ", ")
    }
  }, character(1))
)

var_summary |> gt() |> style_gt()
```

### Variable descriptions

```{r extract-var-descriptions}
ddi_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.xml")
ddi <- ipumsr::read_ipums_ddi(ddi_path)

label_vars <- c("AGE", "EMPSTAT", "LABFORCE", "DURUNEMP", "WHYUNEMP", "EDUC")

var_descriptions <- as.data.table(ddi$var_info)[
  var_name %in% label_vars,
  .(
    Variable = var_name,
    Label = var_label,
    Description = var_desc
  )
]
var_descriptions[, Variable := factor(Variable, levels = label_vars)]
setorder(var_descriptions, Variable)
var_descriptions[, Variable := as.character(Variable)]

var_descriptions |>
  gt() |>
  cols_width(
    Variable ~ px(90),
    Label ~ px(190),
    Description ~ px(520)
  ) |>
  style_gt()
```

### Value labels for key recodes

The table below comes from the IPUMS DDI. For numeric variables such as `AGE` and `DURUNEMP`, ordinary values are interpreted directly as age or weeks; the DDI mainly lists special values.

```{r extract-value-labels}
value_labels <- rbindlist(lapply(label_vars, \(v) {
  labels <- ddi$var_info$val_labels[[which(ddi$var_info$var_name == v)]]
  if (is.null(labels) || nrow(labels) == 0) {
    return(data.table(Variable = v, Code = NA_real_, Label = "No value labels in DDI"))
  }
  data.table(
    Variable = v,
    Code = labels$val,
    Label = labels$lbl
  )
}))

value_labels |>
  gt(groupname_col = "Variable") |>
  fmt_number(columns = Code, decimals = 0, use_seps = FALSE) |>
  style_gt()
```

### Coverage and sample size

```{r extract-coverage}
month_seq <- seq(min(cps$date), max(cps$date), by = "month")
n_months_in_span <- length(month_seq)
n_months_present <- uniqueN(cps$date)

missing_months <- as.Date(
  setdiff(month_seq, unique(cps$date)),
  origin = "1970-01-01"
)
missing_label <- if (length(missing_months) == 0) {
  "None — every month in the span is represented."
} else {
  paste(format(missing_months, "%B %Y"), collapse = ", ")
}

cov <- data.table(
  Metric = c(
    "First month observed",
    "Last month observed",
    "Months in span",
    "Distinct months in data",
    "Missing months (count)",
    "Missing months (which)",
    "Total unweighted rows",
    "Median rows per month",
    "Sum of WTFINL, latest month (implied population)"
  ),
  Value = c(
    format(min(cps$date), "%B %Y"),
    format(max(cps$date), "%B %Y"),
    format(n_months_in_span, big.mark = ","),
    format(n_months_present, big.mark = ","),
    format(n_months_in_span - n_months_present, big.mark = ","),
    missing_label,
    format(nrow(cps), big.mark = ","),
    format(round(cps[, .N, by = date][, median(N)]), big.mark = ","),
    format(round(cps[date == latest_month, sum(WTFINL, na.rm = TRUE)]),
           big.mark = ",")
  )
)

cov |> gt() |> style_gt()
```

```{r extract-rows-per-year, fig.width=10, fig.height=3.5, message=F, warning=F}
yearly <- cps[, .(rows = .N), by = YEAR][order(YEAR)]
ggplot(yearly, aes(x = YEAR, y = rows)) +
  geom_col(fill = jco[1], alpha = 0.85) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1994, max(yearly$YEAR), 5)) +
  labs(title = "Unweighted CPS rows per year",
       subtitle = "Dips toward zero would indicate missing months for that year.",
       x = NULL, y = "Rows")
```

### Sample size and composition over time

These diagnostics show monthly row counts, the weighted 16+ population against FRED `CNP16OV`, and the components of the RCG definition: age 21–26, bachelor's degree or higher, and their intersection.

```{r extract-diag-prep, message=F, warning=F}
monthly_diag <- cps[, .(
  n_rows                = .N,
  weighted_total_16plus = sum(WTFINL * (AGE >= 16), na.rm = TRUE),
  share_2126            = sum(WTFINL * (AGE >= 21 & AGE <= 26), na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE),
  share_bachelor        = sum(WTFINL * (EDUC >= 111),           na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE),
  share_rcg             = sum(WTFINL * is_rcg,                  na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE)
), by = date]
setorder(monthly_diag, date)
monthly_diag[, `:=`(
  weighted_total_16plus_m = weighted_total_16plus / 1e6,
  n_rows_ma               = roll_mean12(n_rows),
  share_2126_pct          = 100 * share_2126,
  share_bachelor_pct      = 100 * share_bachelor,
  share_rcg_pct           = 100 * share_rcg
)]

# FRED civilian noninstitutional population, 16+ (thousands, monthly).
# BLS uses civilian noninstitutional population controls in CPS weighting.
cnp <- as.data.table(fredr(
  series_id = "CNP16OV",
  observation_start = as.Date("1994-01-01")
))[, .(date, cnp16ov_m = value / 1000)]

monthly_diag <- merge(monthly_diag, cnp, by = "date", all.x = TRUE)
monthly_diag[, weighted_total_16plus_m_ma := roll_mean12(weighted_total_16plus_m)]
```

::: panel-tabset
### Unweighted rows per month

```{r extract-sample-size-rows, fig.width=10, fig.height=4, message=F, warning=F}
p_rows <- ggplot(monthly_diag, aes(x = date)) +
  geom_line(aes(y = n_rows),    color = jco[1], alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = n_rows_ma), color = jco[1], linewidth = 1.0) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "Unweighted CPS rows per month",
       subtitle = "Monthly (light); 12-month rolling mean (bold). NBER recessions shaded.",
       x = NULL, y = "Rows")
add_recessions(p_rows)
```

### CPS 16+ population vs FRED CNP16OV

```{r extract-sample-size-pop, fig.width=10, fig.height=4, message=F, warning=F}
wgt_long <- melt(
  monthly_diag,
  id.vars = "date",
  measure.vars = c("weighted_total_16plus_m", "cnp16ov_m"),
  variable.name = "series", value.name = "millions"
)
wgt_long[, series := factor(
  series,
  levels = c("weighted_total_16plus_m", "cnp16ov_m"),
  labels = c("CPS sum(WTFINL) where AGE ≥ 16",
             "FRED CNP16OV (civilian noninst. 16+)")
)]
wgt_long[, ma := roll_mean12(millions), by = series]

wgt_cols <- c(
  "CPS sum(WTFINL) where AGE ≥ 16"        = jco[2],
  "FRED CNP16OV (civilian noninst. 16+)"  = jco[5]
)

p_wgt <- ggplot(wgt_long, aes(x = date, color = series)) +
  geom_line(aes(y = millions), alpha = 0.25, linewidth = 0.4) +
  geom_line(aes(y = ma),       linewidth = 1.0) +
  scale_color_manual(values = wgt_cols) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "CPS 16+ weighted total vs FRED CNP16OV",
       subtitle = "Monthly (light); 12-month rolling mean (bold).",
       x = NULL, y = "Persons (millions)")
add_recessions(p_wgt)
```
:::

```{r extract-composition-age, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_2126_pct",
           "Share of CPS sample aged 21–26 (weighted)", jco[3])
```

```{r extract-composition-bachelor, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_bachelor_pct",
           "Share of CPS sample with bachelor's degree or higher (weighted)", jco[4])
```

```{r extract-composition-rcg, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_rcg_pct",
           "Share of CPS sample that is RCG: age 21–26 & bachelor's+ (weighted)", jco[1])
```

### Recoded indicators

Counts below are person-months from January 1994 through `r format(latest_month, "%B %Y")`, not unique people. CPS has a rotating panel, so the same person can appear in more than one monthly file. The RCG unemployed row gives the sample size behind the RCG-specific unemployment estimates.

```{r extract-recodes}
recode_summary <- data.table(
  Indicator = c(
    "RCG (age 21–26 & bachelor's+)",
    "YLB (age 18–25 & less than bachelor's)",
    "In labor force (LABFORCE == 2)",
    "Employed (EMPSTAT in {10, 12})",
    "Unemployed (EMPSTAT in {20, 21, 22})",
    "RCG ∩ in labor force",
    "RCG ∩ unemployed",
    "YLB ∩ unemployed",
    "Unemployed ∩ New Entrant (WHYUNEMP == 6)",
    "Unemployed ∩ Reentrant (WHYUNEMP == 5)"
  ),
  `Unweighted rows` = c(
    format(sum(cps$is_rcg,                    na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_ylb,                    na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_lf,                     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_emp,                    na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_unemp,                  na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_rcg & cps$is_lf,        na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_rcg & cps$is_unemp,     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_ylb & cps$is_unemp,     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_ne,                     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_re,                     na.rm = TRUE), big.mark = ",")
  )
)

recode_summary |> gt() |> style_gt()
```
