---
title: "Recent College Graduates: Quick Facts"
author: "Kristin Butcher and Aryan Safi"
format:
  html:
    code-fold: true
    embed-resources: true
    toc: true
    toc-depth: 3
editor: visual
---

```{r setup, include=F}
library(data.table)
library(here)
library(ggplot2)
library(ggsci)
library(scales)
library(gt)
library(lubridate)
library(zoo)
library(fredr)

here::i_am("exploring_new_entrants/markdown/rcg_faq.qmd")

fredr_set_key("ccd350785a210a7653a20d02ce92d83a")

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )
theme_set(theme_report)

jco          <- pal_jco()(6)
rcg_col      <- jco[1]
overall_col  <- jco[2]

recessions <- data.table(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-01", "2009-06-01", "2020-04-01"))
)

add_recessions <- function(p, alpha = 0.08) {
  p + annotate("rect",
               xmin = recessions$start, xmax = recessions$end,
               ymin = -Inf, ymax = Inf,
               fill = "grey40", alpha = alpha)
}

weighted_median <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (!any(ok)) return(NA_real_)
  x <- x[ok]; w <- w[ok]
  ord <- order(x)
  x <- x[ord]; w <- w[ord]
  cw <- cumsum(w) / sum(w)
  x[which(cw >= 0.5)[1]]
}

weighted_mean <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (!any(ok)) return(NA_real_)
  sum(x[ok] * w[ok]) / sum(w[ok])
}

# 12-month rolling helpers (right-aligned, fill leading NAs)
roll_mean12   <- function(x) frollmean(x, 12, align = "right", na.rm = TRUE)
roll_median12 <- function(x) zoo::rollapplyr(x, 12, median, na.rm = TRUE, fill = NA)

fmt_pct <- function(x, d = 1) paste0(formatC(x, digits = d, format = "f"), "%")

# Shared styling for gt tables
style_gt <- function(g) {
  g |>
    tab_options(
      table.font.size = px(11),
      column_labels.font.weight = "bold",
      column_labels.background.color = "#2c3e50"
    ) |>
    tab_style(style = cell_text(color = "white"),
              locations = cells_column_labels())
}
```

## Definition

A **Recent College Graduate (RCG)** is a person who is **age 21–26** and holds a **bachelor's degree or higher** (`AGE %in% 21:26` and `EDUC >= 111` in IPUMS CPS). This captures people near the usual age of bachelor's-degree completion and the first several years afterward.

All shares, rates, means, and medians use the IPUMS basic monthly person weight (`WTFINL`). The sample is the regular monthly basic CPS from January 1994 through the latest extract month, with the ASEC supplement excluded.

```{r data-prep, message=F, warning=F}
csv_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.csv.gz")

cps <- fread(csv_path)

cps[, `:=`(
  date     = make_date(YEAR, MONTH, 1L),
  is_rcg   = AGE >= 21 & AGE <= 26 & EDUC >= 111,
  is_lf    = LABFORCE == 2,
  is_emp   = EMPSTAT %in% c(10L, 12L),
  is_unemp = EMPSTAT %in% c(20L, 21L, 22L)
)]
cps[, is_ne_re := is_unemp & WHYUNEMP %in% c(5L, 6L)]
cps[, is_ne    := is_unemp & WHYUNEMP == 6L]
cps[, is_re    := is_unemp & WHYUNEMP == 5L]
cps[, dur_wks  := fifelse(is_unemp & DURUNEMP < 999,
                          as.numeric(DURUNEMP), NA_real_)]

latest_month    <- cps[, max(date)]
latest_12_start <- latest_month - months(11)
format_window   <- paste0(format(latest_12_start, "%b %Y"),
                          " – ", format(latest_month, "%b %Y"))

# ── Monthly aggregates for shares / rates ───────────────────────────────────
monthly <- cps[, .(
  rcg_unemp_w   = sum(WTFINL * (is_rcg & is_unemp), na.rm = TRUE),
  rcg_lf_w      = sum(WTFINL * (is_rcg & is_lf),    na.rm = TRUE),
  total_unemp_w = sum(WTFINL * is_unemp,            na.rm = TRUE),
  total_lf_w    = sum(WTFINL * is_lf,               na.rm = TRUE),
  ne_re_w       = sum(WTFINL * is_ne_re,            na.rm = TRUE),
  ne_w          = sum(WTFINL * is_ne,               na.rm = TRUE),
  re_w          = sum(WTFINL * is_re,               na.rm = TRUE)
), by = date]
setorder(monthly, date)
monthly[, `:=`(
  rcg_unemp_rate     = 100 * rcg_unemp_w / rcg_lf_w,
  overall_unemp_rate = 100 * total_unemp_w / total_lf_w,
  rcg_share_unemp    = 100 * rcg_unemp_w / total_unemp_w,
  rcg_share_lf       = 100 * rcg_lf_w    / total_lf_w,
  ne_re_share        = 100 * ne_re_w     / total_unemp_w,
  ne_share           = 100 * ne_w        / total_unemp_w,
  re_share           = 100 * re_w        / total_unemp_w,
  rcg_lf_millions    = rcg_lf_w / 1e6
)]

# ── Monthly aggregates for duration ──────────────────────────────────────────
unemp_dt    <- cps[is_unemp & !is.na(dur_wks)]
monthly_dur <- unemp_dt[, .(
  rcg_mean       = weighted_mean(dur_wks[is_rcg], WTFINL[is_rcg]),
  rcg_median     = weighted_median(dur_wks[is_rcg], WTFINL[is_rcg]),
  overall_mean   = weighted_mean(dur_wks, WTFINL),
  overall_median = weighted_median(dur_wks, WTFINL)
), by = date]
setorder(monthly_dur, date)

# ── Window-summary helpers (for the headline tables) ────────────────────────
window_label_12  <- paste0("Latest 12 months (", format_window, ")")
window_label_all <- paste0("Pooled Jan 1994–", format(latest_month, "%b %Y"))

in_window_12 <- function(d) d >= latest_12_start & d <= latest_month

# ── Per-question time-series panel helper (for share/rate plots) ────────────
rate_panel <- function(dt, col, title, color,
                       y_label = "Percent", y_suffix = "%") {
  d <- copy(dt)
  d[, .v   := get(col)]
  d[, .ma  := roll_mean12(.v)]
  d[, .med := roll_median12(.v)]
  p <- ggplot(d, aes(x = date)) +
    geom_line(aes(y = .v),   color = color, alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = .ma),  color = color, linewidth = 1.0, alpha = 0.9) +
    geom_line(aes(y = .med), color = color, linewidth = 0.9, alpha = 0.9,
              linetype = "dashed") +
    scale_y_continuous(labels = \(x) paste0(x, y_suffix)) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly (light); 12-mo rolling mean (solid); 12-mo rolling median (dashed). NBER recessions shaded.",
         y = y_label, x = NULL)
  add_recessions(p)
}

# ── Two-series comparison panel (e.g., RCG vs Overall) ──────────────────────
compare_panel <- function(dt, cols, labels, colors, title, y_label = "Percent",
                          y_suffix = "%") {
  d <- copy(dt)[, c("date", cols), with = FALSE]
  d_long <- melt(d, id.vars = "date", measure.vars = cols,
                 variable.name = "series", value.name = "value")
  d_long[, series := factor(series, levels = cols, labels = labels)]
  d_long[, ma := roll_mean12(value), by = series]

  p <- ggplot(d_long, aes(x = date, color = series)) +
    geom_line(aes(y = value), alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = ma),    linewidth = 1.0, alpha = 0.9) +
    scale_color_manual(values = setNames(colors, labels)) +
    scale_y_continuous(labels = \(x) paste0(x, y_suffix)) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly (light); 12-month rolling mean (bold). NBER recessions shaded.",
         y = y_label, x = NULL)
  add_recessions(p)
}
```

The most recent month in the extract is **`r format(latest_month, "%B %Y")`**. Below, *latest 12 months* means **`r format_window`** and *pooled* means **January 1994–`r format(latest_month, "%B %Y")`**.

## Q1. What percentage of recent college graduates are unemployed?

Unemployed RCGs divided by RCGs in the labor force, weighted by `WTFINL`.

```{r q1-table}
q1_value <- function(dt) {
  d <- dt[is_rcg & is_lf]
  d[, sum(WTFINL * is_unemp, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q1_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG unemployment rate` = c(
    fmt_pct(100 * q1_value(cps[in_window_12(date)])),
    fmt_pct(100 * q1_value(cps))
  )
)

q1_tbl |> gt() |> style_gt()
```

::: panel-tabset
### RCG only

```{r q1-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_unemp_rate",
           "RCG unemployment rate (% of RCG in labor force)", jco[1])
```

### RCG vs Overall

```{r q1-ts-compare, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("rcg_unemp_rate", "overall_unemp_rate"),
  labels = c("RCG", "Overall"),
  colors = c(rcg_col, overall_col),
  title  = "Unemployment rate: RCG vs Overall"
)
```
:::

## Q2. What percentage of the unemployed are recent college graduates?

```{r q2-table}
q2_value <- function(dt) {
  d <- dt[is_unemp == TRUE]
  d[, sum(WTFINL * is_rcg, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q2_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG share of unemployed` = c(
    fmt_pct(100 * q2_value(cps[in_window_12(date)])),
    fmt_pct(100 * q2_value(cps))
  )
)

q2_tbl |> gt() |> style_gt()
```

::: panel-tabset
### RCG share of unemployed

```{r q2-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_share_unemp",
           "RCG share of the unemployed (% of all unemployed)", jco[2])
```

### Share of unemployed vs Share of LF

```{r q2-ts-compare, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("rcg_share_unemp", "rcg_share_lf"),
  labels = c("Share of unemployed", "Share of LF"),
  colors = c(jco[2], jco[3]),
  title  = "RCG: share of unemployed vs share of labor force"
)
```

When the "share of unemployed" line sits above the "share of LF" line, RCGs are *over*-represented in the unemployed pool relative to their presence in the labor force, and vice versa.
:::

## Q3. What percentage of the labor force are recent college graduates?

```{r q3-table}
q3_value <- function(dt) {
  d <- dt[is_lf == TRUE]
  d[, sum(WTFINL * is_rcg, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q3_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `RCG share of labor force` = c(
    fmt_pct(100 * q3_value(cps[in_window_12(date)])),
    fmt_pct(100 * q3_value(cps))
  )
)

q3_tbl |> gt() |> style_gt()
```

::: panel-tabset
### Share of LF

```{r q3-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_share_lf",
           "RCG share of the labor force (% of LF)", jco[3])
```

### Counts (millions)

```{r q3-ts-counts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "rcg_lf_millions",
           "RCG in the labor force (weighted, millions of persons)", jco[3],
           y_label = "Millions of persons", y_suffix = "M")
```
:::

## Q4. What share of unemployed people are New Entrants or Reentrants?

Computed from CPS `WHYUNEMP`: codes 5 (Reentrant) and 6 (New Entrant), divided by all unemployed people (`EMPSTAT %in% 20:22`).

```{r q4-table}
q4_value <- function(dt) {
  d <- dt[is_unemp == TRUE]
  d[, sum(WTFINL * is_ne_re, na.rm = TRUE) / sum(WTFINL, na.rm = TRUE)]
}

q4_tbl <- data.table(
  Window = c(window_label_12, window_label_all),
  `NE + Reentrant share` = c(
    fmt_pct(100 * q4_value(cps[in_window_12(date)])),
    fmt_pct(100 * q4_value(cps))
  )
)

q4_tbl |> gt() |> style_gt()
```

::: panel-tabset
### Combined (NE + Reentrant)

```{r q4-ts, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly, "ne_re_share",
           "NE + Reentrant share of the unemployed (%)", jco[4])
```

### Decomposed

```{r q4-ts-decomposed, fig.width=10, fig.height=4, message=F, warning=F}
compare_panel(
  monthly,
  cols   = c("ne_share", "re_share"),
  labels = c("New Entrants", "Reentrants"),
  colors = c(jco[4], jco[5]),
  title  = "NE-only and Reentrant-only shares of the unemployed"
)
```
:::

## Q5. How long are recent college graduates unemployed compared to everyone else?

Weighted **mean** and **median** weeks unemployed (`DURUNEMP`) for unemployed RCGs and for all unemployed people.

```{r q5-table, message=F, warning=F}
q5_value <- function(dt) {
  rcg <- dt[is_unemp & is_rcg]
  all <- dt[is_unemp == TRUE]
  data.table(
    `RCG mean (weeks)`       = weighted_mean(rcg$dur_wks,  rcg$WTFINL),
    `RCG median (weeks)`     = weighted_median(rcg$dur_wks, rcg$WTFINL),
    `Overall mean (weeks)`   = weighted_mean(all$dur_wks,   all$WTFINL),
    `Overall median (weeks)` = weighted_median(all$dur_wks, all$WTFINL)
  )
}

q5_tbl <- rbindlist(list(
  cbind(data.table(Window = window_label_12),  q5_value(cps[in_window_12(date)])),
  cbind(data.table(Window = window_label_all), q5_value(cps))
))

q5_tbl |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 1) |>
  style_gt()
```

```{r q5-ts-setup, include=F}
# Reshape to long for clean ggplot legend (used by both mean and median tabs)
dur_long_mean <- melt(monthly_dur,
                      id.vars = "date",
                      measure.vars = c("rcg_mean", "overall_mean"),
                      variable.name = "group", value.name = "value")
dur_long_mean[, group := factor(group,
                                levels = c("rcg_mean", "overall_mean"),
                                labels = c("RCG", "Overall"))]
dur_long_mean[, ma := roll_mean12(value), by = group]

dur_long_med <- melt(monthly_dur,
                     id.vars = "date",
                     measure.vars = c("rcg_median", "overall_median"),
                     variable.name = "group", value.name = "value")
dur_long_med[, group := factor(group,
                               levels = c("rcg_median", "overall_median"),
                               labels = c("RCG", "Overall"))]
dur_long_med[, ma := roll_mean12(value), by = group]

dur_cols <- c(RCG = rcg_col, Overall = overall_col)

make_dur_panel <- function(d, title) {
  p <- ggplot(d, aes(x = date, color = group)) +
    geom_line(aes(y = value), alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = ma),    linewidth = 1.0, alpha = 0.9) +
    scale_color_manual(values = dur_cols) +
    scale_y_continuous(labels = \(x) paste0(x, " wks")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title,
         subtitle = "Monthly weighted value (light); 12-month rolling mean (bold). NBER recessions shaded.",
         y = "Weeks", x = NULL)
  add_recessions(p)
}
```

::: panel-tabset
### Mean weeks

```{r q5-ts-mean, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_panel(dur_long_mean, "Mean weeks unemployed: RCG vs overall")
```

### Median weeks

```{r q5-ts-median, fig.width=10, fig.height=4, message=F, warning=F}
make_dur_panel(dur_long_med, "Median weeks unemployed: RCG vs overall")
```
:::

::: callout-note
Source: IPUMS CPS basic monthly microdata, January 1994–`r format(latest_month, "%b %Y")`. ASEC samples excluded. All estimates weighted by `WTFINL`.
:::

## Appendix: Extract check

This section checks the IPUMS pull before reporting estimates: variables, month coverage, sample size, and recoded indicators.

### Weighted median

The weighted median asks: at what value have we reached 50% of the total survey weight? The `weighted_median()` helper drops missing values and non-positive weights, sorts the data by the value, adds weights cumulatively, and returns the first observed value where cumulative weight reaches at least 50%.

Suppose three unemployed people report durations of 4, 10, and 20 weeks, with survey weights 1, 8, and 1. Total weight is 10. After sorting by duration, the cumulative weight shares are 10%, 90%, and 100%. The first duration where cumulative weight reaches at least 50% is 10 weeks, so the weighted median is 10 weeks.

```{r weighted-median-example}
example_duration <- c(4, 10, 20)
example_weight   <- c(1, 8, 1)

example_median_steps <- data.table(
  `Duration (weeks)` = example_duration,
  Weight = example_weight
)
example_median_steps[, `Cumulative weight share` := cumsum(Weight) / sum(Weight)]

example_median_steps |>
  gt() |>
  fmt_percent(columns = `Cumulative weight share`, decimals = 0) |>
  style_gt()
```

This approach always returns an observed value, rather than interpolating between adjacent values.

### Variables present

```{r extract-vars}
expected_vars <- c("YEAR", "MONTH", "WTFINL",
                   "AGE", "EDUC",
                   "EMPSTAT", "LABFORCE",
                   "DURUNEMP", "WHYUNEMP")

var_summary <- data.table(
  Variable = expected_vars,
  Present  = expected_vars %in% names(cps),
  Type     = vapply(expected_vars, \(v) {
    if (v %in% names(cps)) class(cps[[v]])[1] else NA_character_
  }, character(1)),
  `Non-NA rows` = vapply(expected_vars, \(v) {
    if (!v %in% names(cps)) return(NA_character_)
    format(sum(!is.na(cps[[v]])), big.mark = ",")
  }, character(1)),
  Range = vapply(expected_vars, \(v) {
    if (!v %in% names(cps)) return(NA_character_)
    x <- cps[[v]]
    if (is.numeric(x)) {
      sprintf("%s – %s",
              format(min(x, na.rm = TRUE), big.mark = ","),
              format(max(x, na.rm = TRUE), big.mark = ","))
    } else {
      paste(head(sort(unique(x)), 4), collapse = ", ")
    }
  }, character(1))
)

var_summary |> gt() |> style_gt()
```

### Variable descriptions

```{r extract-var-descriptions}
ddi_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.xml")
ddi <- ipumsr::read_ipums_ddi(ddi_path)

label_vars <- c("AGE", "EMPSTAT", "LABFORCE", "DURUNEMP", "WHYUNEMP", "EDUC")

var_descriptions <- as.data.table(ddi$var_info)[
  var_name %in% label_vars,
  .(
    Variable = var_name,
    Label = var_label,
    Description = var_desc
  )
]
var_descriptions[, Variable := factor(Variable, levels = label_vars)]
setorder(var_descriptions, Variable)
var_descriptions[, Variable := as.character(Variable)]

var_descriptions |>
  gt() |>
  cols_width(
    Variable ~ px(90),
    Label ~ px(190),
    Description ~ px(520)
  ) |>
  style_gt()
```

### Value labels for key recodes

The table below comes from the IPUMS DDI. For numeric variables such as `AGE` and `DURUNEMP`, ordinary values are interpreted directly as age or weeks; the DDI mainly lists special values.

```{r extract-value-labels}
value_labels <- rbindlist(lapply(label_vars, \(v) {
  labels <- ddi$var_info$val_labels[[which(ddi$var_info$var_name == v)]]
  if (is.null(labels) || nrow(labels) == 0) {
    return(data.table(Variable = v, Code = NA_real_, Label = "No value labels in DDI"))
  }
  data.table(
    Variable = v,
    Code = labels$val,
    Label = labels$lbl
  )
}))

value_labels |>
  gt(groupname_col = "Variable") |>
  fmt_number(columns = Code, decimals = 0, use_seps = FALSE) |>
  style_gt()
```

### Coverage and sample size

```{r extract-coverage}
month_seq <- seq(min(cps$date), max(cps$date), by = "month")
n_months_in_span <- length(month_seq)
n_months_present <- uniqueN(cps$date)

missing_months <- as.Date(
  setdiff(month_seq, unique(cps$date)),
  origin = "1970-01-01"
)
missing_label <- if (length(missing_months) == 0) {
  "None — every month in the span is represented."
} else {
  paste(format(missing_months, "%B %Y"), collapse = ", ")
}

cov <- data.table(
  Metric = c(
    "First month observed",
    "Last month observed",
    "Months in span",
    "Distinct months in data",
    "Missing months (count)",
    "Missing months (which)",
    "Total unweighted rows",
    "Median rows per month",
    "Sum of WTFINL, latest month (implied population)"
  ),
  Value = c(
    format(min(cps$date), "%B %Y"),
    format(max(cps$date), "%B %Y"),
    format(n_months_in_span, big.mark = ","),
    format(n_months_present, big.mark = ","),
    format(n_months_in_span - n_months_present, big.mark = ","),
    missing_label,
    format(nrow(cps), big.mark = ","),
    format(round(cps[, .N, by = date][, median(N)]), big.mark = ","),
    format(round(cps[date == latest_month, sum(WTFINL, na.rm = TRUE)]),
           big.mark = ",")
  )
)

cov |> gt() |> style_gt()
```

```{r extract-rows-per-year, fig.width=10, fig.height=3.5, message=F, warning=F}
yearly <- cps[, .(rows = .N), by = YEAR][order(YEAR)]
ggplot(yearly, aes(x = YEAR, y = rows)) +
  geom_col(fill = jco[1], alpha = 0.85) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_continuous(breaks = seq(1994, max(yearly$YEAR), 5)) +
  labs(title = "Unweighted CPS rows per year",
       subtitle = "Dips toward zero would indicate missing months for that year.",
       x = NULL, y = "Rows")
```

### Sample size and composition over time

These diagnostics show monthly row counts, the weighted 16+ population against FRED `CNP16OV`, and the components of the RCG definition: age 21–26, bachelor's degree or higher, and their intersection.

```{r extract-diag-prep, message=F, warning=F}
monthly_diag <- cps[, .(
  n_rows                = .N,
  weighted_total_16plus = sum(WTFINL * (AGE >= 16), na.rm = TRUE),
  share_2126            = sum(WTFINL * (AGE >= 21 & AGE <= 26), na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE),
  share_bachelor        = sum(WTFINL * (EDUC >= 111),           na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE),
  share_rcg             = sum(WTFINL * is_rcg,                  na.rm = TRUE) /
                          sum(WTFINL, na.rm = TRUE)
), by = date]
setorder(monthly_diag, date)
monthly_diag[, `:=`(
  weighted_total_16plus_m = weighted_total_16plus / 1e6,
  n_rows_ma               = roll_mean12(n_rows),
  share_2126_pct          = 100 * share_2126,
  share_bachelor_pct      = 100 * share_bachelor,
  share_rcg_pct           = 100 * share_rcg
)]

# FRED civilian noninstitutional population, 16+ (thousands, monthly).
# BLS uses civilian noninstitutional population controls in CPS weighting.
cnp <- as.data.table(fredr(
  series_id = "CNP16OV",
  observation_start = as.Date("1994-01-01")
))[, .(date, cnp16ov_m = value / 1000)]

monthly_diag <- merge(monthly_diag, cnp, by = "date", all.x = TRUE)
monthly_diag[, weighted_total_16plus_m_ma := roll_mean12(weighted_total_16plus_m)]
```

::: panel-tabset
### Unweighted rows per month

```{r extract-sample-size-rows, fig.width=10, fig.height=4, message=F, warning=F}
p_rows <- ggplot(monthly_diag, aes(x = date)) +
  geom_line(aes(y = n_rows),    color = jco[1], alpha = 0.3, linewidth = 0.4) +
  geom_line(aes(y = n_rows_ma), color = jco[1], linewidth = 1.0) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "Unweighted CPS rows per month",
       subtitle = "Monthly (light); 12-month rolling mean (bold). NBER recessions shaded.",
       x = NULL, y = "Rows")
add_recessions(p_rows)
```

### CPS 16+ population vs FRED CNP16OV

```{r extract-sample-size-pop, fig.width=10, fig.height=4, message=F, warning=F}
wgt_long <- melt(
  monthly_diag,
  id.vars = "date",
  measure.vars = c("weighted_total_16plus_m", "cnp16ov_m"),
  variable.name = "series", value.name = "millions"
)
wgt_long[, series := factor(
  series,
  levels = c("weighted_total_16plus_m", "cnp16ov_m"),
  labels = c("CPS sum(WTFINL) where AGE ≥ 16",
             "FRED CNP16OV (civilian noninst. 16+)")
)]
wgt_long[, ma := roll_mean12(millions), by = series]

wgt_cols <- c(
  "CPS sum(WTFINL) where AGE ≥ 16"        = jco[2],
  "FRED CNP16OV (civilian noninst. 16+)"  = jco[5]
)

p_wgt <- ggplot(wgt_long, aes(x = date, color = series)) +
  geom_line(aes(y = millions), alpha = 0.25, linewidth = 0.4) +
  geom_line(aes(y = ma),       linewidth = 1.0) +
  scale_color_manual(values = wgt_cols) +
  scale_y_continuous(labels = label_comma()) +
  scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
  labs(title = "CPS 16+ weighted total vs FRED CNP16OV",
       subtitle = "Monthly (light); 12-month rolling mean (bold).",
       x = NULL, y = "Persons (millions)")
add_recessions(p_wgt)
```
:::

```{r extract-composition-age, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_2126_pct",
           "Share of CPS sample aged 21–26 (weighted)", jco[3])
```

```{r extract-composition-bachelor, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_bachelor_pct",
           "Share of CPS sample with bachelor's degree or higher (weighted)", jco[4])
```

```{r extract-composition-rcg, fig.width=10, fig.height=4, message=F, warning=F}
rate_panel(monthly_diag, "share_rcg_pct",
           "Share of CPS sample that is RCG: age 21–26 & bachelor's+ (weighted)", jco[1])
```

### Recoded indicators

Counts below are person-months from January 1994 through `r format(latest_month, "%B %Y")`, not unique people. CPS has a rotating panel, so the same person can appear in more than one monthly file. The RCG unemployed row gives the sample size behind the RCG-specific unemployment estimates.

```{r extract-recodes}
recode_summary <- data.table(
  Indicator = c(
    "RCG (age 21–26 & bachelor's+)",
    "In labor force (LABFORCE == 2)",
    "Employed (EMPSTAT in {10, 12})",
    "Unemployed (EMPSTAT in {20, 21, 22})",
    "RCG ∩ in labor force",
    "RCG ∩ unemployed",
    "Unemployed ∩ New Entrant (WHYUNEMP == 6)",
    "Unemployed ∩ Reentrant (WHYUNEMP == 5)"
  ),
  `Unweighted rows` = c(
    format(sum(cps$is_rcg,                    na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_lf,                     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_emp,                    na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_unemp,                  na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_rcg & cps$is_lf,        na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_rcg & cps$is_unemp,     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_ne,                     na.rm = TRUE), big.mark = ","),
    format(sum(cps$is_re,                     na.rm = TRUE), big.mark = ",")
  )
)

recode_summary |> gt() |> style_gt()
```
