---
title: "Recent College Graduates: Quick Facts"
author: "Kristin Butcher, Scott Brave, Aryan Safi"
format:
  html:
    code-fold: true
    embed-resources: true
editor: visual
---

```{r setup, include=F}
library(tidyverse)
library(ipumsr)
library(here)
library(ggsci)
library(scales)
library(gt)
library(lubridate)
library(patchwork)
library(zoo)

here::i_am("exploring_new_entrants/markdown/rcg_faq.qmd")

theme_report <- theme_minimal(base_size = 11, base_family = "sans") +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(color = "grey50", size = 9),
    plot.caption = element_text(color = "grey60", size = 7, hjust = 0),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = element_blank(),
    axis.title = element_text(size = 9, color = "grey40"),
    axis.text = element_text(size = 8, color = "grey50"),
    legend.position = "top",
    legend.text = element_text(size = 8),
    legend.title = element_blank()
  )
theme_set(theme_report)

jco <- pal_jco()(6)
rcg_col      <- jco[1]
overall_col  <- jco[2]

recessions <- tibble(
  start = as.Date(c("2001-03-01", "2007-12-01", "2020-02-01")),
  end   = as.Date(c("2001-11-01", "2009-06-01", "2020-04-01"))
)

add_recessions <- function(p, alpha = 0.08) {
  p + annotate("rect",
               xmin = recessions$start, xmax = recessions$end,
               ymin = -Inf, ymax = Inf,
               fill = "grey40", alpha = alpha)
}

weighted_median <- function(x, w) {
  ok <- !is.na(x) & !is.na(w) & w > 0
  if (!any(ok)) return(NA_real_)
  x <- x[ok]; w <- w[ok]
  ord <- order(x)
  x <- x[ord]; w <- w[ord]
  cw <- cumsum(w) / sum(w)
  x[which(cw >= 0.5)[1]]
}

roll_mean12   <- function(x) zoo::rollapplyr(x, 12, mean,   na.rm = TRUE, fill = NA)
roll_median12 <- function(x) zoo::rollapplyr(x, 12, median, na.rm = TRUE, fill = NA)

fmt_pct <- function(x, d = 1) paste0(formatC(x, digits = d, format = "f"), "%")
```

## Definition

A **Recent College Graduate (RCG)** is a person who is **age 21–26** and holds a **bachelor's degree or higher**. In the IPUMS CPS harmonized variables this corresponds to `AGE %in% 21:26` and `EDUC >= 111`. The lower age bound captures the typical age of bachelor's-degree completion; the upper bound restricts attention to the first five post-graduation years.

All shares, rates, means, and medians use the IPUMS basic monthly person weight (`WTFINL`). The sample is the regular monthly basic CPS, 1994–present, with the ASEC supplement excluded.

```{r load-data, message=F, warning=F}
ddi_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.xml")
csv_path <- here("exploring_new_entrants", "data", "CPS", "cps_rcg.csv.gz")

ddi <- read_ipums_ddi(ddi_path)
cps <- read_ipums_micro(ddi, data_file = csv_path, verbose = FALSE)

cps <- cps |>
  mutate(
    date     = make_date(YEAR, MONTH, 1),
    is_rcg   = AGE >= 21 & AGE <= 26 & EDUC >= 111,
    is_lf    = LABFORCE == 2,
    is_emp   = EMPSTAT %in% c(1, 10, 12),
    is_unemp = EMPSTAT %in% c(20, 21, 22),
    is_ne_re = is_unemp & WHYUNEMP %in% c(5, 6),
    dur_wks  = if_else(is_unemp & DURUNEMP < 999, as.numeric(DURUNEMP), NA_real_)
  )

latest_month    <- max(cps$date)
latest_12_start <- latest_month - months(11)
is_latest_12    <- function(d) d >= latest_12_start & d <= latest_month
format_window   <- paste0(format(latest_12_start, "%b %Y"), " – ", format(latest_month, "%b %Y"))
```

The most recent month in the extract is **`r format(latest_month, "%B %Y")`**. Below, *latest 12 months* refers to **`r format_window`** and *pooled* refers to **January 1994 – `r format(latest_month, "%B %Y")`**.

## Frequently Asked Questions

### Q1. What percentage of recent college graduates are unemployed?

Unemployed RCGs divided by RCGs in the labor force, weighted by `WTFINL`.

```{r q1, message=F, warning=F}
q1_window <- function(df, label) {
  df |>
    filter(is_rcg, is_lf) |>
    summarise(
      Window = label,
      `RCG unemployment rate` = sum(WTFINL * is_unemp) / sum(WTFINL * is_lf)
    )
}

bind_rows(
  q1_window(filter(cps, is_latest_12(date)), paste0("Latest 12 months (", format_window, ")")),
  q1_window(cps, "Pooled 1994–present")
) |>
  mutate(`RCG unemployment rate` = fmt_pct(100 * `RCG unemployment rate`)) |>
  gt() |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels())
```

### Q2. What percentage of the unemployed are recent college graduates?

```{r q2}
q2_window <- function(df, label) {
  df |>
    filter(is_unemp) |>
    summarise(
      Window = label,
      `RCG share of unemployed` = sum(WTFINL * is_rcg) / sum(WTFINL)
    )
}

bind_rows(
  q2_window(filter(cps, is_latest_12(date)), paste0("Latest 12 months (", format_window, ")")),
  q2_window(cps, "Pooled 1994–present")
) |>
  mutate(`RCG share of unemployed` = fmt_pct(100 * `RCG share of unemployed`)) |>
  gt() |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels())
```

### Q3. What percentage of the labor force are recent college graduates?

```{r q3}
q3_window <- function(df, label) {
  df |>
    filter(is_lf) |>
    summarise(
      Window = label,
      `RCG share of labor force` = sum(WTFINL * is_rcg) / sum(WTFINL)
    )
}

bind_rows(
  q3_window(filter(cps, is_latest_12(date)), paste0("Latest 12 months (", format_window, ")")),
  q3_window(cps, "Pooled 1994–present")
) |>
  mutate(`RCG share of labor force` = fmt_pct(100 * `RCG share of labor force`)) |>
  gt() |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels())
```

### Q4. Of those who say they are unemployed, what percentage classify themselves as New Entrants or Reentrants?

Computed directly from CPS `WHYUNEMP`: codes 5 (Reentrant) and 6 (New Entrant), divided by all unemployed (`EMPSTAT %in% 20:22`).

```{r q4}
q4_window <- function(df, label) {
  df |>
    filter(is_unemp) |>
    summarise(
      Window = label,
      `NE + Reentrant share` = sum(WTFINL * is_ne_re) / sum(WTFINL)
    )
}

bind_rows(
  q4_window(filter(cps, is_latest_12(date)), paste0("Latest 12 months (", format_window, ")")),
  q4_window(cps, "Pooled 1994–present")
) |>
  mutate(`NE + Reentrant share` = fmt_pct(100 * `NE + Reentrant share`)) |>
  gt() |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels())
```

### Q5. How long are recent college graduates unemployed compared to everyone else?

Weighted **mean** and **median** weeks of unemployment (`DURUNEMP`) for unemployed RCGs and for all unemployed persons.

```{r q5, message=F, warning=F}
q5_window <- function(df, label) {
  rcg <- df |> filter(is_unemp, is_rcg)
  all_unemp <- df |> filter(is_unemp)

  tibble(
    Window = label,
    `RCG mean (weeks)`       = weighted.mean(rcg$dur_wks,       rcg$WTFINL,       na.rm = TRUE),
    `RCG median (weeks)`     = weighted_median(rcg$dur_wks,     rcg$WTFINL),
    `Overall mean (weeks)`   = weighted.mean(all_unemp$dur_wks, all_unemp$WTFINL, na.rm = TRUE),
    `Overall median (weeks)` = weighted_median(all_unemp$dur_wks, all_unemp$WTFINL)
  )
}

bind_rows(
  q5_window(filter(cps, is_latest_12(date)), paste0("Latest 12 months (", format_window, ")")),
  q5_window(cps, "Pooled 1994–present")
) |>
  gt() |>
  fmt_number(columns = where(is.numeric), decimals = 1) |>
  tab_options(table.font.size = px(11), column_labels.font.weight = "bold",
              column_labels.background.color = "#2c3e50") |>
  tab_style(style = cell_text(color = "white"), locations = cells_column_labels())
```

## Time Series, 1994 – Present

Each share/rate panel below shows the monthly value (light) overlaid with a 12-month rolling **mean** (solid bold) and a 12-month rolling **median** (dashed bold). NBER recessions shaded.

```{r monthly-shares, fig.width=10, fig.height=11, message=F, warning=F}
monthly <- cps |>
  group_by(date) |>
  summarise(
    rcg_unemp_w   = sum(WTFINL * (is_rcg & is_unemp), na.rm = TRUE),
    rcg_lf_w      = sum(WTFINL * (is_rcg & is_lf),    na.rm = TRUE),
    total_unemp_w = sum(WTFINL * is_unemp,            na.rm = TRUE),
    total_lf_w    = sum(WTFINL * is_lf,               na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(date) |>
  mutate(
    rcg_unemp_rate   = 100 * rcg_unemp_w / rcg_lf_w,
    rcg_share_unemp  = 100 * rcg_unemp_w / total_unemp_w,
    rcg_share_lf     = 100 * rcg_lf_w    / total_lf_w
  )

make_rate_panel <- function(df, col, title, color) {
  d <- df |>
    mutate(
      .v    = .data[[col]],
      .ma   = roll_mean12(.v),
      .med  = roll_median12(.v)
    )
  p <- ggplot(d, aes(x = date)) +
    geom_line(aes(y = .v),   color = color, alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = .ma),  color = color, linewidth = 1.0, alpha = 0.9) +
    geom_line(aes(y = .med), color = color, linewidth = 0.9, alpha = 0.9,
              linetype = "dashed") +
    scale_y_continuous(labels = \(x) paste0(x, "%")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title, y = "Percent", x = NULL)
  add_recessions(p)
}

p1 <- make_rate_panel(monthly, "rcg_unemp_rate",
                      "RCG unemployment rate (% of RCG in labor force)", jco[1])
p2 <- make_rate_panel(monthly, "rcg_share_unemp",
                      "RCG share of the unemployed (% of all unemployed)", jco[2])
p3 <- make_rate_panel(monthly, "rcg_share_lf",
                      "RCG share of the labor force (% of LF)", jco[3])

p1 / p2 / p3 +
  plot_annotation(
    title = "Recent College Graduates, 1994 – Present",
    subtitle = "Monthly (light); 12-month rolling mean (solid); 12-month rolling median (dashed)."
  )
```

### Duration of unemployment: RCG vs overall

Each month we compute the weighted mean and weighted median of `DURUNEMP` for unemployed RCGs and for all unemployed persons, then overlay a 12-month rolling mean on each monthly series.

```{r monthly-duration, fig.width=10, fig.height=8, message=F, warning=F}
monthly_dur <- cps |>
  filter(is_unemp, !is.na(dur_wks)) |>
  group_by(date) |>
  summarise(
    rcg_mean        = weighted.mean(dur_wks[is_rcg], WTFINL[is_rcg], na.rm = TRUE),
    rcg_median      = weighted_median(dur_wks[is_rcg], WTFINL[is_rcg]),
    overall_mean    = weighted.mean(dur_wks, WTFINL, na.rm = TRUE),
    overall_median  = weighted_median(dur_wks, WTFINL),
    .groups = "drop"
  ) |>
  arrange(date) |>
  mutate(
    rcg_mean_ma       = roll_mean12(rcg_mean),
    rcg_median_ma     = roll_mean12(rcg_median),
    overall_mean_ma   = roll_mean12(overall_mean),
    overall_median_ma = roll_mean12(overall_median)
  )

make_dur_panel <- function(df, raw_rcg, raw_all, ma_rcg, ma_all, title) {
  d <- df |>
    transmute(date,
              rcg_raw = .data[[raw_rcg]],
              all_raw = .data[[raw_all]],
              rcg_ma  = .data[[ma_rcg]],
              all_ma  = .data[[ma_all]])
  p <- ggplot(d, aes(x = date)) +
    geom_line(aes(y = rcg_raw), color = rcg_col,     alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = all_raw), color = overall_col, alpha = 0.25, linewidth = 0.4) +
    geom_line(aes(y = rcg_ma),  color = rcg_col,     linewidth = 1.0, alpha = 0.9) +
    geom_line(aes(y = all_ma),  color = overall_col, linewidth = 1.0, alpha = 0.9) +
    scale_y_continuous(labels = \(x) paste0(x, " wks")) +
    scale_x_date(date_breaks = "5 years", date_labels = "%Y") +
    labs(title = title, y = "Weeks", x = NULL)
  add_recessions(p)
}

p_dur_mean <- make_dur_panel(monthly_dur,
                             "rcg_mean", "overall_mean",
                             "rcg_mean_ma", "overall_mean_ma",
                             "Mean weeks unemployed")
p_dur_med  <- make_dur_panel(monthly_dur,
                             "rcg_median", "overall_median",
                             "rcg_median_ma", "overall_median_ma",
                             "Median weeks unemployed")

# Tiny legend strip
legend_df <- tibble(
  x = as.Date(c("1995-01-01", "2005-01-01")),
  y = c(1, 1),
  label = c("RCG", "Overall"),
  color = c(rcg_col, overall_col)
)
legend_strip <- ggplot(legend_df, aes(x = x, y = y, color = color)) +
  geom_point(size = 4) +
  geom_text(aes(label = label), color = "grey25", hjust = -0.4, vjust = 0.5, size = 3.5) +
  scale_color_identity() +
  theme_void() +
  xlim(as.Date("1994-01-01"), as.Date("2012-01-01"))

(legend_strip / p_dur_mean / p_dur_med) +
  patchwork::plot_layout(heights = c(0.15, 1, 1)) +
  plot_annotation(
    title = "Duration of unemployment: RCG vs overall",
    subtitle = "Monthly weighted mean/median (light); 12-month rolling mean of each (bold)."
  )
```

::: {.callout-note}
Source: IPUMS CPS basic monthly microdata, January 1994 – `r format(latest_month, "%b %Y")`. ASEC samples excluded. All estimates weighted by `WTFINL`.
:::
